---
title: "PCRedux package - an overview"
author: "Stefan R&ouml;diger, Micha&#322; Burdukiewicz, Andrej-Nikolai Spiess"
date: "October 31, 2017"
output: 
  rmarkdown::pdf_document:
    number_sections: true
    toc: true
bibliography: "literature.bib"
vignette: >
  %\VignetteIndexEntry{PCRedux package - an overview}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

\begin{figure}[ht]
\centering
\scalebox{0.6}{
\includegraphics[clip=true,trim=1cm 1cm 1cm 1cm]{Logo.pdf}
}
\end{figure}


# Introduction to the analysis of simgmoid shaped curves – or the analysis of amplification curve data from quantitative real-time PCR experiments \label{section_introduction}

`PCRedux` is an R package designed to perform analysis of sigmoid curves. A 
sigmoid function results a S-shaped curve. For example, the equation 
\autoref{sigmoid} produces such curvature.

\begin{equation}
\label{sigmoid}
f(x) = \frac{1}{1 +  e^{-x}}
\end{equation}

These non-linear functions are real-valued and can be differentiated (first derivative 
maximum, with one local minimum and one local maximum).

This kind of curve is common in many biological assays such as quantitative 
real-time PCR (qPCR) experiments. Applications include human diagnostics and 
forensics [@martins_dna_2015, @sauer_differentiation_2016]. qPCRs are performed 
in thermo-cyclers. There are many manufactures of such systems, which offer such 
technologies as commercial products. Moreover, thermo-cyclers are produced in 
house as part of scientific projects. An example is the VideoScan-technology 
[@rodiger_highly_2013].

A challenge for the user is, to make sense from the data produced by these 
devices. This is a manageable task if the volume of data is low, or dedicated 
software is available for the analysis thereof. An example for a low number of 
amplification curves is shown in Figure \ref{samples_of_qPCRs}A. All `r 
ncol(chipPCR::C127EGHP)-1` curves exhibit a sigmoid curve shape. In contrast, 
the example in Figure \ref{samples_of_qPCRs}B is no longer manageable by simple 
visual inspection. The data originate from a high-throughput experiment an 
encompass in total `r ncol(qpcR::htPCR)-1` amplification curves. Similarly, a 
manual analysis of the data is time-consuming and prone to errors. Dedicated 
software that can distinguish the amplification curves automatically is needed.


```{r, echo=FALSE, fig.cap="Sample data from A) the **C127EGHP** data set with 64 amplification curves (*chipPCR* package, [@rodiger_chippcr:_2015]) and B) from the **htPCR** data set with 8858 amplification curves (*qpcR* package, [@Ritz2008]). \\label{samples_of_qPCRs}", fig.height=5.5, fig.width=11}
    
library(chipPCR)
suppressMessages(library(qpcR))


par(mfrow=c(1,2), las=0, bty="n", oma=c(1,1,1,1))
colors <- rainbow(ncol(C127EGHP)-2, alpha=0.3)
matplot(C127EGHP[, 2], C127EGHP[, c(-1, -2)], xlab="Cycle", ylab="RFU", 
        main="C127EGHP dataset", type="l", lty=1, lwd=2, col=colors)
mtext("A", cex = 2, side = 3, adj = 0, font = 2)

colors <- rainbow(ncol(htPCR)-1, alpha=0.2)
matplot(htPCR[, 1], htPCR[, c(-1)], xlab="Cycle", ylab="RFU", 
        main="htPCR dataset", type="l", lty=1, lwd=2, col=colors)
mtext("B", cex = 2, side = 3, adj = 0, font = 2)
```

There are several open source and closed source software tools, which can be 
used 
for the analysis of qPCR data. The software packages deal for example with
    
    * missing values, 
    * noise,
    * inter run calibration, 
    * normalization, 
    * quantification cycle estimation, 
    * amplification efficiency estimation, 
    * data import, 
    * datat export, 
    * data exchange, 
    * realative gene expression analysis,
    
and other tasks [@pabinger_survey_2014, @rodiger_chippcr:_2015, 
@Ritz2008, @rodiger_r_2015, @lefever_rdml:_2009, 
@ruijter_rdml-ninja_2015, @rodiger_enabling_2017, @perkins_readqpcr_2012, 
@pabinger_qpcr:_2009, @neve_unifiedwmwqpcr:_2014, @feuer_lemming:_2015, 
@mccall_non-detects_2014, @ruijter_evaluation_2013, @ruijter_removal_2015, 
@dvinge_htqpcr:_2009, @ronde_practical_2017, @mallona_pcrefficiency:_2011, 
@mallona_chainy:_nodate]. Many of them are implemented in the R statistical 
computing language. It should be mentioned that this list is far from 
completeness. Dedicated literature is available from peer-reviewed publications 
and textbooks.

Others and we have shown that amplification curve analysis is not a simple 
task. For example selected qPCR systems (device and/or detection chemistry) 
cause periodicity in the amplification curve data 
[@spiess_system-specific_2016]. Or the commonly employed smoothing of data 
might 
case alterations to the raw data that affects both the estimation of the 
template material (Cq value) and the amplification efficiency 
[@spiess_impact_2015].

At this level all software assumes that the amplification resemble a sigmoid 
curve shape (ideal positive amplification reaction) or a flat low line (ideal 
negative amplification reaction). For example, @Ritz2008 published an R 
package that contains functions to fit several multi-parameter models. This 
includes the five-parameter Richardson function, which is often used for the 
analysis of qPCR data.

In general, the software packages do not take into account if an amplification 
curve fulfills a certain criterion (e.g., positive or negative, good quality or 
bad quality). This is a needed information for some algorithms. For example, 
the 
*linreg* method by @ruijter_amplification_2009 requires a decision, if an 
amplification curve is positive or negative. 

To illistrate this problem we give and example for the analysis of amplification 
curves of the ``htPCR`` data set in the section about the function 
`decision_modus`.

In summary, during our past research we noticed that a many challenges are well 
addressed. In particular, algorithms for the processing of the positive 
amplification curves are widely available. One bottleneck in qPCR data 
analysis is the lack of amplification curve data classifier. Classifier herein 
refer to a vector of features that can be used to distinguish the amplification 
curves by their shape only.

One reason for this is the lack of features that are known for amplification 
curve data. Only few features for amplification curves are described in the 
literature. For example, tools described by us are the ``amptester`` and the 
``amptester.gui`` functions, which are part of the `chipPCR` package 
([@rodiger_chippcr:_2015]). The **qpcR** package [@Ritz2008] contains 
an amplification curve test via the ``modlist`` function. The parameter check = 
"uni2" offers an analytical approach, as part of a method for the kinetic 
outlier detection. It checks for a sigmoid structure of the amplification curve. 
Then it tests for the location of the first derivative maximum and the second 
derivative maximum. However, multi-parameter functions, like sigmoid models, fit 
"successful" in most cases including noise and give false positive results. 

# Aims of the `PCRedux` package \label{Section_Aims_of_PCRedux}

The `PCRedux` package contains function and data sets for machine learning on 
quantitative PCR data in R. There are numerous software packages for R which 
can 
be used for the analysis of quantitative PCR (qPCR) data [@rodiger_r_2015, 
@pabinger_survey_2014].

This vignette covers important features of the package and should be used as an 
addendum to the manual.

Reproducible research is an important foundation of research. Both, the technical 
and the experimental aspects need to be performed under principles that follow 
good practice of reproducible research. Several authors addressed the matter 
[@huggett_how_2014, @bustin_reproducibility_2014, @bustin_continuing_2017, 
@rodiger_r_2015, @rodiger_enabling_2017, @wilson_good_2016]. During our 
literature studies we found barely material, which can be used for further 
studies. Therefore, we included data set of amplification curves, human rated 
amplification curves (negative, ambiguous, positive) and examples in this 
package.

# Functions of the `PCRedux` package \label{section_Functions_of_PCRedux}

The function described following are aimed for experimental studies. It is 
important to note that the features proposed herein emerged during a critical 
reasoning process. The aim of the package is to propose a set of features, 
functions and data for an independent research.

## ``autocorrelation_test`` - A function to detect positive amplification curves \label{section_autocorrelation_test}

Autocorrelation analysis is a technique that is used in the field of time series 
analysis. It can be used to reveal regularly occurring patterns in 
one-dimensional data [@spiess_system-specific_2016]. The autocorrelation
measures the correlation of a signal $f(t)$ with itself shifted by some 
time delay $f(t - \tau)$.

The `autocorrelation_test` function coercers the amplification curve data to an 
object of the class "zoo" (*zoo* package) as indexed totally ordered 
observations. Next follows the computation of a lagged version of the 
amplification curve data. The shifting the amplification curve data is based 
back by a given number of observations (default $\tau = 3$). Then follows a 
significance test for correlation between paired samples (amplification curve 
data & lagged amplification curve data). The hypothesis is that the paired 
sample of positive amplification curves has a significant correlation 
(`stats::cor.test`, significance level is 0.01) in contrast to negative 
amplification curves (noise).
The application of the `autocorrelation_test` function is shown in the following 
example. 

```{r, echo=TRUE, fig.cap="Autocorrelation analysis for amplification curves of the **testdat** data set (*qpcR* package, [@Ritz2008]). A) All amplification curves of the **testdat** data set. B) Positive curves and negative curves as determined by the `autocorrelation_test` and a human. C) Performance analysis by the `performeR` function (see Section \\ref{section_performeR} for details). \\label{autocorrelation}", fig.height=5.5, fig.width=11}
# Test for autocorrelation in amplification curve data
# Load the libraries magrittr for pipes and qpcR for the data
library(magrittr)
library(qpcR)
library(PCRedux)
# Test for autocorrelation in the testdat dataset
res_ac <- sapply(2:ncol(testdat), function(i) {
    autocorrelation_test(testdat[, i])
}
)

# Plot curve data as overview
# Define the colors for the amplification curves
colors <- rainbow(ncol(testdat)-1, alpha=0.3)
# Names of samples
samples <- colnames(testdat)[-1]
layout(matrix(c(1,2,1,3), 2, 2, byrow = TRUE))
matplot(testdat[, 1], testdat[, -1], xlab="Cycle", ylab="RFU", 
        main="testdat dataset", type="l", lty=1, col=colors, lwd=2)
legend("topleft", samples, pch=19, col=colors, ncol=2, bty="n")
mtext("A", cex = 2, side = 3, adj = 0, font = 2)

# Curves rated by a human after analysis of the overview. 1 = positive, 
# 0 = negative
human_rating <- c(1,1,0,0,1,1,0,0,
                  1,1,0,0,1,1,0,0,
                  1,1,0,0,1,1,0,0)

# Convert the n.s. (not significant) in 0 and others to 1. 
# Combine the results of the aromatic autocorrelation_test as variable "ac", 
# the human rated values as variable "hr" in a new data frame (res_ac_hr).
res_ac_hr <- data.frame(ac=ifelse(res_ac=="n.s.", 0, 1), 
                        hr=human_rating) %>% as.matrix
                        res_performeR <- performeR(res_ac_hr[, "ac"], res_ac_hr[, "hr"])
                        
                        # Add ratings by human and autocorrelation_test to the plot
                        par(las=2)
                        plot(1:nrow(res_ac_hr), res_ac_hr[, "hr"], xlab="Sample", 
			     ylab="Decisions", xaxt="n", yaxt="n", pch=19)
                        axis(2, at=c(0,1), labels=c("negative", "positive"), las=2)
                        axis(1, at=1:nrow(res_ac_hr), labels=colnames(testdat)[-1], las=2)
                        points(1:nrow(res_ac_hr), res_ac_hr[, "ac"], pch=1, cex=2, 
			      col="red")
                        legend("topleft", c("Human", "autocorrelation_test"), pch=c(19,1), 
                               bty="n", col=c("black","red"))
                        mtext("B", cex = 2, side = 3, adj = 0, font = 2, las=0)
                        
                        barplot(as.matrix(res_performeR[, c(1:10,12)]), yaxt="n", ylab="", 
                                main="Performance of autocorrelation_test")
                        axis(2, at=c(0,1), labels=c("0", "1"), las=2)
                        mtext("C", cex = 2, side = 3, adj = 0, font = 2, las=0)
```

As shown in this example, the `autocorrelation_test` function is able to 
distinguish between positive and negative amplification curves. Negative 
amplification curve were in all cases non-significant. In contrast, the 
coefficients of correlation for positive amplification curves ranged between `r 
signif(range(as.numeric(res_ac), na.rm=TRUE), 3)[1]` and `r 
signif(range(as.numeric(res_ac), na.rm=TRUE), 3)[2]` at a significance level of 
0.01 and a lag of 3.

## ``decision_modus`` - A function to get a decision (modus) from a vector of classes \label{section_decision_modus}

Several approaches for machine learning exist. It is a very rich topic. One of them is the supervised 
learning, where the aim is to infer a function from labeled training data. 
Labeled training data are for example created by a human. As described in the 
sections before, amplification curves can be labeled as negative, ambiguous or 
positive for example (Figure \ref{htPCR_nap}) were taken from the ``htPCR`` 
dataset (compare Figure \ref{samples_of_qPCRs}B).


```{r, echo=FALSE, fig.cap="Sample data from the **htPCR** data set (*qpcR* package, [@Ritz2008]) with negative (black), ambiguous (red) or positive (green) amplification curves. The negative amplification curve has no sigmoid curvature, just a strong positive trend. The ambiguous amplification curve is not a perfect sigmoid curve, exhibits a positive slope in the background phase (cylce 1 $\rightarrow$ 5) and a low amplitude. In contrast, the positive amplification curve has a sigmoid curvature, starting with a background phase (cylce 1 $\rightarrow$ 5), an exponential phase (cylce 6 $\rightarrow$ 25) and a plateau phase (cylce 26 $\rightarrow$ 35). \\label{htPCR_nap}", fig.height=5.5, fig.width=11}
suppressMessages(library(qpcR))
matplot(htPCR[, 1], htPCR[, c(552, 512, 616)], xlab="Cycle", ylab="RFU", 
        main="htPCR dataset", type="l", lty=1, lwd=2)
legend("topleft", c(paste("negative ", colnames(htPCR)[552]), 
                    paste("ambiguos ", colnames(htPCR)[512]), 
                    paste("positive ", colnames(htPCR)[616])
                   ), col=1:3, pch=19, bty="n")
```

Provided that different users are given the task to rate the amplification 
curves, it is likely that the ambiguous amplification curve (`r 
colnames(htPCR)[512]`) in Figure \ref{htPCR_nap} will get another rating 
(ambiguos $\rightarrow$ positive) from an inexperienced user. We performed 
exactly this kind of experiment with different users. For this task the 
amplification curve data were analyzed as described in @rodiger_chippcr:_2015 
(supplementary information).

The following table gives the first five and the last five rows for the rating 
of the amplification curves for the ``htPCR`` data set (in total `r 
ncol(qpcR::htPCR)-1` amplification curves). All curves were rated randomly three 
times. All raw data a contained in the CSV file `decision_res_htPCR.csv` as 
part of the ``PCRedux`` package.

```{r, eval=TRUE, echo=TRUE}
library(PCRedux)
suppressMessages(library(data.table))
library(magrittr)
filename <- system.file("decision_res_htPCR.csv", package="PCRedux")
decision_res_htPCR <- fread(filename) %>% as.data.frame()
head(decision_res_htPCR)
tail(decision_res_htPCR)
```

As seen in the example the rating for the same amplification curves vary (e.g., 
row 1 "P01.W01"), while others are in agreement (e.g., row 8856 "P95.W94"). 
``decision_modus`` is a function that can be used to find the most frequent 
(modus) decision in such a data set. The classes were defined by the user 
(e.g., "a", "n", "y" -> "ambiguous", "negative", "positive"). This function is 
useful if large collections of varying decision (e.g., "a", "a", "a", "n", 
"n") need to be condensed to a single decision (3 x "a", 2 x "n" $\rightarrow$ 
"a").

We used the ``decision_modus`` function on the `decision_res_htPCR.csv` data 
set with all the ratings (column 2 to 4) and determined the modus of each 
line.

```{r, eval=TRUE, echo=TRUE}
# Use decision_modus to go rowise through all human ratings

dec <- lapply(1L:nrow(decision_res_htPCR), function(i) {
decision_modus(decision_res_htPCR[i, 2:4])
}) %>% unlist

names(dec) <- decision_res_htPCR[, 1]

# Show statistic of the decisions
summary(dec)
```

```{r, echo=FALSE, fig.cap="Summary of all sample data from the **htPCR** data set (*qpcR* package, [@Ritz2008]) with negative (black), ambiguous (red) or positive (green) amplification curves. \\label{htPCR_nap}", fig.height=5.5, fig.width=11}

# Plot the Frequencies of the decisions
barplot(table(dec), xlab="Decision", ylab="Frequency", 
        main="Amplification curves rated by humans\n htPCR data set", col=c(2,3,1))
```

Another usage of the ``decision_modus`` function is by setting the `max_freq` 
parameter to `FALSE`. This option gives the counts of all decisions.

```{r, eval=TRUE, echo=TRUE}
# Decisions for sample P01.W06
decision_modus(decision_res_htPCR[which(decision_res_htPCR[["sample"]] == "P01.W06"), 
               2:4], max_freq=FALSE)

# Decisions for sample P01.W13
decision_modus(decision_res_htPCR[which(decision_res_htPCR[["sample"]] == "P01.W13"), 
               2:4], max_freq=FALSE)
```


```{r, eval=TRUE, echo=TRUE}
# Load the human rated data sets decision_res_htPCR.csv and the reevaluation 
# decision_res_htPCR_reevaluation.csv thereof.
# Load the decision_res_htPCR.csv data set
filename <- system.file("decision_res_htPCR.csv", package="PCRedux")
decision_res_htPCR <- as.data.frame(fread(filename))

# Load the decision_res_htPCR_reevaluation.csv data set
filename <- system.file("decision_res_htPCR_reevaluation.csv", package="PCRedux")
decision_res_htPCR_reevaluation <- as.data.frame(fread(filename))

# Show the first five elements of both data sets
head(decision_res_htPCR)
head(decision_res_htPCR_reevaluation)

# Renanme the columns to give them the same pattern
pattern <- c("sample", "test.result.1", "test.result.2", "test.result.3", "conformity")
colnames(decision_res_htPCR_reevaluation) <- pattern
colnames(decision_res_htPCR) <- pattern

# Merge the two data sets based on the sample name (amplification curve).
decision_res_htPCR_merged <- merge(decision_res_htPCR[, -5], 
                                   decision_res_htPCR_reevaluation[, -5], 
                                   by="sample")

# Show the first five elements of the merged data sets
head(decision_res_htPCR_merged)

```
Plot everything.

```{r, echo=TRUE, fig.cap="decision plot. \\label{decision_plot}", fig.height=5.5, fig.width=11}

par(mfrow=c(1,3))


data <- decision_res_htPCR_merged
decision_plot <- function(data, name="", rating_columns=2:4){
    dec <- lapply(1L:nrow(data), function(i) {
        decision_modus(data[i, c(rating_columns)])
    }) %>% unlist
    barplot(table(dec), xlab="Decision", ylab="Frequency", 
            main=paste("Amplification curves rated by humans\n htPCR data set", 
                       name), col=c(2,3,1), ylim=c(0,6000))
}

decision_plot(decision_res_htPCR, name="", rating_columns=2:4)
decision_plot(decision_res_htPCR_reevaluation, name="", rating_columns=2:4)
decision_plot(decision_res_htPCR_merged, name="merged", rating_columns=2:7)
```

## ``earlyreg`` - A function to calculate the slope and intercept of an amplification curve data from a quantitative PCR experiment \label{section_earlyreg}

There are numerous segments of an amplification curve, that are potentially 
useful to extract a feature for an amplification curve classification. Figure 
\ref{earlyreg_slopes} shows the the first 500 amplification curves from the 
``htPCR`` data set. Besides the fact that the data stem from a high-throughput 
experiment they exhibit interesting characteristics. Some of the amplification 
curves obviously deviate from the curvature as expected from a sigmoid model. In 
particular, some of the amplification curve just have a positive (linear) trend 
without typical infliction points. Others show a step signal increase even in 
the background phase (cycle 1 to 10) (compare @rodiger_chippcr:_2015, Figure 
S12). The slope and the intercept in background region is an interesting target 
for a classification.

@gunay_machine_2016 used a "modified sigmoid function for CT [Cq] prediction". The 
fundamental assumption of this approach is, that such a model can be applied to 
any data set. There are several reason to why such assumption is flawed. For example is shown in section \ref{section_hookreg} about the ``hookreg`` and ``hookregNL`` functions. There it becomes evident that a three parameter model fit will give wrong 
predictions. Moreover, non-linear function tend to fit even noise (Figure \ref{curve_fit_fail}).


```{r, echo=TRUE, fig.cap="Sample data from the **testdat** data set (*qpcR* package, [@Ritz2008]) with negative and positive amplification curves. \\label{curve_fit_fail}", fig.height=7, fig.width=7}
# Load the drc package for the five-parameter log-logistic fit.
library(drc)
library(chipPCR)

# Load the testdat dataset from the qpcR package without 
# loading the entire package.
data(testdat, package="qpcR")

# Arrange graphs in an orthogonal matrix and set the plot parmeters.
par(mfrow = c(2,2))

# Apply the the amptester function to all amplification curve data
# and write the results to the main of the plots.

curve_data_columns <- c(2,4,9,20)

for(i in 1L:length(curve_data_columns)) {
    res.ampt <- suppressMessages(amptester(testdat[, curve_data_columns[i]]))
    
    # Make a logical connection by two tests (shap.noisy, lrt.test and 
    # tht.dec) of amptester to decide if an amplification reaction is 
    # positive or negative.
    decision <- ifelse(!res.ampt@decisions[1] &&
                       res.ampt@decisions[2] && 
                       res.ampt@decisions[4], 
                       "positive", "negative")
    
    plot(testdat[, 1], testdat[, curve_data_columns[i]], 
         xlab = "Cycle", ylab = "RFU", pch = 19, main = "")
    mtext(LETTERS[i], cex = 2, side = 3, adj = 0, font = 2)
    legend("bottomright", paste0(colnames(testdat)[curve_data_columns[i]], 
                                 "\nDecision: ", decision), 
           bty="n", cex=1.25, col="red")
    # Use the drm function with a five-parameter log-logistic fit model.

        try(lines(predict(drm(testdat[, curve_data_columns[i]] ~ testdat[, 1], 
                          fct = LL.3())), col = 2), silent=TRUE)

        try(lines(predict(drm(testdat[, curve_data_columns[i]] ~ testdat[, 1], 
                          fct = LL.4())), col = 3), silent=TRUE)
 }
```

We suggest using the slope and the intercept for the differentiation of the 
different amplification curves. Therefore, we developed the function 
``earlyreg``. This wrapper function uses a MM-type estimator for a robust linear 
regression [@todorov_object-oriented_2009]. ``earlyreg`` ignores the first cycle 
value and includes by default the next six cycle-dependent amplitude values to 
perform the regression on them. For the analysis we used arbitrarily the first 
500 amplification curves of the ``htPCR`` dataset.

The following example illustrates a potential use of the ``earlyreg`` function 
on the **htPCR** data set. Figure \ref{earlyreg_slopes}A shows all amplification 
curves. Figure \ref{earlyreg_slopes}B shows the first ten cycles only.

```{r, echo=TRUE, fig.cap="Amplification curves from the **htPCR** data set (*qpcR* package). The amplification curves show different slopes in the early phase (cycle 1 to 10) of the qPCR. \\label{earlyreg_slopes}", fig.height=5.5, fig.width=11}
par(bty="n", font=2, font.axis=2, las=1, cex.axis=1.3, cex.lab=1.3, lwd=1.3, 
    bg="white", oma=c(.5,.5,.5,.5))
data <- htPCR[, 1:501]
curve_colors <- c(rainbow(ncol(data)-1, alpha=.5))
range <- 1:10

par(mfrow=c(1,2), las=0, bty="n", oma=c(1,1,1,1))

matplot(data[, 1], data[, -1], col= curve_colors, pch=19, lty=1, type="l", 
        xlab="Cycle", ylab="RFU", main="htPCR dataset")
mtext("A", cex = 2, side = 3, adj = 0, font = 2)

matplot(data[range, 1], data[range, -1], col= curve_colors, pch=19, lty=1, 
        type="l", xlab="Cycle", ylab="RFU", 
        main="htPCR dataset\n Cycle 1 to 10")
mtext("B", cex = 2, side = 3, adj = 0, font = 2)
```
Next we calculated of the slope and intercept with the ``earlyreg`` function 
for 
the first six cycles of the qPCR data. The results were subjected to a cluster 
analysis by k-means clustering. For the analysis we focused on the slope, 
because the intercept is representing the background fluorescence in the early 
phase of the qPCR. As shown in figure \ref{earlyreg_results}B exhibited the 
density function of all intercept values a mono-modal shaped bell curve. In 
contrast, the density function of all slope values had a bi-modal shaped bell 
curve (figure \ref{earlyreg_results}C). Therefore, the slope appears to be an 
indicator for differences between the qPCR amplification curves.

The cluster analysis by k-means (k=3, assumed weak slope, moderate slope, strong 
slope) on the slope data was plotted. The clusters grouped the amplification 
curve data in curves with no-background region and high slope, amplification 
curves with no-background region and a weaker slope and amplification curves 
with background region and a weak slope (figure \ref{earlyreg_results}D-F).

To conclude, this approach might be of use for the classification of 
amplification curves.

```{r, echo=TRUE, fig.cap="\\label{earlyreg_results}Clusters of samples according to their slope and intercept. The **htPCR** data set (*qpcR* package) was used. The amplification curves show different slopes in the early phase (cycle 1 to 10) of the qPCR. Both the slope and the intercept were used for a cluster analysis (k-means).", fig.height=7, fig.width=7}
# Normalize each amplification curve to their 0.99 percentile and use the 
# earlyreg function to determine the slope and intercept of the first 
# 6 cycles

res_slope_intercept <- do.call(rbind, lapply(2L:ncol(data), function(i){
                     earlyreg(x=data[, 1], y=data[, i], range=7, normalize=TRUE)
                    }))
# Label the sample with their original names
rownames(res_slope_intercept) <- colnames(htPCR)[2:ncol(data)]
# Use k-means for cluster analysis
res_kmeans <- kmeans(res_slope_intercept[, "slope"], 3)$"cluster"

par(mfrow=c(2,3))
plot(res_slope_intercept, col=res_kmeans)
mtext("A", cex = 2, side = 3, adj = 0, font = 2)

plot(density(res_slope_intercept[, "intercept"]), main="Intercept")
mtext("B", cex = 2, side = 3, adj = 0, font = 2)

plot(density(res_slope_intercept[, "slope"]), main="Slope")

for(i in unique(res_kmeans)) {
    abline(v=min(res_slope_intercept[res_kmeans==i, "slope"]), col=i)
}
mtext("C", cex = 2, side = 3, adj = 0, font = 2)

matplot(data[, 1], data[, which(res_kmeans==1)+1], col= curve_colors, pch=19, lty=1, 
 type="l", xlab="Cycle", ylab="RFU", 
 main="htPCR dataset (1-500)\n Cluster 1", ylim=c(min(data[, -1]), 0.4))
mtext("D", cex = 2, side = 3, adj = 0, font = 2)

matplot(data[, 1], data[, which(res_kmeans==2)+1], col= curve_colors, pch=19, lty=1, 
 type="l", xlab="Cycle", ylab="RFU", 
 main="htPCR dataset (1-500)\n Cluster 2", ylim=c(min(data[, -1]), 0.4))
mtext("E", cex = 2, side = 3, adj = 0, font = 2)

matplot(data[, 1], data[, which(res_kmeans==3)+1], col= curve_colors, pch=19, lty=1, 
 type="l", xlab="Cycle", ylab="RFU", 
 main="htPCR dataset (1-500)\n Cluster 3", ylim=c(min(data[, -1]), 0.4))
mtext("F", cex = 2, side = 3, adj = 0, font = 2)
```

## ``head2tailratio`` - A function to calculate the ratio of the head and the tail of a quantitative PCR amplification curve\label{section_head2tailratio}


The ``head2tailratio`` function calculates the ratio of the head and the tail of 
a quantitative PCR amplification curve. The begin of the amplification curve 
(head, background region) and the end of the amplification curve (tail, plateau 
region) are in positive amplification. Therefore, these segments are potentially 
useful to extract a feature for an amplification curve classification.

```{r, echo=TRUE, fig.cap="\\label{dil4reps94_head2tailratio}Some text here.", fig.height=5.5, fig.width=11}
suppressMessages(library(qpcR))


colors <- c("#FF00004D", "#80FF004D", "#00FFFF4D", "#8000FF4D")

data <- dil4reps94[, c(1, 2, 98, 194,290)]

res_head2tailratio <- lapply(2L:ncol(data), function(i) {
    head2tailratio(y=data[, i], normalize=TRUE, slope_normalizer=TRUE, 
                   verbose=TRUE)
})

data_normalized <- cbind(data[, 1], 
              sapply(2L:ncol(data), function(i){
                  data[, i] / quantile(data[, i], 0.999)
            })
             )

par(mfrow=c(1,2), las=0, bty="n", oma=c(1,1,1,1))

matplot(data_normalized[, 1], data_normalized[, -1], 
        xlab="Cycle", ylab="normalized RFU", main="dil4reps94 dataset\nsubset", 
        type="l", lty=1, lwd=2, col=colors
       )
for(i in 1L:(ncol(data_normalized)-1)) {
    points(res_head2tailratio[[i]]$x_roi, res_head2tailratio[[i]]$y_roi, 
           col=colors[i], pch=19, cex=1.5)
    abline(res_head2tailratio[[i]]$fit, col=colors[i], lwd=2)
}

mtext("A", cex = 2, side = 3, adj = 0, font = 2)

colors <- c(rep("#FF00004D", 94),
            rep("#80FF004D", 94),
            rep("#00FFFF4D", 94),
            rep("#8000FF4D", 94)
            )

matplot(dil4reps94[, 1], dil4reps94[, -1], xlab="Cycle", ylab="RFU", 
        main="dil4reps94 dataset", type="l", lty=1, lwd=2, col=colors)
mtext("B", cex = 2, side = 3, adj = 0, font = 2)

```

## ``hookreg`` and ``hookregNL``\label{section_hookreg}



## ``mblrr`` - A function perform the Quantile-filter based Local Robust Regression \label{section_mblrr}

```{r, echo=TRUE, fig.cap="\\label{earlyreg_mblrr}Some text here.", fig.height=7, fig.width=7}
suppressMessages(library(qpcR))
par(mfrow=c(3,2))
data <- htPCR[, c(1, 20, 500, 3000, 6000, 6500, 8000)]
for(i in 2:ncol(data)){
        x <- data[, 1]
        y_tmp <- data[, i]/quantile(data[, i], 0.999)
        res_q25 <- y_tmp < quantile(y_tmp, 0.25)
        res_q75 <- y_tmp > quantile(y_tmp, 0.75)
        res_q25_lm <- try(suppressWarnings(lmrob(y_tmp[res_q25] ~ x[res_q25])), 
                          silent=TRUE)
        res_q75_lm <- try(suppressWarnings(lmrob(y_tmp[res_q75] ~ x[res_q75])), 
                          silent=TRUE)
        
        plot(x, y_tmp, xlab="Cylce", ylab="RFU (normalized)", 
             main=colnames(data)[i], 
             type="b", pch=19)
        abline(res_q25_lm, col="red")
        points(x[res_q25], y_tmp[res_q25], cex=2.5, col="red")
        abline(res_q75_lm, col="green")
        points(x[res_q75], y_tmp[res_q75], cex=2.5, col="green")
    }
```

```{r, eval=TRUE, echo=TRUE}
res_mblrr <- do.call(cbind, lapply(2L:ncol(testdat), function(i) {
                mblrr(x=testdat[, 1], y=testdat[, i], 
                      normalize=TRUE) %>% data.frame
}))
colnames(res_mblrr) <- colnames(testdat)[-1]

a <- res_mblrr %>% t
```

## ``pcrfit_single`` \label{section_pcrfit_single_pcrfit_parallel}

The functionality is provided by packages developed by others and us. The 
following table gives an overview and some comments on the function as used by 
the `pcrfit_single` function.

Function                 | Package   | Source                   | Details
-------------------------|-----------|--------------------------|------------------------------
``bcp``                  | `bcp`     | @erdman_bcp:_2007        | Performs a change point analysis based on a Bayesian approach.
``amptester``            | `chipPCR` | @rodiger_chippcr:_2015   | Test with different statistical methods, if the amplification is positive of negative.
``bg.max``               | `chipPCR` | @rodiger_chippcr:_2015   | Calculates estimates for the start and end of the background region and the end of the exponential amplification reaction. The estimates of the ``bg.max`` function are normalized to the total cycle number.
``e.agglo``              | `ecp`     | @james_ecp:_2013         | Performs a multiple change point analysis based on agglomerative hierarchical estimation.
``diffQ``                | `MBmca`   | @roediger_RJ_2013        |
``diffQ2``               | `MBmca`   | @roediger_RJ_2013        |
``mcaPeaks``             | `MBmca`   | @roediger_RJ_2013        |
``head2tailratio``       | `PCRedux` |                          | Calculates an estimate for the head to tail fluorescence, normalized to the slope. See Section \ref{section_head2tailratio}.
``earlyreg``             | `PCRedux` |                          | Calculates the slope and intercept of an amplification curve data from a quantitative PCR experiment. See Section \ref{section_earlyreg}.
``hookreg`` & ``hookregNL`` | `PCRedux` |                          |
``autocorrelation_test`` | `PCRedux` |                          | Test for an autocorrelation of amplification curve data from a quantitative PCR experiment. See Section \ref{section_autocorrelation_test}.
``mblrr``                | `PCRedux` |                          | See Section \ref{section_mblrr}.
``polyarea``             | `pracma`  | @pracma                  | Calculates the area of a polygon defined by the vertices with coordinates of the cycles and fluorescence, based on the Gauss polygon area formula.
``efficiency``           | `qpcR`    | @Ritz2008                | Determine the Cq (first derivative maximum, second derivative maximum) and other parameters.
``LRE``                  | `qpcR`    | @Ritz2008                | Calculates the qPCR efficiency by the 'linear regression of efficiency' method.
``sliwin``               | `qpcR`    | @Ritz2008                | Calculates the qPCR efficiency by the 'window-of-linearity' method.
``takeoff``              | `qpcR`    | @Ritz2008                | Calculates the qPCR takeoff point.

An example is given for the internal parameter ``diffQ2_slope`` which is calculated 
from the slope determined by a linear model of the data points from the minimum 
and maximum of the second derivative. The coordinates of the minimum and the maximum
were determined as described in @roediger_RJ_2013. In the following example, the data 

```{r, echo=TRUE, fig.cap="\\label{diffQ2_slope}Some text here.", fig.height=5, fig.width=11}
# Load example data (sample F6.1) from the testdat data set
library(qpcR)
library(magrittr)
# Load MBmca package to calculate the minimum and the maximum of the second 
# derivative

library(MBmca)
data <- testdat[, c(1,22)]

# Calculate the minimum and the maximum of the second 
# derivative and assign it to the object res_diffQ2

res_diffQ2 <- suppressMessages(diffQ2(data, plot=FALSE, fct=min))

# Build a linear model from der second derivative of res_diffQ2
res_diffQ2_lm <- lm(res_diffQ2[["yTm1.2.D2"]] ~ res_diffQ2[["xTm1.2.D2"]])

par(mfrow=c(1,2))

data %>% plot(., xlab="Cycle", ylab="RFU", main="F6.1 (testdat)", type="l", 
                   lty=1, lwd=2, col="black")
abline(v=res_diffQ2[["xTm1.2.D2"]], col="grey", lwd=2)
mtext("A", cex = 2, side = 3, adj = 0, font = 2)


plot(res_diffQ2[["xTm1.2.D2"]], res_diffQ2[["yTm1.2.D2"]], pch=19, cex=1.5,
     xlab="Cycle", ylab="dd(RFU)/d(T)", 
     main="minimum and the maximum\n of second derivative")
abline(res_diffQ2_lm, col="blue", lwd=2)
legend("bottomright", paste("Slope: ", coefficients(res_diffQ2_lm)[2]), bty="n")
mtext("B", cex = 2, side = 3, adj = 0, font = 2)
```

The ``pcrfit_single`` function is a powerful function, which calculates numerous 
features from amplification curve data. This comes at a cost, since several 
internal functions are computational very intensive [@porzelius_easier_2009, 
@schmidberger_2009]. Recent information about parallelization in **R** is available 
from @eddelbuettel_cran_2017. An example is the ``pcrfit`` 
function from the `qpcR` package. This function fits and optimizes eight 
non-linear (sigmoid) models to the amplification curve data.

The code block below shows an example for a parallelized version of 
``pcrfit_single``. We call this function ``pcrfit_parallel`` is presumably 
beneficial for users who wish to calculate the features of a large amplification 
curve data set. We found that this function works on Linux systems. On Windows 
systems we received selectively error messages. ``pcrfit_parallel`` makes use of 
parallelized code to make use of multi-core architectures. In this function we 
import from the `parallel` package the ``detectCores`` function. This function 
determines the number of available cores. Function from the `foreach` package 
(e.g., `%dopar%`, `foreach`) are used for the further organization of the CPU 
usage. The ``pcrfit_single`` performs the analysis for a single process.

* The parameter `data` is the data set containing the cycles and fluorescence amplitudes.
* The parameter `n_cores` defines the numbers of cores that should be left unused by this function. By default ``pcrfit_parallel`` is using only one core (`n_cores = 1`). `n_cores = "all"` uses all available cores. 

The output of the ``pcrfit_parallel`` function is similar to the ``pcrfit_single`` function.


```{r, echo=TRUE, eval=FALSE}
# Copy and paste the code to an R console to evaluate it

library(parallel)
library(doParallel)

pcrfit_parallel <- function(data, n_cores = 1) {
    # Determine the number of available cores and register them
    if(n_cores == "all")
        n_cores <- detectCores() 
        
        registerDoParallel(n_cores)
        
        # Prepare the data for further processing
        # Normalize RFU values to the alpha percentile (0.999)
        cycles <- data.frame(cycles=data[, 1])
        data_RFU <- data.frame(data[, -1])
        data_RFU_colnames <- colnames(data_RFU)
        data_RFU <- sapply(1L:ncol(data_RFU), function(i) {
            data_RFU[, i] / quantile(data_RFU[, i], 0.999, na.rm=TRUE)
        })
        colnames(data_RFU) <- data_RFU_colnames
        
        # just to shut RCHeck for NSE we define ith_cycle
        ith_cycle <- 1
        
        run_res <- foreach::foreach(ith_cycle = 1L:ncol(data_RFU), 
                           .packages=c("bcp", "changepoint", "chipPCR", "ecp", "MBmca",
                                       "PCRedux", "pracma", "qpcR", "robustbase", 
                                       "zoo"),
                           .combine = rbind) %dopar% {
                               suppressMessages(pcrfit_single(data_RFU[, ith_cycle]))
                           }
                           
                           
                           res <- cbind(runs = colnames(data_RFU), run_res)
                           
                           rownames(res) <- NULL
                           
                           res
                           
}

# Calculate curve features of an amplification curve data. Note that not all 
# available CPU cores are used. If need set "all" to use all available cores.
# In this example the testdat dataset from the qpcR package is used.
# The samples F1.1 and F1.2 are positive amplification curves. The samples 
# F1.3 and F1.4 are negative.
 
library(qpcR)
res_pcrfit_parallel <- pcrfit_parallel(testdat[, 1:5])
res_pcrfit_parallel
```

## ``performeR`` \label{section_performeR}

Statistical modeling and machine learning can be powerful but expose the user to 
the risk of introducing unexpected biases. This may lead to an overestimation of 
the performance.The assessment of the performance by the sensitivity (percentage 
of true decisions that are identified) and specificity (percentage of negative 
decision that are correctly identified) is important to characterize the 
performance of a classifier or screening test [@james_introduction_2013].


Abbreviations: TP, true positive; FP, false positive; TN, true negative; FN, false negative

Measure                                    | Formula
-------------------------------------------|-----------------------------------------------
Sensitivity - TPR, true positive rate      | $TPR = \frac{TP}{TP + FN}$
Specificity - SPC, true negative rate      | $SPC = \frac{TN}{TN + FP}$
Precision - PPV, positive predictive value | $PPV = \frac{TP}{TP +  FP}$
Negative predictive value - NPV            | $NPV = \frac{TN}{TN + FN}$
Fall-out, FPR, false positive rate         | $FPR = \frac{FP}{FP + TN} = 1 - SPC$
False negative rate - FNR                  | $FNR = \frac{FN}{TN + FN} = 1 - TPR$
False discovery rate - FDR                 | $FDR = \frac{FP}{TP + FP} = 1 - PPV$
Accuracy - ACC                             | $ACC =  \frac{(TP + TN)}{(TP + FP + FN + TN)}$
F1 score - F1                              | $F1 = \frac{2TP}{(2TP + FP + FN)}$
Matthews correlation coefficient - MCC     | $MCC = \frac{(TP*TN - FP*FN)}{\sqrt{(TP+FP)*(TP+FN)*(TN+FP)*(TN+FN)}}$
Likelihood ratio positive - LRp            | $LRp = \frac{TPR}{1-SPC}$
Cohen''s kappa (binary classification)     | $\kappa=\frac{p_{0}-p_{c}}{1-p_{0}}$


## ``qPCR2fdata`` \label{section_qPCR2fdata}

``qPCR2fdata`` is a helper function to convert qPCR data to the functional 
`fdata` class as introduced by @Febrero_Bande_2012. This function prepares the 
data for further analysis, which includes utilities for functional data 
analysis. For example, it this can be used to determine the similarity measures 
between amplification curves shapes by the Hausdorff distance 
[@charpiat_shape_2003].

The ``qPCR2fdata`` function takes a data set containing the amplification cycles 
(first column) and the fluorescence amplitudes (subsequent columns) as input. 
Since noise and missing values may effect the analysis adversely we integrated 
an instance of ``CPP`` function from the `chipPCR` package 
[@rodiger_chippcr:_2015] in `qPCR2fdata`.

The following example illustrates the usage for the `testdat` data set.

```{r echo=TRUE, message=FALSE, warning=FALSE}
# Calculate slope and intercept on noise (negative) amplification curve data
# for the last eight cycles.
# Load additional packages for data and pipes.
library(qpcR)
library(chipPCR)
library(fda.usc)
library(magrittr)

# Convert the qPCR data set to the fdata format
# Use unprocessed data from the testdat data set
res_fdata <- qPCR2fdata(testdat)

# Use preprocessed data (preprocess=TRUE) from the testdat data set
res_fdata_preprocessed <- qPCR2fdata(testdat, preprocess=TRUE)

# Extract column names and create rainbow color to label the data
res_fdata_colnames <- testdat[-1] %>% colnames()
data_colors <- rainbow(length(res_fdata_colnames), alpha=0.5)
```

```{r, echo=TRUE, fig.cap="\\label{qPCR2fdata}Some text here.", fig.height=11, fig.width=11}
# Plot the converted qPCR data
par(mfrow=c(2,2))
res_fdata %>% plot(., xlab="Cycle", ylab="RFU", main="testdat", type="l", 
                   lty=1, lwd=2, col=data_colors)
legend("topleft", as.character(res_fdata_colnames), pch=19, 
       col=data_colors, bty="n", ncol=2)
mtext("A", cex = 2, side = 3, adj = 0, font = 2)
# Calculate the Hausdorff distance (fda.usc) package and plot the distances
# as clustered data.

res_fdata_hclust <- metric.hausdorff(res_fdata)
plot(hclust(as.dist(res_fdata_hclust)), main="Clusters of the amplification\n
curves as calculated by the Hausdorff distance")
mtext("B", cex = 2, side = 3, adj = 0, font = 2)
rect(0.5,-3,12.25,0.5, border = "red")
text(7, 1, "negative", col="red")
rect(12.5,-3,24.5,0.5, border = "green")
text(14, 1, "positive", col="green")

# Repeat the plot and the cluster analysis for the preprocessed data
res_fdata_preprocessed %>% plot(., xlab="Cycle", ylab="RFU", main="testdat", type="l", 
                   lty=1, lwd=2, col=data_colors)
legend("topleft", as.character(res_fdata_colnames), pch=19, 
       col=data_colors, bty="n", ncol=2)
mtext("C", cex = 2, side = 3, adj = 0, font = 2)
# Calculate the Hausdorff distance (fda.usc) package and plot the distances
# as clustered data from preprocessed amplification curves.

res_fdata_hclust_preprocessed <- metric.hausdorff(res_fdata_preprocessed)
plot(hclust(as.dist(res_fdata_hclust_preprocessed)), 
     main="Clusters of the preprocessed amplification\n
curves as calculated by the Hausdorff distance")
rect(0.5,-3,12.25,0.5, border = "red")
text(7, 1, "negative", col="red")
rect(12.5,-3,24.5,0.5, border = "green")
text(14, 1, "positive", col="green")
mtext("D", cex = 2, side = 3, adj = 0, font = 2)
```

## ``visdat_pcrfit`` - A function to visualize the content of data from an analysis with the `pcrfit_single` function  \label{section_visdat_pcrfit_parallel}

The ``visdat_pcrfit`` function makes use of the ``vis_dat`` function from the `visdat` package by @Tierney2017.
The samples "A01", "A06" and "A10" from the ``C126EG685`` of the ``chipPCR`` package were analyzed.

```{r, eval=TRUE, echo=TRUE}
# Calculate curve features of an amplification curve data set. Note that not all 
# available CPU cores are used. If need set "all" to use all available cores.
library(chipPCR)
dat <- C126EG685[, c(2, 7, 11)]

res_1 <- cbind(runs="A01", pcrfit_single(dat[, 1]))
res_2 <- cbind(runs="A06", pcrfit_single(dat[, 2]))
res_3 <- cbind(runs="A10", pcrfit_single(dat[, 3]))

res <- rbind(A01=res_1, A06=res_2, A10=res_3)
```
Finally the data were visualized with the ``visdat_pcrfit`` function. In this example the static plot is shown. It is also possible to run the function in an interactive mode by setting the parameter "interactive = TRUE". In this case starts a `ggplotly` instance.

```{r, echo=TRUE, fig.cap="visdat_pcrfit. \\label{visdat_pcrfit_plot}", fig.height=11, fig.width=11}
# Show all results in a plot. Not that the interactive parameter is set to 
# FALSE.

visdat_pcrfit(res, type="all", interactive=FALSE)
```

# Summary and conclusions  \label{section_Summary_and_conclusions}

The `PCRedux` enables the user to extract features from amplification curve 
data. We implemented numerous features that can be extracted from the 
amplification curve data. Some of them have not been described in the 
literature. We consider `PCRedux` as enabling technology for further research. 
For example, the proposed features can be used for machine learning applications 
or quality assessment of data.

The ``pcrfit_helper`` function is a wrapper that can be extended if new features 
emerge.

Of note, we would like to emphasis that the functionality of this package is not 
limited to amplification curve data from qPCR experiments. As stated before, 
amplification curves have a sigmoid curve shape. So do many other processes.

# References
