---
title: "PCRedux Package - An Overview"
author: "Stefan R&ouml;diger"
date: "`r Sys.Date()`"
output: 
  rmarkdown::pdf_document:
    number_sections: true
    toc: true
    toc_depth: 5
header-includes:
    - \usepackage[font={small}]{caption}
bibliography: "literature.bib"
classoption: a4paper
vignette: >
  %\VignetteIndexEntry{PCRedux package - an overview}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---



```{r, echo=FALSE}
knitr::opts_chunk$set(fig.lp = "", out.extra='', warnings = FALSE, message = FALSE)

amplification_curve_ROI <- "Phases of application curves as Region of Interest (ROI). For amplification curves, the fluorescence signal (RFU, relative fluorescence units) of the reporter dye is plotted against the cycle number. Positive amplification curves have the three ROIs: ground phase, exponential phase and plateau phase. These ROIs can be used to determine features such as the takedown point (`tdp`) or the standard deviation within the ground phase (`sd\\_bg`). The exponential range (red dots) is used to determine the Cq values and amplification efficiency (not shown). A linear regression model (red) can be used to calculate the slope in this region. B) PCRs without amplification reaction usually show a flat (non-sigmoides) signal. C) The exponential phase of PCR reactions can vary greatly depending on the DNA starting quantity and other factors. Amplification curves that appear in later cycles often have a lower slope in the exponential phase."

amplification_curve_ROI_short <- "Phases of application curves as Region of Interest (ROI)"


figure_quntifcation_points <- "Frequently used methods for the analysis of quantification points. A) The amplification curve is intersected by a grey horizontal line. This is the background signal (3$\\sigma$) determined from the \\textit{68-95-99.7 rule} from the fluorescence emission of cycles 1 to 10. The black horizontal line is the user-defined threshold (Ct value) in the exponential phase. Based on this, the cycle at which the amplification curve differs significantly from the background is calculated. B) The amplification curve can also be analyzed by fitting a multi-parametric model (black line, five parameters). The red line is the first derivative of the amplification curve with a maximum of 17.59 cycles. The first derivative maximum (`cpD1`) is used as a quantification point (Cq value) in some qPCR systems. The green line shows the second derivative of the amplification curve, with a maximum at 15.68 cycles a minimum at 19.5 cycles. The maximum of the second derivative (`cpD2`) is used as Cq value in many systems. The blue line shows the amplification efficiency estimated from the trajectory of the exponential region. The `Eff` value of 1.795 means that the amplification efficiency is approximately 89\\%. `cpDdiff` is the difference between the Cq values calculated from the first and second derivative maximum ($cpDdiff = |cpD1 - cpD2|$)."

figure_quntifcation_points_short <- "Frequently used methods for the analysis of quantification points"

figure_curve_classification <- "Variations of the classification of amplification curves. A prerequisite for the development of machine-learning models is the availability of manually classified amplification curves. Amplification curves (n = 8858) from the `htPCR` data set have been classified by one user eight times at different points over time (classes: ambiguous (a), positive (y) or negative (n)). During this process, the amplification curves were presented in random order. The example shows that different (subjective) class mappings may occur for the same dataset. While only a few amplification curves were classified as negative in the first three classification cycles (A-C), their proportion increased almost tenfold in later classification cycles (D-H)."

figure_curve_classification_short <- "Variations of the classification of amplification curves"


htPCR_nap <- "Examples of negative, ambiguous and positive amplification curves. A negative (black), ambiguous (red) and positive (green) amplification curve were selected from the `htPCR` data set. The negative amplification curve is non-sigmoid and has a positive trend. The ambiguous amplification curve is similar to a sigmodic amplification curve, but shows a positive slope in ground phase (cycle 1 $\\rightarrow$ 5). The positive amplification curve (green) is sigmoid. It starts with a flat baseline (cycle 5 $\\rightarrow$ 25). This is followed by the exponential phase (cycle 5 $\\rightarrow$ 25) and ends in a flat plateau phase (cycle 26 $\\rightarrow$ 35)."

htPCR_nap_short <- "Examples of negative, ambiguous and positive amplification curves"


htPCR_nap_frequency <- "Frequency of amplification curve classes and conformity in the `htPCR` dataset. The `htPCR` dataset was classified by hand eight times. Due to the unusual amplification curve shape and input errors during classification, many amplification curves were classified differently. A) Frequency of negative (black), ambiguous (red) and positive (green) amplification curves in the `htPCR` dataset. The combined number of ambiguous and negative amplification curves appears to be higher, than the number of positive amplification curves. B) The number of observations where all classification cycles made the same decision (conformity == TRUE) accounts for only 5\\% of the total number of observations. TRUE, all classes of the amplification curve matched. FALSE, at least one in eight observations had a different class."

htPCR_nap_short_frequency <- "Frequency of amplification curve classes and conformity in the `htPCR` dataset."


qPCR2fdata <- "Shape-based grouping of amplification curves. A) The grouping of the amplification curves of the `testdat` data set (A) was based on the Hausdorff distance. B) The amplification curves were converted with the ``qPCR2fdata()`` function and the Hausdorff distance of the curves to each other was determined by a cluster analysis. There was no errors in distinguishing between negative (n) and positive (y) amplification curves."

qPCR2fdata_short <- "Shape-based grouping of amplification curves"


HCU32 <- "Clustering and variation analysis of amplification curves. The amplification curves of the 32HCU were processed with the ``qPCR2fdata()`` function and then processed by cluster analysis (Hausdorff distance). A) Amplification curves were plotted from the raw data. B) Overall, signal-to-noise ratios of the amplification curves between all cavities were similar. C) The Cq values and amplification efficiency (eff) were calculated using the ``efficiency(pcrfit())`` function (``qpcR`` package). The median Cq is shown as a vertical line. Cqs greater or less than 0,1 of Cq $\tilde{x}$ are marked with observation labels. D) The cluster analysis showed no specific pattern with respect to the amplification curve signals. It appears that the observations D1, E1, F1, F3, G3 and H1 differ most from the other amplification curves."

HCU32_short <- "Clustering and variation analysis of amplification curves"

curve_fit_fail <- "Incorrect model adjustment for amplification curves. A positive (black) and a negative amplification curve (red) were randomly selected from the `RAS002` dataset. The positive amplification curve has a baseline signal of about 2500 RFU and has a definite sigmoidal shape. The negative amplification curve has a baseline signal of approx. 4200 RFU, but only moderately positive slope (no sigmoidal shape). A logistic function with seven parameters (`l7`) has been fitted to both amplification curves. A Cq value of 25.95 was determined for the positive amplification curve. The negative amplification curve had a Cq value of 9.41. However, it can be seen that the latter model fitting is not appropriate for calculating a trustworthy Cq value. An automatic calculation without user control would give a false-positive result."

curve_fit_fail_short <- "Incorrect model adjustment for amplification curves"

plot_models <- "Frequencies of the fitted multiparametric models and Cq values. The amplification curves (n = 3302) of the `data\\_sample` dataset were analyzed with the ``encu()`` function. The amplification curves have been broken down according to their classes (negative: grey, positive: green). A) The optimal multiparametric model was selected for each amplification curve based on the Akaike information criterion. lNA stands for `no model` and l4 \\ldots l7 for a model with four to seven parameters. B) All Cq values were calculated from optimal multiparametric models. Cqs of positive amplification curves accumulate in the range between 15 and 30 PCR cycles (50\\%). For the negative amplification curves, the Cqs are distributed over the entire span of the cycles. Note: The Cqs of the negative amplification curves are false-positive!"

plot_models_short <- "Frequencies of the fitted multiparametric models and Cq values"

figure_cpD2_range <- "Location of the of the features `cpD2\\_range`, `bg.start`, `bg.stop` within an amplification curve. The minimum (cpD2m) and maximum (cpD2) of the second derivative were calculated numerically using the ``diffQ2()`` function. This function also returns the maximum of the first derivative (cpD1). The `cpD2\\_range` is defined as $cpD2\\_range = |cpD2 - cpD2m|$. Large `cpD2\\_range` values indicate a low amplification efficiency or a negative amplification reaction. The feature `bg.start` is an estimate for the end of the ground phase. `bg.start` is an approximation for the onset of the plateau phase."

figure_cpD2_range_short <- "Location of the of the features `cpD2\\_range`, `bg.start`, `bg.stop` within an amplification curve"


plot_dat_EffTop <- "Values of features calculated from negative and positive amplification curves. Amplification curves from the datasets `stepone\\_std`, `RAS002`, `RAS003`, `lc96\\_bACTXY`, `C126EG595` and `dil4reps94` were analyzed with the ``encu()`` function. These datasets contain positive and negative amplification curves.  Furthermore, the meta dataset contains amplification curves that exhibit a hook effect or non-sigmoid shapes, for instance. All amplification curves are manually classified. Altogether 626 positive and 317 negative amplification curves were included in the analysis. A) `eff`, optimized PCR efficiency found within a sliding window. B) `sliwin`, PCR efficiency by the ‘window-of-linearity’ method. C) `cpDdiff`, difference between the Cq values calculated from the first and the second derivative maximum. D) `loglin\\_slope`, slope from the cycle at the second derivative maximum to the second derivative minimum. E) `cpD2\\_range`, absolute value of the difference between the minimum and the maximum of the second derivative maximum. F) `top`, takeoff point. G) `f.top`, fluorescence intensity at takeoff point. H) `tdp`,  takedown point. I) `f.tdp`, fluorescence intensity at takedown point. J) `bg.stop`, estimated end of the ground phase. K) `amp.stop`, estimated end of the exponential phase. L) `convInfo\\_iteratons`, number of iterations until convergence."

plot_dat_EffTop_short <- "Analysis of location features"

loglin_slope <- "Concept of the `loglin\\_slope` feature. The algorithm determines the fluorescence values of the raw data at the approximate positions of the maximum of the first derivative, the minimum of the second derivative and the maximum of the second derivative, which are in the exponential phase of the amplification curve. A linear model is created from these parameter sets and the slope is determined. A) Positive amplification curves have a clearly positive slope. B) Negative amplification curves usually have a low, sometimes negative slope. The data were taken from the `RAS002` dataset."

loglin_slope_short <- "Concept of the `loglin\\_slope` feature"

plot_sd_bg <- "Standard deviation in the ground phase of various qPCR devices. The `sd\\_bg` feature was used to determine if the standard deviation between the thermo-cyclers and between positive and negative amplification curves was different. The standard deviation was determined from the fluorescence values from the first cycle to the takeoff point. If the takeoff point could not be determined, the standard deviation from the first cycle to the eighth cycle was calculated. The Mann-Whitney test was used to compare the medians of the two populations (y, positive; n, negative). The differences were significant for A) LC\\_480 (Roche), B) ABI\\_Prism\\_7700 (ABI), C) LC1.0 (Roche), E) CFX96 (Bio-Rad) and F) LC96 (Roche). The difference was not significant for D) StepOne (Thermo Fisher)."

plot_sd_bg_short <- "Standard deviation in the ground phase of various qPCR devices"

plot_bg_pt <- "Values of features calculated from negative and positive amplification curves. Using the ``encu()`` function, 626 positive and 317 negative amplification curves were analyzed from various datasets. The \\textit{metadata set} set also includes amplification curves with a hook effect or a non-sigmoid shape. A) `eff`, optimized PCR efficiency in a sliding window. B) `sliwin`, PCR efficiency according to the window-of-linearity method. C) `cpDdiff`, difference between the Cq values calculated from the first and the second derivative maximum. D) `loglin\\_slope`, slope from cycle at second derivative maximum to second derivative minimum. E) `cpD2\\_range`, absolute difference between the minimum and maximum of the second derivative. F) `top`, takeoff point. G) `f.top`, fluorescence intensity at takeoff point. H) `tdp`, takedown point. I) `f.tdp`, fluorescence intensity at the takedown point. J) `bg.stop`, estimated end of the ground phase. K) `amp.stop`, estimated end of the exponential phase. L) `convInfo\\_iteratons`, number of iterations until convergence when fitting a multiparametric model. The classes were compared using the Wilcoxon Rank Sum Test."

plot_bg_pt_short <- "Values of features calculated from negative and positive amplification curves"

plot_Logistic_Regression <- "Machine classification by means of binomial logistic regression using the `polyarea` feature. A) For the calculation of a binomial logistic regression model, the categorical response variable $Y$ (decision with classes: negative and positive) must be converted to a numerical value. With binomial logistic regression, the probability of a categorical response can be estimated using the $X$ predictor variable. In this example, the predictor variable `polyarea` is used. Grey measurement points (70\\% of the dataset) were used for training. Red dots represent the values used for testing. The regression curve of the binomial logistic regression is shown in blue. The grey horizontal line at 0.5 marks the threshold of probability above which it is determined whether an amplification curve is negative or positive. B) The performance indicators were calculated using the ``performeR()`` function. Sensitivity, TPR; Specificity, SPC; Precision, PPV; Negative prediction value, NPV; Fall-out, FPR; False negative rate, FNR; False detection rate, FDR; Accuracy, ACC; F1 score, F1; Matthews correlation coefficient, MCC, Cohens kappa (binary classification), kappa ($\\kappa$)."

plot_Logistic_Regression_short <- "Machine classification by means of binomial logistic regression using the `polyarea` feature"


statistical_methods_amptester <- "Analysis of amplification curves with the ``amptester()`` function. A \\& B) The threshold test (THt) is based on the Wilcoxon ranksum test and compares 20\\% of the fluorescence values of the ground phase with 15\\% of the plateau phase. In the example, a significant difference ($p = 0.000512$) was found for the positive amplification curve. However, this did not apply to the negative amplification curve ($p = 0.621$). C \\& D) A Q-Q diagram is used to graphically compare two probability distributions.  In this study the probability distribution of the amplification curve was compared with a theoretical normal distribution. The orange line is the theoretically normal quantil-quantile plot that passes through the probabilities of the first and third quartiles. The Shapiro-Wilk test (SHt) of normality checks whether the underlying measurement data of the amplification curve is significantly normally distributed. Since the p-value of the positive amplification curve ($7.09 e^{-9}$) is $\\alpha \\leq 5e^{-4}$, the null hypothesis is rejected. However, this does not apply to the negative amplification curve ($p = 0.895$). E \\& F) The linear regression test (LRt) calculates the coefficient of determination ($R^{2}$) using an ordinary least square regression where all measured values are integrated into the model in a cycle-dependent manner. Experience shows that the non-linear part of a amplification curve has a $R^{2}$ smaller than 0.8, which is also shown in the example."

statistical_methods_amptester_short <- "Analysis of amplification curves with the ``amptester()`` function"

figure_autocorrelation_tau <- "Effect of tau"

figure_autocorrelation_tau_short <- "Effect of tau"


autocorrelation <- "Autocorrelation analysis of the amplification curves of the `RAS002` data set. A) Display of all amplification curves of the data set `RAS002`. Negative amplification curves are shown in red and positive amplification curves in black. The `autocorrelation\\_test()` function was used to analyze all amplification curves. B) The density diagram of the autocorrelation of positive and negative amplification curves shows a bimodal distribution. C) The cdplot calculates the conditional densities of x based on the values of y weighted by the boundary distribution of y. The densities are derived cumulatively via the values of y. The probability that the decision is negative (n) when autocorrelation equals 0.85 is approximately 100\\%. D) Performance analysis using the ``performeR()`` function (see \\autoref{section_performeR} for details)."

autocorrelation_short <- "Autocorrelation analysis for amplification curves of the `RAS002` dataset)"


earlyreg_slopes <- "Analysis of the ground phase with the ``earlyreg()`` function . The amplification curves (`C127EGHP` data set, n = 64 amplification curves). This dataset consists of 32 samples, which were simultaneously monitored with the intercalator EvaGreen or hydrolysis probes . A) All amplification curves have slightly different slopes and intercepts in the first cycles of the ground phase (ROI: Cycles 1 to 8). Both the slope and the intercept of each amplification curve were used for a cluster analysis (k-means, Hartigan-Wong algorithm, number of centers \\textit{k = 2}). B) The amplification curves were assigned to five clusters, depending on their slope and their intersection (red, black). C) Finally, the clusters were associated to the detection chemistries (EvaGreen (EG) or hydrolysis probes (HP))."

earlyreg_slopes_short <- "Analysis of the ground phase with the ``earlyreg()`` function"


figure_head2tailratio <- "Ratio between the head and the tail of a quantitative PCR amplification curve. A) Plot of quantile normalized amplification curves from the `RAS002` dataset. Data points used in the head and and tail are highlighted by circles. The intervals for the Robust Linear Regression are automatically selected using the 25\\% and 75\\% quantiles. Therefore, not all data points are used in the regression model. The straight line is the regression line from the robust linear model. The slopes of the positive and negative amplification curves differ. B) Boxplot for the comparison of the $head/tail$ ratio. Positive amplification curves have a lower ratio than negative curves. The difference between the classes is significant."

figure_head2tailratio_short <- "Ratio between the head and the tail of a quantitative PCR amplification curve"

plot_mblrr <- "Robust local regression to analyze amplification curves. The amplification curves were arbitrarily selected from the `RAS002` dataset. Not the differences in slopes and intercepts (blue and orange lines and dots). The ``mblrr()`` function is presumably useful for datasets which are accompanied by noise and artifacts. m, slop; n, intercept."

plot_mblrr_short <- "Robust local regression to analyze amplification curves"

plot_FFTrees <- "Visualization of decisions in  Fast and Frugal Trees after data analysis of amplification curves via the ``mblrr()`` function. \\textbf{Top row} `Data`) Overview of the dataset, with displaying the total number of observations (N = 192) and percentage of positive (22\\%) and negative (78\\%) amplification curves. \\textbf{Middle row} `FFT \\#1 (of 6)`) Decision Tree with the number of observations classified at each level of the tree. For the analysis, six features (nBG, intercept of head region; mBG, slope of head region; rBG, Pearson correlation of head region; nTP, intercept of tail region; mTP, slope of tail region; rBG, Pearson correlation of tail region) have been used for the analysis. After two tree levels (nBG, nTP) already the decision tree is created. All positive amplification curves (N = 40) are correctly classified. Two observations are classified as false-negative in the negative amplification curves. \\textbf{Lower row} `Performance`)  The ``FFTrees()`` function determines several performance statistics. For the training data, there is a classification table on the left side showing the relationship between tree `decision` and the `truth`. The correct rejection (`Cor Rej`) and `Hit` are the right decisions. `Miss` and false alarm (`False Al`) are wrong decisions. The centre shows the cumulative tree performance in terms of mean of used cues (`mcu`), Percent of ignored cues (`pci`), sensitivity (`sens`), specificity (`spec`), accuracy (`acc`) and weighted Accuracy (`wacc`). The receiver operating characteristic (ROC) curve on the right-hand side compares the performance of all trees in the FFTrees object. The system also displays the performance of the fast frugal trees (`\\#`, green), CART (`C`, red), logistical regression (`L`, blue), random forest (`R`, violet) and the support vector machine (`S`, yellow)."

plot_FFTrees_short <- "Visualization of decisions in  Fast and Frugal Trees after data analysis of amplification curves via the ``mblrr()`` function"

plot_peaks_ratio <- "Working principle of the `peaks\\_ratio` feature. The computation is based on a sequential linking of functions. The ``diffQ()`` function (\\texttt{MBmca}) determines numerically the first derivative of an amplification curve. This derivative is passed to the ``mcaPeaks()`` function (\\texttt{MBmca}). In the output all minima and all maxima are contained. The ranges are calculated from the minima and maxima. The Lagged Difference is determined from the ranges of the minima and maxima. Finally, the ratio of the differences (maximum/minimum) is calculated."

plot_peaks_ratio_short <- "Working principle of the `peaks\\_ratio` feature"

plot_cp_area <- "Analysis of area and changepoint features. Amplification curves from the datasets `stepone\\_std`, `RAS002`, `RAS003`, `lc96\\_bACTXY`, `C126EG595` and `dil4reps94` were analyzed with the ``encu()`` function. These datasets contain positive and negative amplification curves. Furthermore, the meta dataset contains amplification curves that exhibit a hook effect or non-sigmoid shapes, for instance. All amplification curves are manually classified. Altogether 626 positive and 317 negative amplification curves were included in the analysis. A) `polyarea`, is the area under the amplification curve determined by the Gauss polygon area formula. B) `peaks\\_ratio`, is the ratio of the local minima and the local maxima. C) `changepoint\\_e.agglo`, makes use of energy agglomerative clustering. Positive amplification curves have fewer change points than negative amplification curves. These two change point analyses generally separate positive and negative amplification curves. D) `changepoint\\_bcp`, analyses change points by a Bayesian approach. Positive amplification curves appear to contain more change points than negative amplification curves. Nevertheless, there is an overlap between the positive and negative amplification curves in both methods. This can lead to false-positive or false-negative classifications. E) `amptester\\_polygon`, is the cycle normalized order of a polygon.  F) `amptester\\_slope.ratio`, is the slope (linear model) of the raw fluorescence values at the approximate first derivate maximum, second derivative minimum and second derivative maximum."

plot_cp_area_short <- "Analysis of area and changepoint features"

plot_cpa <- "Application of Bayesian change point analysis and energy agglomerative change point analysis methods to the `RAS002` dataset. An analysis of a negative and a positive amplification curve from the `RAS002` dataset was performed using the ``pcrfit\\_single()`` function. In this process, the amplification curves were analysed for change points using Bayesian change point analysis and energy agglomerative clustering. A) The negative amplification curve has a base signal of circa 2450 RFU and only a small signal increase to 2650 RFU. There is a clear indication of the signal variation (noise). B) The first negative derivative amplifies the noise so that some peaks are visible. C) The change point analysis shows changes in energy agglomerative clustering at several positions (green vertical line). The Bayesian change point analysis rarely exceeds a probability of 0.6 (grey vert line). D) The positive amplification curve has a lower base signal ($\\sim 2450$ RFU) and increases up to the 40th cycle ($\\sim 3400$ RFU). A sigmoid shape of the curve is visible. E) The first negative derivation of the positive amplification curve shows a distinctive peak with a minimum at cycle 25. F) The change point analysis in energy agglomerative clustering shows changes (green vertical line) only at two positions. The Bayesian change point analysis shows a probability higher than 0.6 (grey horizontal line) at several positions."

plot_cpa_short <- "Application of Bayesian change point analysis and energy agglomerative change point analysis methods to the `RAS002` dataset"

plot_random_forest <- "Random Forest."

plot_random_forest_short <- "Random Forest"


#----------Tables-------------------------------------
```


\begin{figure}[ht]
\centering
    \scalebox{0.6}{
    \includegraphics[clip=true,trim=1cm 1cm 1cm 1cm]{Logo.pdf}
}
\end{figure}


## Development, Implementation and Installation \label{section_Development_Implementation_and_Installation}

\texttt{PCRedux} is an open source software package ([MIT license](https://opensource.org/licenses/MIT)\footnote{\url{https://opensource.org/licenses/MIT}}) for the statistical computing language \texttt{R}. Data with sigmoid curves are common in bioanalytical methods. A widely used bioanalytical method for the detecition and quantification of DNA is the quantitative real-time PCR (qPCR). qPCRs are applied in human diagnostics, life sciences and forensics [@martins_dna_2015, @sauer_differentiation_2016]. qPCR amplification curves are an example for sigmoid shaped curves \texttt{PCRedux} contains function for the calculation of features from amplification curves and classified datasets for machine learning applications. 

All technical and experimental aspects should be performed under principles that follow good practices of reproducible research. Numerous authors addressed the matter for experimental design and data report. Examples are the *Minimum Information for Publication of Quantitative PCR Experiments* guidelines (MIQE) and the *Real-time PCR Data Markup Language* (RDML). MIQE is a recommended standard of the minimum information for publication of quantitative real-time PCR experiments guidelines and RDML is a data exchange format [@bustin_continuing_2017, @roediger2015r, @roediger_enabling_2017].

The development of scientific software is a complex process. In particular, if a developer team works in different time zones with no face-to-face meetings. End users need releases with stable software that delivers reproducible results. Developers need well documented software to modify the software to their needs. 

Under the umbrella \textit{Agile Software Development} and \textit{Extreme Programming}, several principles were proposed to deliver high quality software, which meet the needs of end users and developers. This includes version control, collaborative editing, unit testing and continuous integration [@lanubile_collaboration_2010, @myers_art_2004, @roediger2015r]. The following paragraphs describe methods applied for the \texttt{PCRedux} package.

### Version Control and Continuous Integration \label{section_Version_Control_and_Continuous_Integration}

The development of the \texttt{PCRedux} package started 2017 with the submission of a functional, yet immature source code, to GitHub (GitHub, Inc.). GitHub is a web-based version control repository hosting service. Both distributed version control and source code management are based on Git. [@lanubile_collaboration_2010]. Additional functionality of GitHub includes the administration of access management, bug tracking, moderation of feature requests, task management, some metrics for the software development, and wikis. The source code of \texttt{PCRedux} is available at:

> \url{https://github.com/devSJR/PCRedux/}

In continuous integration development team members can commit and integrate their contributions several times a day. Team members may include coders, artists and translators. An automated build and test system verifies each integration and gives the development team members a timely feedback about the effect of their commit. In contrast to deferred integration leads this to a reduced number of integration problems and less workload because most errors are solved shortly after they were integrated [@myers_art_2004].

TravisCI was chosen as continues integration service for \texttt{PCRedux}. The TravisCI server communicates with the GitHub version control system and manages the  \texttt{PCRedux} package building process. Continuous interaction is available for the \texttt{R} releases \textit{oldrel}, \textit{release} and \textit{devel}. The history of the build tests are available at:

> \url{https://travis-ci.org/devSJR/PCRedux}

### Naming Convention and Literate Programming \label{section_Naming_Convention_and_Literate_Programming}

\texttt{PCRedux} is an \text{R} ($\geq$ v. 3.3.3) package, written as \textit{S3} object system. \textit{S3} has characteristics of object orientated programming but eases the development due to the use of the naming conventions [@noauthor_compstat_2008]. In most places functions and parameter names are written as underscore separated (underscore$\_$sep), which is a widely used style in \texttt{R} packages [@Baaaath_2012]. This convention had to be violated in coding sections where functionality from other packages was used.

Literate programming, as proposed by @knuth_literate_1984, is a concept where the logic of the source code and documentation is integrated in a single file. Markup conventions (e.&nbsp;g., '#') tell in literate programming how to typeset the documentation. This produces outputs in a typesetting language such as the lightweight markup language **Markdown**, or the document preparation system \LaTeX. 

The \texttt{roxygen2}, \texttt{rmarkdown} and \texttt{knitr} packages were used to write the documentation in-line with code for the \texttt{PCRedux} package.

### Installation of the \texttt{PCRedux} Package \label{section_Installation_of_the_PCRedux_Package}

The development version of \texttt{PCRedux} can be installed using the \texttt{devtools} package.

```{r, eval=FALSE, echo=TRUE}
# Install devtools, if not already installed.
install.packages("devtools")

# Install PCRedux
devtools::install_github("devSJR/PCRedux")
```

\texttt{PCRedux} is available as stable version from the \textbf{C}omprehensive \texttt{R} \textbf{A}rchive \textbf{N}etwork (CRAN) at \url{https://CRAN.R-project.org/package=PCRedux}. Package published at CRAN undergo intensive checking procedures. In addition, CRAN tests whether the package can be built for common operating systems and whether all version dependencies are solved. To install \texttt{PCRedux} first install \texttt{R} ($\geq$ v. 3.3.3). Then start \texttt{R} and type in the prompt:

```{r, eval=FALSE, echo=TRUE}
# Select your local mirror
install.packages("PCRedux")
```

The \texttt{PCRedux} package should just install. If this failed make sure that write access is permitted to the destination directory and that all package dependencies are met.

```{r, eval=FALSE, echo=TRUE}
# The following command points to the help for download and install of packages
# from CRAN-like repositories or from local files.
?install.packages()
```

If this fails try to follow the instructions given  by @de_vries_r_2012.

> R CMD check 

Results from CRAN check can be found at 

> \url{http://cran.us.r-project.org/web/checks/check_results_PCRedux.html}.


### Unit Testing of the \texttt{PCRedux} Package \label{section_Unit_Testing_of_the_PCRedux_Package}

Modules testing, better known as unit testing, is an approach to simplify the refactoring of source code during software development. Unit Testing is not a guarantee for error-free software. The goal is to minimize errors and regressions. It is also intended to ensure that the numerical results from the calculations are reproducible and of high quality. An unintended behavior of the software should be detected at the latest during the package building process [@myers_art_2004].

*Checkpoints* are used to check whether the software performs calculations and data transformations correctly for all builds. For this, numerous (logical) queries have to be defined by the developer in advance. They are refereed to \textit{expectations}. It should be ensured that as many errors as possible are covered. A logical query can be, for example, whether the calculation has a numeric or Boolean value as output. If the data type is incorrect during output, this is a sufficient termination criterion. Or it can be checked whether the length of the result vector is correct after the calculation. There are different approaches for unit tests in \texttt{R}. This also includes testing of units from the packages \texttt{RUnit}, \texttt{covr}, \texttt{svUnit} and \texttt{testthat}. (@wickham_testthat_2011).

The package \texttt{testthat} was used in \texttt{PCRedux} because it could be well implemented and its maintenance is relatively simple. The logic is that an \textit{expectation} defines how the result, class or error in the corresponding unit (e.&nbsp;g., function) should behave. Unit tests can be found in the `/test/testthat` subdirectory of the \texttt{PCRedux} package. They are run automatically during the creation of the package. The following example for the ``qPCR2fdata()`` function (for details see \autoref{section_qPCR2fdata}) uses the ``test_that()`` function from the \texttt{testthat} package with the \textit{expectations} that:

* an object of the class *fdata* is created (see @Febrero_Bande_2012 for details of the class *fdata*),
* the parameter `rangeval` has a length of two,
* is the second value of parameter `rangeval` 49 (last cycle number) and 
*whether the object structure of the function ``qPCR2fdata()`` does not change if the parameter `preprocess=TRUE` is set.

```{r, eval=FALSE, echo=TRUE}
# Expectations used for the unit testing of the qPCR2fdata() function.
library(PCRedux)

context("qPCR2fdata")

test_that("qPCR2fdata gives the correct dimensions and properties", {
    library(qpcR)
    res_fdata <- qPCR2fdata(testdat)
    res_fdata_preprocess <- qPCR2fdata(testdat, preprocess = TRUE)
    
    expect_that(res_fdata, is_a("fdata"))
    expect_that(length(res_fdata$rangeval) == 2 &&
    res_fdata$rangeval[2] == 49, is_true())
    
    expect_that(res_fdata_preprocess, is_a("fdata"))
    expect_that(length(res_fdata_preprocess$rangeval) == 2 &&
    res_fdata_preprocess$rangeval[2] == 49, is_true())
})
```

Further unit tests were implemented for all functions of the \texttt{PCRedux} package. The coverage by \texttt{PCRedux} package can be calculated by the ``package_coverage()`` function from the \texttt{covr} package [@covr] or visual analyzed via web-interface at:

> \url{https://codecov.io/gh/devSJR/PCRedux/list/master/}.


\newpage

## Analysis of Sigmoid Shaped Curves for Data Mining and Machine Learning Applications\label{section_reasoning_and_analysis}


The following sections describe the \texttt{PCRedux} regarding the analysis, numerical description and feature calculation from a sigmoid curve. A feature herein refers to a quantifiable *informative* property of a sigmoid curve. The features, sometimes referred to as descriptors, (\autoref{section_Functions_of_PCRedux}) can be used for applications such as data mining, machine learning and automatic classification (e.&nbsp;g., negative or positiv amplification). The determination of quantification points such as the Cq value is a typical task during the analysis of qPCR experiments. This is briefly described in dedicated sections (\autoref{section_software}ff.).

Characteristics of amplification curves that can be used for the statistical and analytical description are discussed (\autoref{section_analysis_functions}) more in detail. The examples described focus on the concepts for **binary (dichotomous) classification** [@kruppa_probability_2014] as negative or positive. The mere binary classification into positive or negative is not necessarily the aim of the \texttt{PCRedux} package. Instead, it is aimed to provide a tool set for automatic **multicategory (polychotomus) classification** of amplification curves by any class conceivable. Such classification could be used for the quality of an amplification curve as negative, ambiguous and positive (\autoref{htPCR_nap}) \footnote{A definition for binary (dichotomous) classification and multicategory (polychotomus) classification is presented in \cite{kruppa_probability_2014}.}.

```{r htPCR_nap, echo=FALSE, fig.cap=htPCR_nap, fig.scap=htPCR_nap_short, fig.height=4}

library(qpcR)
matplot(
    htPCR[, 1], htPCR[, c(552, 512, 616)], xlab = "Cycles", ylab = "RFU",
        main = "", type = "l", lty = 1, lwd = 2
)
legend("topleft", c(
    paste("negative ", colnames(htPCR)[552]),
                    paste("ambiguos ", colnames(htPCR)[512]),
                    paste("positive ", colnames(htPCR)[616])
), col = 1:3, pch = 19, bty = "n")
```

### Concepts of Machine Learning\label{section_technologies_amplification_curves_ML}

Data mining and machine learning can be used for descriptive and predictive tasks during the analysis of complex datasets. Data mining uses specific methods from statistical interference, software engineering and domain knowledge to get a better understanding of the data and to extract *hidden knowledge* from the pre-processed data [@kruppa_probability_2014, @herrera_multiple_2016]. All this implies that a human being interacts with the data at the different stages of the whole process as part of the workflow in data mining. Elements of the data mining process are the pre-processing of the data, the description of the data, the exploration of the data and the search for connections and causes.

The availability of classified amplification curve datasets and technologies for the classification of amplification curves is of high importance to train and validate models. This is dealt with in \autoref{section_data_sets} and \autoref{chapter_humanrater}, respectively.

In contrast, machine learning uses instructions and data in software modules to create models that can be used to make predictions on novel data. In machine learning, the human being is much less necessary in the entire process. Processes (algorithms) are used to create models with tunable parameters. These models automatically adapt their performance to the information (features) from the data. Well-known examples of machine learning technologies are Decision Trees (DT), Boosting, Random Forests (RF), Support Vector Machines (SVM), generalized linear models (GLM), logistic regression (LR) and deep neural networks (DNN) [@lee_statistical_2010]. The three following concepts of machine learning are frequently described in the literature:

*Supervised learning*: These algorithms (e.&nbsp;g., SVM, DT, RF) learn from a training dataset of labeled and annotated data (e.&nbsp;g., "positive" and "negative"). Classified training data can be created by one or more individuals. It is used for building a generalized model of all data. These algorithms use error or reward signals to evaluate the quality of a solution found [@bischl_mlr:_2010, @greene_big_2014, @igual_introduction_2017]. Binomial logistic regression\footnote{Logistical regression can also be used to predict a dependent variable that can assume more than two states. In this case, it is called a multinomial logistic regression. An example would be the classification $y$ of amplification curves as \textit{slightly noisy}, \textit{medium noisy} or \textit{heavily noisy}.} is used to gain knowledge about a binary relationship, by fitting a regression model $y = f(x)$. $y$ is a categorical variable with two states (negative $\rightarrow 0$, positive $\rightarrow 1$). Typically, this model is used for predicting $y$ with a mixture of $n$ continuous and categorical predictors (features) $x_{i1}, \ldots, x_{k1}, (i = 1, \ldots, n)$. 

The logit model is a robust and versatile classification method to explain a dependent binary variable. Their codomain of real numbers is limited to [0,1]. Probabilities can therefore be utilized. The logistical distribution function $F(\eta)$, also known as the response function, is strictly monotone increasing and limited to this range.

$\eta_{i}$ establishes the link between the probability of the occurrence and the independent variables. For this reason, $\eta_{i}$ is referred to as a link function. The distribution function of the normal distribution is an alternative to the logistical distribution function. By using the normal distribution, the Probit model is obtained. However, since this is more difficult to interpret, it is less widely used in practice. Since probabilities are used, it is possible to make a prediction about the probability of occurrence of an event. 

When analyzing amplification curves, diagnosis can be made whether a reaction was unsuccessful (0) or successful (1). For the prediction independent metric variables (features) are used. The metric variables have interpretable distances with a defined order. Their codomain is [-$\infty$,$\infty$]. The logistic distribution function on the independent variables determines the probability for $Y_{i} = 0$ or $Y_{i} = 1$. A logistic regression model can be formulated as follows:

$F(\eta)=\frac{1}{1+exp(-\eta)}$

The logistic regression analysis is based on the maximum-likelihood estimation (MLE). In contrast to linear regression, the probability for $Y=1$ is not modeled from explanatory variables. Rather, the logarithmic chance (logit) is used for the occurrence of $Y=1$. The term *chance* refers to the ratio of the probability of occurrence of an event (e.&nbsp;g., amplification curve is positive) and the counter-probability (e.&nbsp;g., amplification curve is negative) of an event.

*Unsupervised learning*: Algorithms, such as k-means clustering, kernel density estimation, LDA or PCA learn from training datasets of unlabeled or non-annotated data to find hidden structures according to geometric or statistical criteria [@bischl_mlr:_2010, @greene_big_2014, @igual_introduction_2017].

*Reinforcement Learning*: The algorithms learn by reinforcement from `criticism`. The criticisms inform the algorithm about the quality of the solution found but nothing about how to improve. These algorithms iteratively search the improved solution in the entire solution space [@bischl_mlr:_2010, @igual_introduction_2017].

\newpage


### Why is there is need for this software?\label{why_PCRedux}

The binary classification of an amplification curve, for instance as negative or positive, is feasible using bioanalytical methods such as melting curve analysis [@roediger_RJ_2013] or electrophoretic separation [@westermeier2004]. However, this is not always possible or desirable. 

- Melting curve analysis is used in some qPCRs as a post-processing step to identify samples which contain the specific target sequence (*positive*) based on a specific melting temperature. However, some detection probe systems like hydrolysis probes do not permit such classification. Moreover, nucleic acids with similar biochemical properties but different sequences may have the same melting temperature. 
- An electrophoretic separation (classification of target DNA sequences by size and quantity) often requires too much effort for experiments with high sample throughput.
- There are mathematical qPCR analysis algorithms such as \texttt{linreg} [@ruijter_amplification_2009] that require information on whether an amplification curve is negative or positive for subsequent calculation. 
- Raw data of amplification curves can be fitted with sigmoid functions. Sigmoid functions are non-linear, real-valued, have an S-shaped curvature (\autoref{figure_sigmoid_curve_models}) and are differentiable (e.&nbsp;g., first derivative maximum, with one local minimum and one local maximum). With the model obtained, predictions can be made. For example, the position of the second derivative maximum can be calculated from this (\autoref{section_DataAnalysis}). In the context of amplification curves, the second derivative maximum is commonly used to describe the relationship between the cycle number and the PCR product formation (\autoref{section_DataAnalysis}). All softwares assume that the amplification resemble a sigmoid curve shape (ideal positive amplification reaction), or a flat low line (ideal negative amplification reaction). For example, @Ritz2008 published the \texttt{qpcR} \texttt{R} package that contains functions to fit several multi-parameter models. This includes the five-parameter Richardson function [@richards_flexible_1959] (\autoref{l5}). The \texttt{qpcR} package [@Ritz2008] contains an amplification curve test via the ``modlist()`` function. The parameter `check="uni2"` offers an analytical approach, as part of a method for the kinetic outlier detection. It tries to checks for a sigmoid structure of the amplification curve. Then ``modlist()`` tests for the location of the first derivative maximum and the second derivative maximum. However, multi-parameter functions fit "successful" in most cases including noise and give false positive results. This will be shown in later sections. This is exemplary shown in later sections in combination with the ``amptester()`` function, which is part of the \texttt{chipPCR} package [@roediger2015chippcr]. This function uses static thresholds and frequentist inference to identify amplification curves that exceed the threshold ($\mapsto$ classified as positive). However, the analysis can also lead to false-positive classifications as exemplified in the example below and in \autoref{curve_fit_fail}. Therefore, additional classification concepts would be beneficial.


```{r curve_fit_fail, echo=TRUE, fig.cap=curve_fit_fail, fig.scap=curve_fit_fail_short, fig.width=6}
# Load the qpcR package for the model fit.
library(qpcR)
library(chipPCR)

# Select one positive and one negative amplification curve from the PCRedux 
# package.

amp_data <- PCRedux::RAS002[, c("cyc", "A01_gDNA.._unkn_B.Globin", 
                                "B07_gDNA.._unkn_HPRT1")]

colnames(amp_data) <- c("cyc", "positive", "negative")

# Arrange graphs in an matrix and set the plot parameters. An plot the positive
# and negative amplification curve.
hight <- c(3100, 4100)

plot(NA, NA, xlim = range(amp_data[, "cyc"]), 
        ylim = range(amp_data[, c("positive", "negative")]),
        xlab = "Cycles", ylab = "RFU", main = "")

# Apply the amptester function from the chipPCR package to the amplification 
# curve data and write the results to the main of the plots.

for (i in 2:3) {
    res.ampt <-  suppressMessages(amptester(amp_data[, i]))
    
    # Make a logical connection by two tests (shap.noisy, lrt.test and
    # tht.dec) of amptester to decide if an amplification reaction is
    # positive or negative.
    decision <- ifelse(!res.ampt@decisions[1] &&
    res.ampt@decisions[2] &&
    res.ampt@decisions[4],
    "positive", "negative"
    )
    # The amplification curves were fitted (l7 model) with pcrfit() function. 
    # The Cq was determined with the efficiency() function.
    
    fit <- pcrfit(data = amp_data, cyc = 1, fluo = i, model = l7)
    res <- efficiency(fit, plot = FALSE)
    lines(predict(fit), pch = 19, lty = 1, xlab = "Cycles", ylab = "RFU", 
          main = "", col = i - 1)
    abline(h = res[["fluo"]], col = "grey")
    points(res[["cpD2"]], res[["fluo"]], pch = 19)

    legend(1, hight[i-1], paste0(colnames(amp_data)[i], 
                                 "  curve ->  Decision: ", 
                                 decision, "    Cq: ", res[["cpD2"]]), 
           bty = "n", cex = 1, col = "red"
          )
}
```

- The analysis and classification of sigmoid data (e.&nbsp;g., quantitative PCR) it is a manageable task if the data volume is low, or dedicated analysis software is available. An example for a low number of amplification curves is shown in \autoref{figure_sigmoid_curve}A. All `r ncol(chipPCR::C127EGHP)-1` curves exhibit a sigmoid curve shape. It is trivial to classify them as positive by hand. In contrast, the vast number of amplification curves in \autoref{figure_sigmoid_curve}B is barely manageable with a reasonable effort by simple visual inspection. These data originate from a high-throughput experiment that encompasses in total `r suppressMessages(ncol(qpcR::htPCR))-1` amplification curves of which only 200 are shown. A manual analysis of the data is time-consuming and prone to errors. Even for an experienced user it is difficult to classify the amplification curves unambiguously and reproducible as will be later shown in \autoref{section_data_sets}.
- qPCRs are performed in thermo-cyclers, which are equipped with a real-time monitoring technology. There are numerous commercial manufactures, which produce thermo-cyclers (\autoref{table-datasets}). An example for a thermo-cycler that originated in scientific project is the VideoScan technology [@roediger_highly_2013]. Most of the thermo-cyclers have a thermal block with wells at certain positions. Reaction vessels containing the PCR mix are inserted into the wells. There are also thermo-cyclers that use capillary tubes that are heated and cooled by air (e.&nbsp;g., Roche Light Cycler 1.0). The thermo-cycler raises and lowers the temperature in the reaction vessels in discrete, pre-programmed steps so that the PCR can take place. Instruments with a real-time monitoring function have sensors to measure changes of the fluorescence intensity in the reaction vessel. All thermo-cycler systems use software to processes the amplification curves. Plots of the fluorescence observations versus cycle number obtained from two different qPCR systems is shown in \autoref{figure_sigmoid_curve}A and B. The thermo-cyclers produce different amplification curve shapes even with the same sample material and PCR mastermix because of their technical design, sensors, and software. These factors need to be taken into account during the development of analysis algorithms.


```{r figure_sigmoid_curve_models, echo=FALSE, fig.scap="Sigmoid models of amplification curves", fig.height=8, out.extra='', fig.cap="A) Model function of a one-parameter sigmoid function. B) Model function of a sigmoid function with an intercept $n$ = 0.2 RFU (shift in base-line). C) Model function of a sigmoid function with an intercept ($n$ \\textasciitilde 0.2 RFU) and a square portion $m * x^{2}, m = -0.0005, n = 0.2 RFU$ (hook-effect-like). D) Model function of a sigmoid function with an intercept ($n$) and a square portion of $m * x^{2}$ and additional noise $\\epsilon$ (normal distributed, $\\mu = 0.01, \\sigma = 0.05$)."}

x_val <- seq(-10, 10, 0.5)
y_val <- 1 / (1 + exp(-x_val))
y_val_slope <- 1 / (1 + exp(-x_val)) + 0.2
y_val_slope_quadratic <- 1 / (1 + exp(-x_val)) + -0.0005 * x_val ^ 2 + 0.2
y_val_slope_quadratic_noise <- 1 / (1 + exp(-x_val)) + -0.0005 * x_val ^ 2 + 0.2 + rnorm(length(x_val), mean = 0.01, sd = 0.05)

y_lim <- c(-0.05, max(c(
    y_val, y_val_slope, y_val_slope_quadratic,
    y_val_slope_quadratic_noise
)) * 1.2)

par(mfrow=c(2,2))
plot(x_val, y_val, type = "l", xlab = "x", ylab = "f(x)", ylim = y_lim)
abline(h = 0, col = "grey")
legend("topleft", expression(y == frac(1, (1 + e ^ {
    -x
}))), bty = "n", cex = 0.9)
mtext("A", cex = 1.2, side = 3, adj = 0, font = 2)

plot(x_val, y_val_slope, type = "l", xlab = "x", ylab = "f(x)", ylim = y_lim)
abline(h = 0, col = "grey")
legend("topleft", expression(y == frac(1, (1 + e ^ {
    -x
})) + n), bty = "n", cex = 0.9)
mtext("B", cex = 1.2, side = 3, adj = 0, font = 2)

plot(
    x_val, y_val_slope_quadratic, type = "l", xlab = "x", ylab = "f(x)",
     ylim = y_lim
)
abline(h = 0, col = "grey")
legend("topleft", expression(y == frac(1, (1 + e ^ {
    -x
})) + m * x ^ 2 + n), bty = "n", cex = 0.9)
mtext("C", cex = 1.2, side = 3, adj = 0, font = 2)

plot(
    x_val, y_val_slope_quadratic_noise, type = "l", xlab = "x", ylab = "f(x)",
     ylim = y_lim
)
abline(h = 0, col = "grey")
legend("topleft", expression(y == frac(1, (1 + e ^ {
    -x
})) + m * x ^ 2 + n + epsilon, epsilon %~% N(0, sigma)), bty = "n", cex = 0.9)
mtext("D", cex = 1.2, side = 3, adj = 0, font = 2)
```


```{r figure_sigmoid_curve, echo=FALSE, fig.scap="Shape of amplification curves", fig.height=8, out.extra='', fig.cap="Amplification curve data from an iQ5 (Bio-Rad) thermo-cycler and a high throughput experiment in the Biomark HD (Fluidigm). A) The `C127EGHP` dataset with 64 amplification curves was produced in conventional thermo-cycler with a 8 x 12 PCR grid. B) The `htPCR` dataset, which contains 8858 amplification curves, was produced in a 95 x 96 PCR grid. Only 200 amplification curves are shown. In contrast to `A)` have all amplification curves in `B)` an off-set (intercept) between 0.09 and 0.40 RFU."}
library(chipPCR)
library(qpcR)

par(mfrow = c(2, 1), las = 0, bty = "o", oma = c(0, 0, 0, 0))

colors <- rainbow(ncol(C127EGHP) - 2, alpha = 0.5)
matplot(
    C127EGHP[, 2], C127EGHP[, c(-1, -2)], xlab = "Cycles", ylab = "RFU",
        main = "", type = "l", lty = 1, lwd = 2, col = colors
)
abline(h = 0, col = "grey")
mtext("A    iQ5 C127EGHP dataset", cex = 1.2, side = 3, adj = 0, font = 2)

colors <- rainbow(200, alpha = 0.5)
matplot(
    htPCR[, 1], htPCR[, c(2L:201)], xlab = "Cycles", ylab = "RFU", ylim = c(0, 1.75),
        main = "", type = "l", lty = 1, lwd = 2, col = colors
)
abline(h = 0, col = "grey")
mtext("B    Biomark HD htPCR dataset", cex = 1.2, side = 3, adj = 0, font = 2)
```

### Software for the Analysis of Amplification Curve Data \label{section_software}

There are several open source and closed source software tools for the analysis of qPCR data [@pabinger_2014]. The software packages deal for example with

* missing values and non-detects [@mccall_non-detects_2014], 
* noise and artifact removal [@roediger2015chippcr, @roediger2015r, @spiess_impact_2015, @spiess_system-specific_2016],
* inter run calibration [@ruijter_removal_2015], 
* normalization [@roediger2015chippcr, @ruijter_evaluation_2013, @feuer_lemming:_2015, @matz_no_2013], 
* quantification cycle estimation [@Ritz2008, @ruijter_evaluation_2013], 
* amplification efficiency estimation [@Ritz2008, @ruijter_evaluation_2013], 
* data exchange [@lefever_rdml_2009, @perkins_readqpcr_2012, @roediger_enabling_2017],
* relative gene expression analysis [@dvinge_htqpcr:_2009, @pabinger_qpcr:_2009, @neve_unifiedwmwqpcr:_2014] and 
* data analysis pipelines [@pabinger_qpcr:_2009, @ronde_practical_2017, @mallona_pcrefficiency:_2011, @mallona_chainy:_nodate].

However, a bottleneck of qPCR data analysis is the lack of features and software to build classifiers for amplification curves. A classifier herein refers to a vector of features that can be used to distinguish the amplification curves by their shape only. A feature is an entity that characterizes an object. A few potential features for amplification curves are described in the literature. These include:

* the starting point (*takeoff*) of the amplification curve,
* the Cq value and amplification efficiency, and 
* the signal level (e.g., slope and intercept of the ground phase).

These alone are presumably not enough to describe amplification curves sufficiently. The number of features should be large enough to describe the object accurately and small enough not to interfere with the learning process with redundant or information. There are no references of algorithms in the scientific literature for the calculation of additional features from amplification curves. This makes studies on machine learning and modeling difficult.

### Principles of Amplification Curve Data Analysis and Feature Calculation\label{section_DataAnalysis}

The shape of a positive amplification curve has in most cases a sigmoid shape. Many factors, such as the sample quality, qPCR chemistry, and technical problems (e.&nbsp;g., sensor errors) contribute to various curve shapes [@ruijter_2014]. The curvature of the amplification curve can be used as a quality measure. For example, fragmentation, inhibitors, and sample material handling errors during the extraction can be identified. The kinetic of fluoresce emission is proportional to the quantity of the synthesized DNA. Typical amplification curves have three phases. 

1. **Ground phase**: This phase occurs during the first cycles of the PCR. The fluorescence emission is in most cases flat. During the ground phase, only a weak and flat fluorescence signal is generated. Noise but no product formation is detected by the sensor system. The PCR product signal is an insignificantly small component of the total signal. This is often referred to as base-line or background signal. Apparently, there is only a phase shift or no signal at all. This is primarily due to the limited sensitivity of the instrument. Even in a perfect PCR reaction (double amplification per cycle), qPCR instruments cannot detect the fluorescence signal from the amplification. Fragmentation, inhibitors and sample handling errors would result in a prolonged ground phase. Nevertheless, this may indicate some typical properties of the qPCR system or probe system. In many instruments, this phase is used to determine the base-line level for the calculation of the Cycle threshold (Ct). The Ct value is considered statistically relevant increase outside the noise range. A signal that is far enough above this threshold is considered coming from the amplicon. In some qPCR systems a flat amplification signal is expected in this phase. Slight deviations from this trend are presumably due to changes (e.&nbsp;g., disintegration of probes) in the fluorophores. Background correction algorithms are often used here to ensure that flat amplification curves without slope are generated. However, this can result in errors and inevitably leads to a loss of information via the waveform of the raw data [@nolan_2006]. The slope, level and variance of this phase can serve as features.
2. **Exponential phase**: This phase follows the ground phase and is also called *log phase*. This phase is characterized by a strong increase of the emitted fluorescence. In this phase, the DNA amount doubles in each cycle under ideal conditions. The amount of the synthesized fluorescent labeled PCR product is high enough to be detected by the sensor system. This phase is used for the calculation of the quantification point (Cq) and for the calculation of the curve specific amplification efficiency. The most important measurement from qPCRs is the cycle of quantification (Cq), which signifies at which PCR cycle the fluorescence exceeds a ``threshold value``. There is an ongoing debate as to what a significant and robust threshold value is. An overview and performance comparison of Cq methods is given in @ruijter_evaluation_2013. There are several mathematical methods to calculate the Cq. 
    - The 'classical' threshold value (cycle threshold, Ct) is the intersect between a manually defined straight horizontal line with the quasi-linear phase in the exponential amplification phase (\autoref{figure_quntifcation_points}A & B). This simple to implement method requires that amplification curves are properly base-lined prior to the analysis. The Ct method makes the assumption that the amplification efficiency (~ slope in the log-linear phase) is equal across all amplification curves compared [@ruijter_evaluation_2013]. Evidently, this is not always case as exemplified in \autoref{amplification_curve_ROI}C. The Ct method is widely used presumably due to the familiarity of users with this approach (e.g., chemical analysis procedures).  However, this method is statistically unreliable [@ruijter_evaluation_2013, @spiess_impact_2015, @spiess_system-specific_2016]. Moreover, the Ct method gives no stable in predictions if different users are given the same dataset to be analyzed. *Therefore, this method is not used within the \texttt{PCRedux} package*.
    - Another Cq method uses the maximum of second derivative (SDM) [@roediger2015r] (\autoref{figure_quntifcation_points}C). In all cases the Cq value can be used to calculate the concentration of target sequence in a sample (low Cq \textrightarrow high target concentration). In contrast, negative or ambiguous amplification curves loosely resemble noise. This noise may appear linear or exhibit an curvature similar to a specific amplification curve (\autoref{htPCR_nap}). This however, may result in faulty interpretation of the amplification curves. Fragmentation, inhibitors and sample handling errors would decrease the slop of the amplification curve [@spiess_highly_2008, @Ritz2008]. The slope and variation can be considered as features. Since the Cq depends on the initial template amount, and the amplification efficiency there is no immediate use of the Cq as feature. 
    3. **Plateau phase**: This phase follows the exponential phase. The cause for this lies in the exploitation of the limited resources (incl. primers, nucleotides, enzyme activity) in the reaction vessel. This limits the amplification reaction, so that the theoretical maximum amplification efficiency (doubling per cycle) no longer prevails. This turning point and the progressive limitation of resources finally leads to a plateau. In the plateau phase, there is sometime a signal decrease called *hook effect* (\autoref{why_PCRedux} and [@barratt_improving_2002, @isaac_essentials_2009]). The slope (*hook effect*), level and variation can be considered as features. 

If the amplification curve has only a slight positive slope and no perceptible exponential phase, it can be assumed that the amplification reaction did not occur (\autoref{amplification_curve_ROI}B). Causes may include poor specificity of the PCR primers (non-specific PCR products), degraded sample material, degraded probes or detector failures. If a lot of input DNA is present in a sample, the amplification curve starts to increase in early PCR cycles (1 - 12 cycles). Some PCR devices have software that corrects this data without checking it. This results in an amplification curve with a negative trend.

The discussed phases are considered as regions of interest (ROI). As an example, the \textit{ground phase} is in the head area, while the \textit{plateau phase} is in the tail area. The \textit{exponential phase} is located between these two ROIs.


```{r amplification_curve_ROI, echo=FALSE, fig.scap=amplification_curve_ROI_short, fig.cap=amplification_curve_ROI, fig.height=8, fig.width=8}

library(qpcR)
library(PCRedux)

colors <- rainbow(10, alpha = 0.15)

x_range <- 1L:35
d <- testdat[x_range, ]
amp_data <- data.frame(
  d[, 1],
  pos = d[, 3] + 0.9,
  posReverse = (max(d[, 3]) - rev(d[, 3])) + 0.9,
  neg = d[, 4] + 0.9 + 0.0005 * d[, 1] ^ 2
)

# Calculation for the normal data
res_amp_data <- pcrfit(amp_data, 1, 2, l5)
res_takeoff <- takeoff(res_amp_data)

# Calculation of sd_bg

res_sd_bg <- sd(amp_data[1:res_takeoff[[1]], 2])

# Calculation for the reversed data
res_amp_data_reverse <- pcrfit(amp_data, 1, 3, l5)
res_takeoff_reverse <- takeoff(res_amp_data_reverse)
res_takeoff_reverse[[1]] <- nrow(d) - res_takeoff_reverse[[1]]
res_takeoff_reverse[[2]] <- amp_data[res_takeoff_reverse[[1]], 2] - res_takeoff_reverse[[2]] + min(amp_data[, 3])

exponentialRange <- c((res_takeoff[[1]] + 1):(res_takeoff_reverse[[1]] - 1))

backgroundplateu <- function(x) {
  bg <- mean(head(x, res_takeoff[[1]])) + 3 * sd(head(x, res_takeoff[[1]]))
  plat <- mean(tail(x, 10)) - 3 * sd(tail(x, 10))
  list(bg = bg, plateau = plat)
}

res_lm <- lm(amp_data[exponentialRange, 2] ~ amp_data[exponentialRange, 1])

y_lim <- max(amp_data[, 2:4]) * 1.15

res_bgpl <- unlist(backgroundplateu(amp_data[, 2]))

layout(matrix(c(1, 1, 2, 3), 2, 2, byrow = TRUE), respect = TRUE)

plot(amp_data[, 1], amp_data[, 2], ylim = c(-0.1, y_lim), xlab = "Cycles", ylab = "RFU", type = "b", lwd = 2, pch = 19)

text(c(2,30), c(10.5,10.5), c("Head", "Tail"), cex = 1.2, col = "red")

rect(0, 0, res_takeoff[[1]] + 1, res_takeoff[[2]] * 1.25, col = colors[1], border = NA)
text(5, res_bgpl[1] * 1.45, "Ground phase")

rect(res_takeoff_reverse[[1]] - 1, res_takeoff_reverse[[2]] * 0.95, nrow(amp_data), y_lim, col = colors[5], border = NA)
text(32, res_bgpl[2] * 1.1, "Plateau phase")

text(res_takeoff_reverse[[1]], mean(amp_data[, 2]), "Exponential\nregion")


points(
  c(res_takeoff[[1]], res_takeoff_reverse[[1]]),
  c(res_takeoff[[2]], res_takeoff_reverse[[2]]), pch = 12, cex = 2.5
)

text(
    c(res_takeoff[[1]], res_takeoff_reverse[[1]]),
    c(res_takeoff[[2]], res_takeoff_reverse[[2]]) + c(1.05, -1.05), c("top", "tdp")
)


arrows(20, 0, 20, res_bgpl[1], code = 3, length = 0.1)
text(30, res_bgpl[1] / 2, "Background")

arrows(5, res_bgpl[2], 5, max(amp_data[, 2]), code = 3, length = 0.1)
text(15, res_bgpl[2] * 0.95, "Plateau")


abline(res_lm, col = "red")
points(amp_data[exponentialRange, 1], amp_data[exponentialRange, 2], pch = 19, col = "red")

abline(h = res_bgpl, col = c("green", "blue"))
abline(h = 0, col = "grey")

legend(2, 12, paste0(
  "Slope: ", signif(coef(res_lm)[2], 3),
  "\nBackground (mean): ", signif(res_bgpl[1], 3),
  "\nsd_bg: ", signif(res_sd_bg, 3),
  "\nPlateau: ", signif(res_bgpl[2], 3),
  "\ntop: ", signif(res_takeoff[[1]], 3),
  "\ntdp: ", signif(res_takeoff_reverse[[1]], 3)
), bty = "n")

mtext("A    Positive", cex = 1, side = 3, adj = 0, font = 2)


y_lim <- 2
plot(amp_data[, 1], amp_data[, 4], ylim = c(-0.1, y_lim), xlab = "Cycles", ylab = "RFU", type = "b", lwd = 2, pch = 19)
res_bgpl <- unlist(backgroundplateu(amp_data[, 4]))
abline(h = res_bgpl, col = c("green", "blue"))
abline(h = 0, col = "grey")

mtext("B    Negative", cex = 1, side = 3, adj = 0, font = 2)

curve_colors <- c(rainbow(ncol(boggy) - 1, alpha = .5))
matplot(boggy[, 1], boggy[, -1], type = "l", col = curve_colors, xlab = "Cycles", ylab = "RFU", lty = 1)
    rect(22, 2, 40, 2.3, border = "blue")
    text(27.5, 2.1, "Hook effect", col = "blue")
mtext("C    boggy dataset", cex = 1, side = 3, adj = 0, font = 2)
```

The amplification curve shape, the amplification efficiency and the Cq value are important measures to judge the outcome of a qPCR reaction. In all phases of PCR the curves should be smooth. Possible artifacts in the curves may be due to unstable light sources from the instrument or problems during sample preparation, such as the presence of bubbles in the reaction vessel,incorrectly assigned dye detectors, errors during the calibration of dyes for the instrument, errors during the preparation of the PCR master mix, sample degradation, lack of a sample in the PCR, too much sample material in the PCR mix or a low detection probe concentration [@ruijter_amplification_2009, @ruijter_2014, @spiess_impact_2015]. Smoothing and filtering cause alterations to the raw data that affects the Cq value and the amplification efficiency.

Most commercial qPCR systems do not display the raw data of the amplification curves on the screen. Instead, raw data are often processed by the instrument software to remove fluorophore-specific effects and noise in all ROI's. Commonly employed pre-processing step of qPCR is smoothing and filtering to remove noise. Noise in amplification curves can have different causes [@spiess_impact_2015]. 

The ordinate often does not display the measured fluorescence, but rather the change in fluorescence per cycle ($\varDelta RFU = RFU_{cycle  + 1} - RFU_{cycle}$). Some qPCR systems have a periodicity in the amplification curve data. Periodicity exposes the risk of introducing artificially shifts in the Cq values [@spiess_system-specific_2016]. 

In particular the cycle threshold method (Ct method) (\autoref{section_DataAnalysis}) is affected by these factors [@spiess_impact_2015, @spiess_system-specific_2016]. Therefore, it is advisable to clarify beforehand, which processing steps the amplification curves have been subjected to. Failure to do so may result in misinterpretations and incorrect models [@nolan_2006, @roediger2015r, @roediger2015chippcr, @spiess_impact_2015].



```{r figure_quntifcation_points, results='hide', message=FALSE, echo=FALSE, fig.scap=figure_quntifcation_points_short, fig.cap=figure_quntifcation_points, fig.height=8, fig.width=8} 
library(qpcR)
library(chipPCR)
library(magrittr)

res_model <- pcrfit(testdat, cyc = 1, fluo = 2, model = l5)
res_takeoff <- takeoff(res_model, pval = 0.05, nsig = 3)

res_model_predict <- predict(res_model)

r_user <- 2.356

res_th.cyc <- th.cyc(testdat[, 1], testdat[, 2], r = r_user, linear = FALSE)


par(las = 0, oma = c(0, 0, 0, 0))

layout(matrix(c(1, 2, 3, 3), 2, 2, byrow = TRUE), respect = TRUE)


plot(testdat[, 1], testdat[, 2], xlab = "Cycles", ylab = "Raw fluorescence")

abline(h = (mean(testdat[1:10, 2]) + 3 * sd(testdat[1:10, 2])), col = "grey")

abline(h = res_th.cyc[1, 2], col = "black")
text(28, r_user + 0.3, paste0("Threshold: ", r_user))
arrows(res_th.cyc[1, 1], res_th.cyc[1, 2], res_th.cyc[1, 1], 0, angle = 25, length = 0.1, lwd = 2)
mtext(paste0("A     ", "Ct = ", signif(res_th.cyc[1, 1], 4)), cex = 1.2, side = 3, adj = 0, font = 2)

plot(testdat[, 1], log(testdat[, 2]), xlab = "Cycles", ylab = "log(Raw fluorescence)")
abline(h = log(res_th.cyc[1, 2]), col = "black")
arrows(res_th.cyc[1, 1], log(res_th.cyc[1, 2]), res_th.cyc[1, 1], min(log(testdat[, 2]), na.rm = TRUE), angle = 25, length = 0.1, lwd = 2)
mtext(paste0("B     ", "Ct = ", signif(res_th.cyc[1, 1], 4)), cex = 1.2, side = 3, adj = 0, font = 2)

res_efficiency <- efficiency(res_model)
cpDdiff <- sqrt((res_efficiency$cpD1 - res_efficiency$cpD2)^2)

arrows(res_takeoff[[1]], res_takeoff[[2]], res_takeoff[[1]], -0.2, angle = 25, length = 0.1, lwd = 2)

abline(v = 19.5)

mtext(paste0("C   ",  "cpDdiff: ", cpDdiff), cex = 1.2, side = 3, adj = 0, font = 2)
```


\newpage


## Technologies for Amplification Curve Classification and Classified Amplification Curves\label{section_data_sets}

Many machine learning concepts exist. One method is supervised machine learning, where the goal is to derive a property from user-defined (classified) training data. Categories such as negative, ambiguous or positive are assigned depending on the form of the amplification curve. An extensive literature research showed that there are no openly accessible classified amplification curve datasets. Open Data is meant in the sense that data are freely available, free of charge, free to use and that data can be republished, without restrictions from copyright, patents or other mechanisms of control [@kitchin2014].

Therefore, a large number of records with amplification curves and their classification (negative, ambiguous, positive) were added to the \texttt{PCRedux} package.

For the amplification curves in \autoref{table-datasets}, a dichotomous classification was performed (roughly sigmoid or negative amplification reaction with a flat curve shape). Consequently, this does not rule out 

- if a specific amplification product has been synthesized, 
- if a contamination has been amplified or 
- if only primer-dimers have been amplified. 

To answer this question, other methods such as agarosegel electrophoresis need to be used.

### Manual Amplification Curve Classification\label{Manual_Amplification_Curve_Classification}

For machine learning and method validation it was important to classify the amplification curves individually. In @roediger2015chippcr the  ``humanrater()``  function was introduced. This function was developed to help the user during the classification of amplification curves and melting curves. The user has to define classes (e.&nbsp;g., negative ("n"), ambiguous ("a"), positive ("p")) which get assigned to an amplification curve after expert has entered the class in input mask. All amplification curve datasets listed in \autoref{table-datasets} were classified in interactive, semi-blinded sessions. ``humanrater()`` was set to randomly select individual amplification curves. All datasets were manually classified at least three times. The `htPCR` dataset (\autoref{figure_sigmoid_curve}B) was in total classified eight times (see \autoref{figure_curve_classification}). Most of the amplification curves are neither unequivocal classifiable as positive or negative.


```{r figure_curve_classification, echo=TRUE, eval=TRUE, fig.cap=figure_curve_classification, fig.scap=figure_curve_classification_short}
# Suppress messages and load the packages for reading the data of the classified
# amplification curves.
library(data.table)
library(PCRedux)

# Load the decision_res_htPCR.csv dataset from a csv file.
filename <- system.file("decision_res_htPCR.csv", package = "PCRedux")
decision_res_htPCR <- fread(filename, data.table = FALSE)

#
par(mfrow = c(2, 4))
for (i in 2L:9) {
    data_tmp <- table(as.factor(decision_res_htPCR[, i]))
    
    barplot(data_tmp, col = adjustcolor("grey", alpha.f = 0.5),
            xlab = "Class", ylab = "Counts", border = "white")
    text(c(0.7, 1.9, 3.1), rep(quantile(data_tmp, 0.25), 3), data_tmp, srt = 90)
    mtext(LETTERS[i - 1], cex = 1.2, side = 3, adj = 0, font = 2)
}
```

This approach is well suited and has been applied to classify a variety of amplification curves during the development of the \texttt{PCRedux} package. From experience this is time-consuming and tiring for large datasets, especially when the amplification curves are similar in shape. A high similarity between amplification curves exists, for example, in replicates and negative controls.

### Shape-based Group-wise Classification of Amplification Curves\label{shape_Amplification_Curve_Classification}

The similarity of amplification curves can be used to form groups of similar shapes. The amplification curves in the groups can then be classified in a bulk. In this way, a higher throughput can be achieved. This concept has not been described for the analysis of qPCR data in the literature. 

The ``tReem()`` function was developed to perform a *shape-based group classification*. To use the ``tReem()`` function, the first column must contain the qPCR cycles and all subsequent columns must contain the amplification curves. Two measures of similarity are used within the ``tReem()`` function.

- In the first measure (default), the  Pearson *correlation coefficient*s (*r*) are determined in pairs for all combinations of the amplification curves. The correlation coefficient is a statistical measure to describe the strength of the correlation between two or more variables. The correlation coefficient *r* is regarded as distance between the amplification curves. *r* is a dimensionless value and only takes values between -1 and 1. If *r = -1*, there is a maximum reciprocal relationship. If *r = 0* there is no correlation between the two variables. If *r = 1*, there is a maximum rectified correlation.

- In the second measure, the *Hausdorff distance* is used to determine the similarity between amplification curves. The Hausdorff distance is "the maximum of the distances from a point in any of the sets to the nearest point in the other set" [@rote_computing_1991, @herrera_multiple_2016]. The amplification curves are converted within the ``tReem()`` function using the ``qPCR2data()`` function.

Both methods process the distances in the same steps. This involves the calculation of the distance matrix using the Euclidean distances of all distance measures to determine the distance between the lines of the data matrix. 

This is used to perform a hierarchical cluster analysis. In the last step, the cluster is divided into groups based on a user-defined *k* value. For example, two groups are created for *k = 2*. If the amplification curves shapes are highly diverse, a larger *k* should be used. After a chain of processing steps presents the ``tReem()`` function a series of plots with grouped of amplification curves. The corresponding classes can then be assigned to the groups of amplification curves by the user using an input mask.

Grouping the amplification curves with the Pearson correlation coefficient as a distance measure is usually faster than the Hausdorff distance. The Hausdorff distance is an approximation of a shape metrics to define similarity measures between shapes. [@charpiat_shape_2003].


```{r, echo=TRUE, eval=FALSE}
# Classify amplification curve data by correlation coefficients (r)
library(qpcR)
classification_result <- tReem(testdat[, 1:15], k = 3)
classification_result
```

#### ``decision_modus()`` - A Function to Get a Decision (Modus) from a Vector of Classes \label{section_decision_modus}

For the systematic statistical analysis of classification datasets, the ``decision_modus()`` function has been developed. This allows the most common decision (mode) to be determined. The mode is useful to consolidate large collections of different decisions into a single (most frequent) decision. 

> Observed:*a*, *a*, *a*, *a*, *a*, *n*, *n*, *n* $\rightarrow$ frequencies 5 x *a*, 3 x *n* $\rightarrow$ mode: *a*. Since the class names are known, they only have to be interpreted by the user (e.&nbsp;g., "a",*n*,"y" -> "ambivalent","negative","positive"). 

A manual classification was performed out for the `htPCR` dataset (for an example plot \autoref{figure_sigmoid_curve}B) with the ``humanrater()`` function. The classification of each amplification curve was performed eight times at different time points since many of the amplification curves did not resemble optimal curvatures (e.g., \autoref{htPCR_nap}). It is likely that the amplification curve (`r colnames (htPCR)[512]`, \autoref{htPCR_nap}) is considered as ambiguous or even positive (positive $\leftrightarrow$ ambivalent) by the users. 

\autoref{tableheaddecision} shows from a total of `r ncol(qpcR:: htPCR)-1` amplification curves the first 25 lines classified as negative (*conformity=TRUE*) and the first 25 lines classified as positive. Since in total, the curves were classified eight times (`test.result.1` $\ldots$ `test.result.8`) a whole of 70864 amplification curves was analysed. In this classification experiment the amplification curves have been classified differently in `r paste0(signif(sum(decision_res_htPCR$conformity == FALSE) / nrow(decision_res_htPCR) * 100, 3), "%")` of the cases (e.&nbsp;g., line 1 "P01. W01").


```{r, eval=TRUE, echo=FALSE, results='asis'}
library(PCRedux)
library(xtable)
library(data.table)
library(magrittr)

# filename <- system.file("decision_res_htPCR.csv", package = "PCRedux")
# decision_res_htPCR <- fread(filename, data.table = FALSE)

print(
    xtable(
        rbind(
            head(subset(decision_res_htPCR, conformity == FALSE), 25),
              head(subset(decision_res_htPCR, conformity == TRUE), 25)), 
           caption = "Results of the `htPCR` dataset classification. 
           All amplification curves of the `htPCR` dataset were classified as 
           `negative`, `ambiguous` and `positive` by individuals in eight 
           analysis cycles (`test.result.1` $\\ldots$ `test.result.8`). If an 
           amplification curve has always been classified with the same class, 
           the last column (`conformity`) shows `TRUE`. As an example, the table 
           shows 25 amplification curves with consistent classes and 25 
           amplification curves with differing classes (`conformity = FALSE`).", 
           label = "tableheaddecision" ), include.rownames = FALSE, 
      comment = FALSE, caption.placement = "top", size = "\\tiny"
)
```

The ``decision_modus()`` function was applied to the record `decision_res_htPCR.csv` with all classification rounds (columns 2 to 9) and the mode was determined for each amplitude curve \autoref{htPCR_nap_frequency}.

```{r, eval=TRUE, echo=TRUE}
# Use decision_modus() to go through each row of all classification done by
# a human.

# Determine the number of observations where all classifications were 
# the same (conformity == TRUE).
conformity <- decision_res_htPCR[["conformity"]]

# List all classifications.
dec <- lapply(1L:nrow(decision_res_htPCR), function(i) {
    decision_modus(decision_res_htPCR[i, 2:9])[1]
}) %>% unlist()

# Show statistic of the decisions
summary(dec)
```

```{r htPCR_nap_frequency, echo=FALSE, fig.cap=htPCR_nap_frequency, fig.scap=htPCR_nap_short_frequency, fig.higth=3}
# Plot the Frequencies of the decisions

layout(matrix(c(1,1,2), 1, 3, byrow = TRUE))
barplot(
    table(dec), xlab = "Class decision", ylab = "Frequency",
        main = "", col = c("red", "black", "green"), border = "white"
)
text(c(0.7, 1.9, 3.1), rep(4000,3), 
        c(paste("n = ", table(dec)[1]), 
          paste("n = ", table(dec)[2]), 
          paste("n = ", table(dec)[3])),
    col = c("black", "white", "black"))
mtext("A", cex = 1.2, side = 3, adj = 0, font = 2)

barplot(
    table(conformity), xlab = "Comformity", ylab = "Frequency",
        main = "", col = c("black", "green"), border = "white"
)
text(c(0.7, 1.85), rep(8000,2), 
        c(paste("n = ", table(conformity)[1]), 
        paste("n = ", table(conformity)[2])), 
        col = c("white", "black"))
mtext("B", cex = 1.2, side = 3, adj = 0, font = 2)
```

Another usage mode of ``decision_modus()`` is to set the parameter as `max_freq=FALSE`. This option specifies the number of all classifications.

```{r, eval=TRUE, echo=TRUE}
library(PCRedux)
# Decisions for observation P01.W06
res_dec_P01.W06 <- decision_modus(decision_res_htPCR[
which(decision_res_htPCR[["htPCR"]] == "P01.W06"),
                                  2L:9
], max_freq = FALSE)
print(res_dec_P01.W06)
```

The amplification curve `P01. W06` was classified as `r paste0(res_dec_P01.W06$variable[1], "=", res_dec_P01.W06$freq[1])` times and as `r paste0(res_dec_P01.W06$variable[2], "=", res_dec_P01.W06$freq[2])` times. Therefore, the decision would turn into a `negative` decision.


### Classified Amplification Curve Datasets\label{chapter_humanrater}

Amplification curves from different sources (e.g., detection chemistries, thermo-cyclers) were manually classified with the ``humanrater()`` function (\autoref{Manual_Amplification_Curve_Classification}) or with the ``tReem()`` function  (\autoref{shape_Amplification_Curve_Classification}). Raw amplification curve data were exported as comma separated values or in the Real-time PCR Data Markup Language (RDML) format via the \texttt{RDML} package. RDML is human readable data exchange format for qPCR experiments. A detailed description can be found in @roediger_enabling_2017. The following code section describes the import of an RDML file from the \texttt{PCRedux} package. The RDML file contains amplification curve data of a duplex qPCR (HPV 16 & HPV 18) performed in the CFX96 (Bio-Rad).

```{r, echo=TRUE, eval=FALSE}
library(RDML)
# Load the RDML package and use its functions to import the amplification curve
#  data
library(RDML)
filename <- system.file("RAS002.rdml", package = "PCRedux")
raw_data <- RDML$new(filename = filename)
```

The following example shows the export of the `RAS002.rdml` file from the RDML format to the csv format.

```{r, echo=TRUE, eval=FALSE}
# Export the RDML data from the PCRedux package as the objects RAS002 and RAS003.
library(RDML)
library(PCRedux)
library(magrittr)
library(data.table)

RAS002 <- data.frame(RDML$new(paste0(
    path.package("PCRedux"),
                                     "/", "RAS002.rdml"
))$GetFData())

# The obbject RAS002 can be stored in the working directory as CSV file with
# the name RAS002_amp.csv.
write.csv(RAS002, "RAS002_amp.csv", row.names = FALSE)
```

RDML data file           | Device    | Target gene           | Detection chemistry
-------------------------|-----------|-----------------------|------------------------------
RAS002.rdml              | CFX96     | HPV16, HPV18, HPRT1   | Taqman
RAS003.rdml              | CFX96     | HPV16, HPV18, HPRT1   | Taqman
hookreg.rdml             | Bio-Rad   | various               | Taqman, DNA binding dyes

\begin{table}[]
\centering
\caption{Classified amplification curve datasets. Decision Datasets in \texttt{PCRedux}: table with results of manual classification as comma separated values. qPCR Dataset: name of original amplification curve data set. Package: name of the \texttt{R} package containing the amplification curves. Device: is the device used to measure the amplification reaction. \texttt{Note: The original data sets contain inofrmation about the detection chemistry used within the corresponding qPCR experiments.} AB, Applied Biosystems.}
\label{table-datasets}
\begin{tabular}{llll}
\hline
Decision Datasets in PCRedux                    & qPCR Dataset               & Package              & Device         \\ \hline
decision\_res\_RAS002.csv                        & RAS002.rdml                 & \texttt{PCRedux}   & CFX96, Bio-Rad \\
decision\_res\_RAS003.csv                        & RAS003.rdml                 & \texttt{PCRedux}   & CFX96, Bio-Rad \\
decision\_res\_batsch1.csv                       & batsch1                     & \texttt{qpcR}      & Light Cycler 1.0, Roche \\
decision\_res\_batsch2.csv                       & batsch2                     & \texttt{qpcR}      & Light Cycler 1.0, Roche \\
decision\_res\_batsch3.csv                       & batsch3                     & \texttt{qpcR}      & Light Cycler 1.0, Roche \\
decision\_res\_batsch4.csv                       & batsch4                     & \texttt{qpcR}      & Light Cycler 1.0, Roche \\
decision\_res\_batsch5.csv                       & batsch5                     & \texttt{qpcR}      & Light Cycler 1.0, Roche \\
decision\_res\_lc96\_bACTXY.csv                  & lc96\_bACTXY.rdml           & \texttt{RDML}      & Light Cycler 1.0, Roche \\
decision\_res\_boggy.csv                         & boggy                       & \texttt{qpcR}      & Light Cycler 96, Roche \\
decision\_res\_C126EG595.csv                     & C126EG595                   & \texttt{chipPCR}   & Chromo4, Bio-Rad \\
decision\_res\_C127EGHP.csv                      & C127EGHP                    & \texttt{chipPCR}   & iQ5, Bio-Rad \\
decision\_res\_C316.amp.csv                      & C316.amp                    & \texttt{chipPCR}   & iQ5, Bio-Rad \\
decision\_res\_C317.amp.csv                      & C317.amp                    & \texttt{chipPCR}   & iQ5, Bio-Rad \\
decision\_res\_C60.amp.csv                       & C60.amp                     & \texttt{chipPCR}   & iQ5, Bio-Rad \\
decision\_res\_CD74.csv                          & CD74                        & \texttt{chipPCR}   & iQ5, Bio-Rad \\
decision\_res\_competimer.csv                    & competimer                  & \texttt{qpcR}      & Light Cycler 480, Roche \\
decision\_res\_dil4reps94.csv                    & dil4reps94                  & \texttt{qpcR}      & CFX384, Bio-Rad \\
decision\_res\_guescini1.csv                     & guescini1                   & \texttt{qpcR}      & Light Cycler 480, Roche \\
decision\_res\_guescini2.csv                     & guescini2                   & \texttt{qpcR}      & Light Cycler 480, Roche \\
decision\_res\_htPCR.csv                         & htPCR                       & \texttt{qpcR}      & Biomark HD, Fluidigm \\
decision\_HCU32\_aggR.csv                        & HCU32\_aggR.csv             & \texttt{PCRedux}   & VideoScan \\
decision\_res\_karlen1.csv                       & karlen1                     & \texttt{qpcR}      & ABI Prism 7700, AB \\
decision\_res\_karlen2.csv                       & karlen2                     & \texttt{qpcR}      & ABI Prism 7700, AB \\
decision\_res\_karlen3.csv                       & karlen3                     & \texttt{qpcR}      & ABI Prism 7700, AB \\
decision\_res\_lievens1.csv                      & lievens1                    & \texttt{qpcR}      & ABI7300, ABI \\
decision\_res\_lievens2.csv                      & lievens2                    & \texttt{qpcR}      & ABI7300, ABI \\
decision\_res\_lievens3.csv                      & lievens3                    & \texttt{qpcR}      & ABI7300, ABI \\
decision\_res\_reps.csv                          & reps                        & \texttt{qpcR}      & MXPro3000P, Stratagene \\
decision\_res\_reps2.csv                         & reps2                       & \texttt{qpcR}      & MXPro3000P, Stratagene \\
decision\_res\_reps3.csv                         & reps3                       & \texttt{qpcR}      & MXPro3000P, Stratagene \\
decision\_res\_reps384.csv                       & reps384                     & \texttt{qpcR}      & CFX384, Bio-Rad \\
decision\_res\_rutledge.csv                      & rutledge                    & \texttt{qpcR}      & Opticon 2, MJ Research \\
decision\_res\_stepone\_std.csv                  & stepone\_std                & \texttt{RDML}      & StepOne, AB
\\
decision\_res\_testdat.csv                       & testdat                     & \texttt{qpcR}      & Light Cycler 1.0, Roche \\
decision\_res\_vermeulen1.csv                    & vermeulen1                  & \texttt{qpcR}      & Light Cycler 480, Roche \\
decision\_res\_vermeulen2.csv                    & vermeulen2                  & \texttt{qpcR}      & Light Cycler 480, Roche \\
decision\_res\_VIMCFX96\_60.csv                  & VIMCFX96\_60                & \texttt{chipPCR}   & CFX96, Bio-Rad \\ \hline
\end{tabular}
\end{table}


\newpage


## Data Analysis Functions of the \texttt{PCRedux} Package \label{section_Functions_of_PCRedux}

The \texttt{PCRedux} package contains functions for analyzing amplification curves. In the following, these are distinguished into helper functions  (\autoref{section_helper_functions}) and analysis functions (\autoref{section_analysis_functions}). 

### Helper Functions of the \texttt{PCRedux} Package  \label{section_helper_functions}

#### ``performeR()`` - Performance Analysis for Binary Classification \label{section_performeR}

Statistical modeling and machine learning is powerful but expose a risk to the user by introducing an unexpected bias. This may lead to an overestimation of the performance. The assessment of the performance by the sensitivity and specificity is fundamental to characterize a classifier or screening test [@james_introduction_2013]. Sensitivity is the percentage of true decisions that are identified and specificity is the percentage of negative decision that are correctly identified (\autoref{table-performance}). An example for the application of the ``performeR()`` function is shown in \autoref{section_autocorrelation_test}.


\begin{table}[]
\centering
\caption{Measures for performance analysis for binary classification. TP, true positive; FP, false positive; TN, true negative; FN, false negative}
\label{table-performance}
\begin{tabular}{ll}
Measure                                    & Formula   \\ \hline
Sensitivity - TPR, true positive rate      & $TPR=\frac{TP}{TP + FN}$   \\
Specificity - SPC, true negative rate      & $SPC=\frac{TN}{TN + FP}$   \\
Precision - PPV, positive predictive value & $PPV=\frac{TP}{TP +  FP}$   \\
Negative predictive value - NPV            & $NPV=\frac{TN}{TN + FN}$   \\
Fall-out, FPR, false positive rate         & $FPR=\frac{FP}{FP + TN}=1 - SPC$   \\
False negative rate - FNR                  & $FNR=\frac{FN}{TN + FN}=1 - TPR$   \\
False discovery rate - FDR                 & $FDR=\frac{FP}{TP + FP}=1 - PPV$   \\
Accuracy - ACC                             & $ACC= \frac{(TP + TN)}{(TP + FP + FN + TN)}$   \\
F1 score - F1                              & $F1=\frac{2TP}{(2TP + FP + FN)}$   \\
Matthews correlation coefficient - MCC     & $MCC=\frac{(TP*TN - FP*FN)}{\sqrt{(TP+FP)*(TP+FN)*(TN+FP)*(TN+FN)}}$   \\
Likelihood ratio positive - LRp            & $LRp=\frac{TPR}{1-SPC}$   \\
Cohen''s kappa (binary classification)     & $\kappa=\frac{p_{0}-p_{c}}{1-p_{0}}$   \\ \hline
\hline
\end{tabular}
\end{table}


#### ``qPCR2fdata()`` - A Helper Function to Convert Amplification Curve Data to the `fdata` Format \label{section_qPCR2fdata}

``qPCR2fdata()`` is a helper function to convert amplification curve data to the functional `fdata` class [@Febrero_Bande_2012]. The `fdata` format is used for functional data analysis to determine the similarity measures between amplification curves shapes by the Hausdorff distance. Similarity herein refers to the difference in spatial location of two \textit{objects} (e.&nbsp;g., amplification curves). Objects with a close distance are presumably more similar. For single objects (e.&nbsp;g., points) one can use a vector distance, such as the Euclidean distance [@herrera_multiple_2016].

The ``qPCR2fdata()`` function takes a `data.frame` containing the amplification cycles (first column) and the fluorescence amplitudes (subsequent columns) as input. 

Noise and missing values may affect the analysis adversely. Therefore, an instance of the ``CPP()`` function (\texttt{chipPCR} package [@roediger2015chippcr]) was integrated in ``qPCR2fdata()``. If \textit{preprocess=TRUE} in ``qPCR2fdata()``, then all curves are smoothed (Savitzky-Golay smoother), missing values are imputated and outliers in the ground phase get removed as described in @roediger2015chippcr.

The following example illustrates a hierarchical cluster analysis the `testdat` dataset. The amplification curves of the `testdat` dataset remained as raw data or were pre-processed (smoothed). Subsequent, the amplification curves were converted by the ``qPCR2fdata()``. The converted data were subjected to a cluster analysis (Hausdorff distance). This method uses the elements of a proximity matrix to generate a dendrogram. The dendrogram can can be used to further analyze the clusters. There are methods to determine the number of clusters automatically \textit{k} [@cook_interactive_2007]. However, for simplicity the number of clusters was determined visually.

The distance based on the Hausdorff metric was already done the next steps involved the ``cutree()`` function from the \texttt{stats} package to split the dendrogram into smaller junks. *A priori* was defined that two classes (\textit{positive} & \textit{negative}) are expected. Therefore, the \textit{group} parameter was set to \textit{k}=2 in the ``cutree()``.


```{r echo=TRUE, message=FALSE, warning=FALSE}
# Calculate the Hausdorff distance of the amplification curves
# cluster the curves.
# Load additional packages for data and pipes.
library(qpcR)
library(chipPCR)
library(fda.usc)
library(magrittr)

# Convert the qPCR dataset to the fdata format
# Use unprocessed data from the testdat dataset
res_fdata <- qPCR2fdata(testdat)

# Extract column names and create rainbow color to label the data
columnames <- testdat[-1] %>% colnames()
colors <- rainbow(length(columnames), alpha = 0.5)

# Calculate the Hausdorff distance (fda.usc) package and plot the distances
# as clustered data.

res_fdata_hclust <- metric.hausdorff(res_fdata)
res_hclust <- hclust(as.dist(res_fdata_hclust))
```

The distance based on the Hausdorff metric was already done the next steps involved the ``cutree()`` function from the \texttt{stats} package to split the dendrogram into smaller junks. *A priori* was defined that two classes (\textit{positive} & \textit{negative}) are expected. Therefore, the \textit{group} parameter was set to \textit{k}=2 in the ``cutree()``.

```{r echo=TRUE, message=FALSE, warning=FALSE}
# Cluster of the unprocessed amplification curves
res_cutree <- cutree(res_hclust, k = 2)
res_cutree <- factor(res_cutree)
levels(res_cutree) <- list(y = "1", n = "2")
```

The dendrogram shows that

- the observations are correctly assigned to a cluster of positive or negative amplification curves and that 
- the shift of the Cq (late increase of the fluorescence) is reflected in the positive cluster (\autoref{qPCR2fdata}).


```{r qPCR2fdata, echo=TRUE, fig.cap=qPCR2fdata, fig.scap=qPCR2fdata_short, fig.height=4, fig.width=8}
# Plot the converted qPCR data
par(mfrow = c(1, 2))
res_fdata %>% plot(
    ., xlab = "Cycles", ylab = "RFU", main = "", type = "l",
    lty = 1, lwd = 2, col = colors
)
legend(
    "topleft", paste0(as.character(columnames), ": ", res_cutree),
       pch = 19, col = colors, bty = "n", ncol = 2, cex = 0.7
)
mtext("A", cex = 1.2, side = 3, adj = 0, font = 2)

plot(res_hclust, main = "", xlab = "", sub="")
text(c(5.5,18), rep(6.5,2), c("negative", "positive"), col = c("black", "green"), cex = 0.9, srt = 90)
mtext("B", cex = 1.2, side = 3, adj = 0, font = 2)
```

This workflow can be used to cluster amplification curve data according to their shape into groups of amplification curves with similar shape. Classification tasks can be preformed in batches of amplification curves. The calculation of the distances is a computing expensive step dependent on the number of amplification curves. 

The following example illustrates the usage for the `HCU32_aggR.csv` dataset 
from the VideoScan platform with 32 heating and cooling units (equivalent of 32 PCR vessels). In this experiment the
bacterial gene *aggR* from *E. coli* was amplified in 32 replicate qPCR reactions. Details of the experiment are described in the manual of the \texttt{PCRedux} package.
The ambition was to test if the 32 amplification curves of the qPCR reaction 
are identical. As before, the data were processed with the ``qPCR2fdata()`` 
function and compared by the Hausdorff distance. Ideally, the amplification 
curves form only few clusters.

```{r echo=TRUE, message=FALSE, warning=FALSE}
# Calculate slope and intercept on positive amplification curve data from the
# VideoScan 32 cavity real-time PCR device.
# Load additional packages for data and pipes.
library(data.table)
library(fda.usc)
library(magrittr)

# Load the qPCR data from the HCU32_aggR.csv dataset
# Convert the qPCR dataset to the fdata format

filename <- system.file("HCU32_aggR.csv", package = "PCRedux")
data_32HCU <- data.table::fread(filename, data.table = FALSE)

res_fdata <- qPCR2fdata(data_32HCU)
# Extract column names and create rainbow color to label the data
columnames <- data_32HCU[-1] %>% colnames()
colors <- rainbow(length(columnames), alpha = 0.55)
```
In advance the Cq values were calculated by the following code:

```{r, echo=TRUE, eval=FALSE}
# Load the qpcR package to calculate the Cq values by the second derivative
# maximum method.
library(qpcR)

res_Cq <- sapply(2L:ncol(data_32HCU), function(i) {
    efficiency(pcrfit(data_32HCU, cyc = 1, fluo = i, model = l6))
})

data.frame(
    obs = colnames(data_32HCU)[-1],
           Cq = unlist(res_Cq["cpD2", ]), eff = unlist(res_Cq["eff", ])
)

#        Results
#
# obs    Cq      eff
# 1      A1 14.89 1.092963
# 2      B1 15.68 1.110480
# 3      C1 15.63 1.111474
# ...
# 30     F4 15.71 1.109634
# 31     G4 15.70 1.110373
# 32     H4 15.73 1.117827
```

Next, the amplification curves (\autoref{HCU32}A), the differences between base-line region and plateau region (\autoref{HCU32}B), the correlation between the Cq value and amplification efficiency (\autoref{HCU32}C) and the clusters based on the Hausdorff distance were taken into account.

Some amplification curves (\autoref{HCU32}A) had stronger noise and all curves have a negative non-linear trend and a shift in the ground phase. The comparison of the ground phase and the plateau phase showed a difference between the 32 amplification curves. The observations `E1`, `F1` and `H1` were most close in the ground phase and plateau phase. The comparison of Cq values and amplification efficiency showed that most amplification curves are similar. However, there are also amplification curves that show a greater deviation from the median of all Cq values (\autoref{HCU32}C). The cluster analysis confirmed the shape similarity (\autoref{HCU32}D).

```{r HCU32, echo=TRUE, fig.cap=HCU32, fig.scap=HCU32_short, fig.height=8.5, fig.width=11} 

library(fda.usc)
library(magrittr)

# To save computing time, the Cq values and amplification efficiencies were 
# calculated beforehand and transferred as a hard copy here.

calculated_Cqs <- c(
    14.89, 15.68, 15.63, 15.5, 15.54, 15.37, 15.78, 15.24, 15.94,
    15.88, 15.91, 15.77, 15.78, 15.74, 15.84, 15.78, 15.64, 15.61,
    15.66, 15.63, 15.77, 15.71, 15.7, 15.79, 15.8, 15.72, 15.7, 15.82,
    15.62, 15.71, 15.7, 15.73
)

calculated_effs <- c(
    1.09296326515231, 1.11047987547324, 1.11147389307153, 1.10308929700635,
    1.10012176315852, 1.09136717687619, 1.11871308210321, 1.08006168654712,
    1.09500422011318, 1.1078777171126, 1.11269436700649, 1.10628580163733,
    1.1082009954558, 1.11069683827291, 1.11074914659374, 1.10722949813473,
    1.10754282514113, 1.10098387264025, 1.1107026749644, 1.11599641663658,
    1.11388510347017, 1.11398547396991, 1.09410798249025, 1.12422338092929,
    1.11977386646464, 1.11212436173214, 1.12145338871426, 1.12180879952503,
    1.1080276005651, 1.10963449004393, 1.11037302758388, 1.11782689816295
)

# Plot the converted qPCR data
layout(matrix(c(1, 2, 3, 4, 4, 4), 2, 3, byrow = TRUE))
res_fdata %>% plot(
    ., xlab = "Cycles", ylab = "RFU", main = "HCU32_aggR", type = "l",
    lty = 1, lwd = 2, col = colors
)
legend(
    "topleft", as.character(columnames), pch = 19,
       col = colors, bty = "n", ncol = 4
)
mtext("A", cex = 1.2, side = 3, adj = 0, font = 2)

# Plot the background and plateau phase.

boxplot(
    data_32HCU[, -1] - apply(data_32HCU[, -1], 2, min),
        col = colors, las = 2, main = "Signal to noise ratio",
        xlab = "Sample", ylab = "RFU"
)
mtext("B", cex = 1.2, side = 3, adj = 0, font = 2)

# Plot the Cqs and the amplification efficiencies.
# Determine the median of the Cq values and label all Cqs, which a less 0.1 Cqs
# of the median or more then 0.1 Cqs of the median Cq.

plot(
    calculated_Cqs, calculated_effs, xlab = "Cq (SDM)",
     ylab = "eff", main = "Cq vs. Amplification Efficiency",
     type = "p", pch = 19, lty = 1, lwd = 2, col = colors
)

median_Cq <- median(calculated_Cqs)
abline(v = median_Cq)

text(median_Cq + 0.01, 1.085, expression(paste(tilde(x))))
labeled <- c(
    which(calculated_Cqs < median_Cq - 0.1),
             which(calculated_Cqs > median_Cq + 0.1)
)

text(
    calculated_Cqs[labeled], calculated_effs[labeled],
     as.character(columnames)[labeled]
)
mtext("C", cex = 1.2, side = 3, adj = 0, font = 2)

# Calculate the Hausdorff distance using the fda.usc package and cluster the
# the distances.

res_fdata_hclust <- metric.hausdorff(res_fdata)
cluster <- hclust(as.dist(res_fdata_hclust))

# plot the distances as clustered data and label the leafs with the Cq values
# and colored dots.

plot(cluster, main = "Clusters of the amplification\n
curves as calculated by the Hausdorff distance", xlab = "", sub="")
mtext("D", cex = 1.2, side = 3, adj = 0, font = 2)
```

The analysis gives an overview of the variation of the amplification curve data.

\newpage

### Amplification Curve Analysis Functions of the \texttt{PCRedux} package  \label{section_analysis_functions}

On the basis of this observation, various concepts were developed and implemented in algorithms to describe amplification curves. The function described following are aimed for experimental studies. It is important to note that the features proposed herein emerged during a critical reasoning process. The aim of the package is to propose a set of features, functions and data for an independent research.


#### ``pcrfit_single()`` and ``encu()``- Functions to Calculate Features from an Amplification Curve \label{section_pcrfit_single_pcrfit_parallel}

The following sections give a concise description of the algorithms used to calculate feature vectors by the ``pcrfit_single()`` function . Based on considerations and experience the algorithms of the ``pcrfit_single()`` function are restricted to ROIs (\autoref{amplification_curve_ROI}) to calculate specific features. 

> The ``encu()`` function is a wrapper for the  ``pcrfit_single()`` function.   ``encu()`` can be used to process large records of amplification curve data arranged in columns. The progress of processing is displayed in the form of a progress bar and the estimated run-time. Additionally, the ``encu()`` allows to specify which monitoring chemistry (e.&nbsp;g., DNA binding dye, sequence specific probes) and which thermo-cycler was used. @ruijter_2014 have shown that the monitoring chemistry and the type of input DNA (single stranded, double stranded) are important when analysing qPCR data, because they have an influence on the shape of the amplification curve. For simplicity, the documentation will describe the ``pcrfit_single()`` only.


The underlying hypotheses and concepts of the features are formulated and supported by  *exemplary applications*. Different representative data sets were used to support a concept or features. For example, the `RAS002` dataset represents a typical qPCR. This means that the positive amplification curves start with a flat plateau phase and then transition into the sigmoid shape with a plateau. The negative amplification curves display no significant peculiarities. For both positive and negative amplification curves there is a shift from the origin. The `htPCR` data set serves as a problem example in several places, since it contains many observations (amplification curves from high-throughput experiments). In addition, the amplification curves have a high diversity of curve shapes that cannot be uniquely and reproducibly classified even by experienced users. Other datasets are used in the documentation, but these are not discussed in detail. The goal is not to examine the limits of their applicability, but rather to prove their basic functionality. Similar concepts are presented in groups.  The algorithms are divided into the following broad categories:

- algorithms that determine slopes, signal levels,
- algorithms that determine turning points and
- algorithms that determine areas.

The algorithms in 

- ``earlyreg()`` (\autoref{section_earlyreg}),
- ``head2tailratio()`` (\autoref{section_head2tailratio}),
- ``hookreg()``& ``hookregNL()`` (\autoref{section_hookreg}) and 
- ``mblrr()`` (\autoref{section_mblrr}),
- ``autocorrelation_test()`` (\autoref{section_autocorrelation_test})

were implemented as standalone functions to make them available for other applications. 

The output below shows the features and their data type (`num`, numeric; `int`, integer; `Factor`, factor; `logi`, boolean) that were determined with the ``pcrfit_single()`` function.


```{r, echo=TRUE}
library(PCRedux)
# Calculate feature vector of column two from the RAS002 dataset.
str(pcrfit_single(RAS002[, 2]))
```

> To underscore the usability of the algorithms and their features, `r d <- PCRedux::data_sample; paste0(nrow(d), " observations ", "(", sum(d[["decision"]] == "n"), " negative amplification curves, ", sum(d[["decision"]] == "y"), " positive amplification curves)")` from the `batsch1`, `boggy`, `C126EG595`, `competimer`, `dil4reps94`, `guescini1`, `karlen1`, `lievens1`, `reps384`, `rutledge`, `testdat`, `vermeulen1`, `VIMCFX96_60`, `stepone_std`, `RAS002`, `RAS003`, `HCU32_aggR` and `lc96_bACTXY` were analyzed with the ``encu()`` function and the results (features) were combined in the file **`data_sample.rda`**. Users of this function should independently verify and validate the results of the methods for their applications.

##### Amplification Curve Pre-Processing

The ``pcrfit_single()`` function performs pre-processing steps before each calculation. That includes checking whether an amplification curve contains missing values. Missing values (NA) are measuring points in a dataset where no measured values are available or if they have been removed arbitrarily. NAs may occur if no measurement has been carried out (e.&nbsp;g., defective detector) or lengths of the vectors differ (number of cyles) between the observation. Such missing values are automatically imputed by spline interpolation as described in @roediger2015chippcr. 

All values of an amplification curve are normalized to their 99\% quantile. The normalization is used to equalize the amplitudes differences of amplification curves from thermo-cyclers (sensor technology, software processing) and detection chemistries. To compare amplification curves from different thermo-cyclers, the values should always be scaled systematically using the same method.\footnote{Although there are other normalization methods (e. g., \textit{minimum-maximum normalization}, see \cite{roediger2015chippcr}, the normalization by the 99\% quantile preserves the information about the level of the background phase. A normalization to the maximum is not used to avoid strong extenuation by outliers.} The data in \autoref{plot_bg_pt}D show that the `maxRFU` values after normalization are approximately 1. There is no statistical significant difference between `maxRFU` values of positive and negative amplification curves. `minRFU` is the minimum of the amplification curve, which is determined at the 1\% quantile to minimize the influence of outliers (\autoref{plot_bg_pt}C).

Selected algorithms of the ``pcrfit_single()`` function use the ``CPP()`` function from the \texttt{chipPCR} package to pre-process (e.&nbsp;g., base-lining, smoothing, imputation of missing values) the amplification curves. Further details are given in @roediger2015chippcr.

##### Handling of Missing Features

Missing values (NA) can occur if a calculation of a feature is impossible (e.g., if a logistic function cannot be adapted to noisy raw data). The lack of a feature is nevertheless a useful information (no feature calculate $\mapsto$ amplification curves deviate from sigmoid shape). The NAs were left unchanged in the \texttt{PCRedux} package up to version 0.2.5-1. Since version 0.2.6, however, the NAs have been replaced by numerical values (e.&nbsp;g., total number of cycles) or factors (e.&nbsp;g., *lNA* for non-fitted model). Under the term "imputation" there are a number of procedures based on statistical methods (e.&nbsp;g., neighboring median, spline interpolation) or on user-defined rules [@williams_rattle:_2009, @cook_interactive_2007, @hothorn_handbook_2014]. Rules are mainly used in function of \texttt{PCRedux} to relieve the user from the decision as to how to deal with missing values. For example, slope parameters of a model are to zero when it cannot be determined. The disadvantage is that rules do not necessarily concur to real world values. 

#### Multi-parametric Models for Amplification Curve Fitting\label{section_models}

Both the ``pcrfit_single()`` function and the ``encu()`` function use four multi-parametric models based on the findings by @spiess_highly_2008 and @Ritz2008. The ``pcrfit_single()`` function starts by adjusting a seven-parameter model since this adapts *easier* to an dataset (\autoref{plot_models}). 

* **l7**:

\begin{equation}\label{l7}
f(x) = c + k1 \cdot x + k2 \cdot x^2 + \frac{d - c}{(1 + exp(b(log(x) - log(e))))^f}
\end{equation}

From that model, the ``pcrfit_single()`` function outputs the variables `b_slope` and `f_intercept`, which describe the slope and the intercept. The number of iterations required to adapt the model is also stored. That value is returned by the ``pcrfit_single()`` function as `convInfo_iteratons`. The higher the `convInfo_iteratons` value, the more iterations were necessary to converge from the start parameters (\autoref{plot_dat_EffTop}D). A low `convInfo_iteratons` value is an indicator for a sigmoid curve shape. High numbers of iterations imply noisy or non-sigmoid curves (\autoref{plot_dat_EffTop}L). 

The amplification curve fitting process continues with four-parameter model (*l4*, \autoref{l4}). This is followed by a model with five parameters (*l5*, \autoref{l5}) and six parameters (*l6*, \autoref{l6}).

* **l4**: 

\begin{equation}\label{l4}
    f(x) = c + \frac{d - c}{1 + exp(b(log(x) - log(e)))}
\end{equation}


* **l5**: 

\begin{equation}\label{l5}
    f(x) = c + \frac{d - c}{(1 + exp(b(log(x) - log(e))))^f}
\end{equation}
                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                * **l6**: 

\begin{equation}\label{l6}
    f(x) = c + k \cdot x + \frac{d - c}{(1 + exp(b(log(x) - log(e))))^f}
\end{equation}

The optimal model is selected on the basis of the Akaike information criterion and used for all further calculations. The ``pcrfit_single()`` function returns `qPCRmodel` as a factor (*l4*, *l5*, *l6*, *l7*). In case no model could be fitted, an *lNA* is returned.

The model is an indicator of the amplification curve shape. Model with many parameters deviate more from an ideal sigmoid model. For instance, a four-parameter model, unlike the six-parameter model, does not have a linear component. A negative linear slope in the plateau phase is an indicator of a *hook effect* [CITATION BDQ].


```{r plot_models, echo=FALSE, fig.cap=plot_models, fig.scap=plot_models_short, fig.height=3.5}
library(PCRedux)

x <- data_sample$decision
y <- factor(data_sample[["qPCRmodel"]], levels = c("lNA", "l4", "l5", "l6", "l7"))

res_fw <- rbind(negative = table(y[x == "n"]),
                positive = table(y[x == "y"])
)


colors <- c(adjustcolor("black", alpha.f = 0.25), adjustcolor("green", alpha.f = 0.25))

par(mfrow = c(1,2))

barplot(res_fw / sum(res_fw) * 100, beside = TRUE, col = colors, xlab = "", 
        ylab = "Percentage", border = "white")
legend("top", c(
    paste0("Negative (n =", length(y[x == "n"]),")"), paste0("Positive (n =", length(y[x == "y"]),")")
), fill = colors, bty = "n")
mtext("A    Fitted models", cex = 1.2, side = 3, adj = 0, font = 2)


data <- data_sample
feature <- c("cpD2")

x <- data$decision

for(i in 1L:length(feature)) {
    y <- data[, colnames(data) == feature[i]]
    y_density_neg <- density(y[x == "n"])
    y_density_pos <- density(y[x == "y"])
    res <- stats::wilcox.test(y ~ x)
    h <- max(na.omit(y))
    l <- min(na.omit(y))
    h_text <- rep(h * 0.976, 2)
    
    par(bg=NA)
    stripchart(y ~ x, vertical = TRUE, ylab = "Cq (cpD2)",
               method = "jitter", pch = 20, cex = 0.7, 
               col = adjustcolor("black", alpha.f = 0.8), 
               ylim = c(l * 0.95, h * 1.05))
    
    polygon(y_density_neg$y * 5 + 1.35, y_density_neg$x, col = colors[1], border = NA)
    polygon(y_density_pos$y * 5 + 1.35, y_density_pos$x, col = adjustcolor("green", alpha.f = 0.25), border = NA)
    arrows(1,-0.25,1.5,25, length = 0.1)
    arrows(2,-0.25,1.5,5, length = 0.1)
    
    boxplot(y ~ x, outline = FALSE, add = TRUE, boxwex = 0.35)
    
    legend("topleft", paste0("P = ", signif(res[["p.value"]])), 
           cex = 1, bty = "n")
    
    mtext(paste0("B", "   ", feature[i]), cex = 1.2, side = 3, 
          adj = 0, font = 2, col = ifelse(signif(res[["p.value"]], 2) < 0.05, 
                                          "black", "red"))
}
```


#### Quantification Points, Ratios and Slopes\label{section_Areas_Ratios_Slopes}

The first derivative maximum (`cpD1`) and the second derivative maximum (`cpD2`) are used to describe the amplification reaction quantitatively (\autoref{plot_models}B). The ``pcrfit_single()`` function  calculates the `cpD1` and `cpD2` and uses them for further analysis. For example, low `cpD1` and `cpD2` values (< 5 cycles) indicate that the PCR reaction was negative or that the amount of input DNA was to high. Further features from the ``pcrfit_single()`` function are: 

- `eff` is the optimized PCR efficiency found within a sliding window (\autoref{figure_quntifcation_points}C). A linear model of cycles versus log(Fluorescence) is fit within a sliding window (for details see ``sliwin()`` function from the \texttt{qpcR} package). The comparison of positive and negative amplification curves in \autoref{plot_dat_EffTop}A demonstrates that the classes are significantly different from each other.
- `sliwin` is the PCR efficiency by the 'window-of-linearity' method [@spiess_highly_2008].
- `cpDdiff` is the difference between the Cq values calculated from the first and the second derivative maximum ($cpDdiff = |cpD1 - cpD2|$) **from the fitted model** (\autoref{figure_quntifcation_points}C). Provided that a model can be exactly fitted, the estimates of the difference are reliable. Higher `cpDdiff` values indicate a negative amplification reaction or a very low amplification efficiency. The comparison of positive and negative amplification curves in \autoref{plot_dat_EffTop}C demonstrates that the classes are significantly different from each other. In the event that the `cpDdiff` value cannot be determined (NA), it is replaced by zero.
- `cpD2_range` is the absolute value of the difference between the minimum and the maximum of the second derivative maximum ($cpD2\_range = |cpD2m - cpD2|$) from the ``diffQ2()`` function (**no model fitted**) (\autoref{figure_cpD2_range}). The `cpD2_range` value does not require an adjustment of a multiparametric model. The approximate first and second derivatives are determined using a five-point stencil [@roediger2015chippcr]. The comparison of positive and negative amplification curves in \autoref{plot_dat_EffTop}E shows that the classes differ significantly from each other. In the event that the `cpD2_range` value cannot be determined (NA), it is replaced by zero.


```{r figure_cpD2_range, echo=FALSE, results='hide', fig.cap=figure_cpD2_range, fig.scap=figure_cpD2_range_short, fig.height=4} 
library(chipPCR)
data <- cbind(RAS002[, 1], CPP(RAS002[, c(1,2)], method.norm = "minm")$y.norm)

# Invoke the inder function for the object data to interpolate 
# the derivatives of the simulated data as object res. The Nip 
# parameter was set to 5. This leads to smoother curves. res is
# an object of the class "der".
res <- inder(data[-c(1,2),], Nip = 5)

# Plot the object res and add descriptions to the elements.

par(las = 0, bty = "n", oma = c(.5,.5,.5,.5))

plot(data, xlab = "Cycles", ylab = "RFU", ylim = c(0,1),
     main = "", type = "b", pch = 20, lwd = 2)

# Add graphical elements for the derivatives and the calculated
# Cq values FDM, SDM and SDm.

lines(res[, "x"], res[, "d1y"], col = "blue", lwd = 2)
lines(res[, "x"], res[, "d2y"], col = "red", lwd = 2)

# Fetch the Cq values from res with the summary function
summ <- summary(res, print = FALSE)

abline(v = c(summ[[1]], summ[[2]], summ[[3]]), col = c("blue", "red", "red"), 
       lwd = 2)

text(summ[[2]] - 2, 0.5, paste("cpD2\n", round(summ["SDM"], 2)), 
     cex = 1.1, col = "red")
text(summ[[1]] + 2, 0.7, paste("cpD1\n", round(summ["FDM"], 2)), 
     cex = 1.1, col = "blue")
text(summ[[3]] + 2, 0.5, paste("cpD2m\n", round(summ["SDm"], 2)), 
     cex = 1.1, col = "red")

arrows(summ[[2]], 0.85, summ[[3]], 0.85, code = 3, length = 0.1)
text(summ[[1]] - 6, 0.85, paste("cpD2_range \n", 
                           round(abs(summ["SDM"] - summ["SDm"]), 2)), 
     cex = 1.1, col = "black")

legend("topleft", c("Amplification curve", "First derivative", "Second derivative"), 
       col = c(1,4,2), lty = c(2,1,1), bty = "n")

res <- data
background <- bg.max(res[-c(1,2), 1], res[-c(1,2), 2])


abline(v = background@bg.stop, col = "grey")
text(background@bg.stop, 0.3, "bg.stop", pos = 4, col = "grey", srt=90)
abline(v = background@amp.stop, col = "grey")
text(background@amp.stop, 0.3, "amp.stop", pos = 4, col = "grey", srt=90)
```

- `bg.stop` (\autoref{figure_cpD2_range}) is the end of the ground phase estimated by the ``bg.max()`` function [@roediger2015chippcr].
- `amp.stop` (\autoref{figure_cpD2_range}) is the end of the exponential phase estimated by the ``bg.max()`` function [@roediger2015chippcr].

- `top` is the *takeoff point*) as proposed by @tichopad_standardized_2003. The `top` is calculated using externally studentized residuals, which tested to be an outlier in terms of the t-distribution. The `top` signifies to first PCR cycle entering the exponential phase. 
- `tdp` is the *takedown point*. This is an implementation in the ``pcrfit_single()`` function, which uses the rotated $f(x) \mapsto f_{1}(f(x))$ and flipped $g(x) = -(x)$ amplification curve for calculation \autoref{amplification_curve_ROI}A describes the location of `top` and `tdp`. The position (`f.top`, `f.tdp`) on the ordinate is also determined from these points. If an amplification curve is negative or neither `top` nor `tdp` can be calculated, then `top` & `tdp` will be assigned the number of cycles and `f.top` & `f.tdp` the value 1.
The distribution of `top`, `tdp`, `f.top` and `f.tdp` is shown in \autoref{plot_dat_EffTop}I-L. This figure shows that a `top` value and a `tdp` value enable a qualitative classification of the amplification reaction. An interesting aspect is that the positive `f.top` values are markedly lower than the negative `f.top` values (\autoref{plot_dat_EffTop}J). The same applies inversely to the `tdp` values (\autoref{plot_dat_EffTop}L). In this way, amplification curves can be classified according to these values.
- `peaks_ratio` is based on a sequential chaining of functions. The ``diffQ()`` function (\\texttt{MBmca}) determines numerically the first derivative of an amplification curve. This derivative is passed to the ``mcaPeaks()`` function (\\texttt{MBmca}). In the output all minima and all maxima are contained. The ranges are calculated from the minima and maxima. The Lagged Difference is determined from the ranges of the minima and maxima. Finally, the ratio of the differences (maximum/minimum) is calculated.

```{r plot_dat_EffTop, echo=FALSE, fig.cap=plot_dat_EffTop, fig.scap=plot_dat_EffTop_short, fig.height=11, fig.width=11} 

library(PCRedux)

data <- data_sample[data_sample$dataset %in% c("stepone_std", "RAS002", "RAS003", "lc96_bACTXY", "C126EG595", "dil4reps94"), ]

feature <- c("eff", "sliwin", 
             "cpDdiff", "loglin_slope", "cpD2_range", "top", 
             "f.top", "tdp", "f.tdp", "bg.stop", 
             "amp.stop", "convInfo_iteratons")

par(mfrow = c(3,4))

x <- data$decision

for(i in 1L:length(feature)) {
    y <- data[, colnames(data) == feature[i]]
    res <- stats::wilcox.test(y ~ x)
    h <- max(na.omit(y))
    l <- min(na.omit(y))
    h_text <- rep(h * 0.976, 2)
    
    par(bg=NA)
    stripchart(y ~ x, vertical = TRUE, ylab = feature[i],
               method = "jitter", pch = 19, cex = 1, 
               col = adjustcolor("darkgrey", alpha.f = 0.65), 
               ylim = c(l * 0.95, h * 1.05))
    
    boxplot(y ~ x, outline = FALSE, add = TRUE, boxwex = 0.35)
    
    legend("topleft", paste0("P = ", signif(res[["p.value"]])), 
           cex = 1, bty = "n")
    
    mtext(paste0(LETTERS[i], "   ", feature[i]), cex = 1.2, side = 3, 
          adj = 0, font = 2, col = ifelse(signif(res[["p.value"]]) < 0.05, 
                                          "black", "red"))
}
```

- `loglin_slope` is calculated from the slope determined by a linear model of the data points from the cycle dependent fluorescence at the minimum of the second derivative and maximum of the second derivative (\autoref{loglin_slope}). Provided that the locations of the minimum of the second derivative and the maximum of the second derivative yield a *suitable* interval. As a precaution, the algorithm checks, for example, whether the distance between the minimum of the second derivative and the maximum of the second derivative is not more than nine PCR cycles. Failing this, the `loglin_slope` value is set to zero (no slope). In the following example, the data \autoref{loglin_slope}.


```{r loglin_slope, eval=TRUE, echo=FALSE, fig.cap=loglin_slope, fig.scap=loglin_slope_short, fig.height=3}
# Load example data (observation F6.1) from the testdat dataset
library(magrittr)
# Load MBmca package to calculate the minimum, and the maximum of the second
# derivative

library(MBmca)

par(mfrow = c(1, 2))

cyc <- 1

data <- RAS002[, c("cyc", "A01_gDNA.._unkn_B.Globin", "H10_ntc_ntc_B.Globin")]
col_names <- colnames(data)
data <- cbind(data[, 1], data[, 2]/quantile(data[, 2], .99), data[, 3]/quantile(data[, 3], .99))
colnames(data) <- col_names
# Calculate the minimum, and the maximum of the second
# derivative and assign it to the object res_diffQ2

rfu <- c(2,3)

for (i in rfu){
obs <- colnames(data)[i]
data_tmp <- data.frame(cyc = data[, cyc], rfu = data[, i])
res_diffQ2 <- suppressMessages(diffQ2(data_tmp, plot = FALSE, fct = min, inder = TRUE))
ROI <- round(c(res_diffQ2[[1]], res_diffQ2[[3]]))
# Build a linear model from der second derivative of res_diffQ2
res_loglin_lm <- lm(data_tmp[ROI, 2] ~ ROI)

data_tmp %>% plot(
    ., xlab = "Cycles", ylab = "normalized RFU", main = "", type = "l",
    lty = 1, lwd = 2, col = "black"
)
abline(res_loglin_lm, col = "red", lwd = 2)
abline(v = ROI, col = "grey", lwd = 2)
mtext(paste(LETTERS[i - 1], obs), side = 3, adj = 0, font = 1)
legend("topleft", paste("Slope: ", signif(
    coefficients(res_loglin_lm)["ROI"],
                                            4
)), bty = "n")
mtext("", side = 3, adj = 0, font = 2)
}
```

- `sd_bg` is the standard deviation from the first PCR cycle to the takeoff point (\autoref{amplification_curve_ROI}A). Manufacturers of thermo-cyclers use different sensors and data processing algorithms. Same applies to the detection chemistry used in experiments \autoref{why_PCRedux}. The signal variation in the ground phase differs between the different systems (\autoref{figure_sigmoid_curve_models}D). If no takeoff point can be determined from an amplification curve, the value for `sd_bg` is calculated from the first to the eighth PCR cycle. The results for the feature `sd_bg` were broken down by the thermo-cycler and the output of the amplification reaction (negative, positive). It can be seen that the signal variation between the thermo-cyclers seems to be different. There is also a difference between negative and positive amplification curves \autoref{plot_sd_bg}.


```{r plot_sd_bg, echo=FALSE, fig.cap=plot_sd_bg, fig.scap=plot_sd_bg_short, fig.height=7}
library(PCRedux)
data <- data_sample[, c("device", "decision", "sd_bg")]

devices <- unique(data[data[["decision"]] == "n", "device"])

par(mfrow = c(3,3))
for(i in 1L:length(devices)) {
    data_tmp <- data[data[, 1] == devices[i], ]
    x <- data_tmp[["decision"]]
    y <- data_tmp[["sd_bg"]]
    pos_neg <- summary(x)
    c(paste0("y (", pos_neg[["y"]], ")"), paste0("n (", pos_neg[["n"]], ")"))
    res <- stats::wilcox.test(y ~ x)
    
    par(bg=NA)
    stripchart(y ~ x, vertical = TRUE, 
               xlab = "Class decision",
               ylab = "Standard Deviation",
               method = "jitter", pch = 19, cex = 1, 
               col = adjustcolor("darkgrey", alpha.f = 0.65))
    
    boxplot(y ~ x, outline = FALSE, add = TRUE, boxwex = 0.35)
    legend("topleft", paste0("P = ", signif(res[["p.value"]])), 
           cex = 1, bty = "n")
    legend("bottomleft", c(paste0("y (", pos_neg[["y"]], ")"), 
                           paste0("n (", pos_neg[["n"]], ")")), 
           cex = 1, bty = "n")
    mtext(paste0(LETTERS[i], "   ", devices[i]), cex = 1.2, side = 3, 
          adj = 0, font = 2, col = ifelse(signif(res[["p.value"]]) < 0.05, 
                                          "black", "red"))
}
```


```{r plot_bg_pt, echo=FALSE, fig.cap=plot_bg_pt, fig.scap=plot_bg_pt_short, fig.height=11, fig.width=11}
library(PCRedux)

data <- data_sample[data_sample$dataset %in% c("stepone_std", "RAS002", "RAS003", "lc96_bACTXY", "C126EG595", "dil4reps94"), ]

# feature <- c("b_slope", "f_intercept", "minRFU", "maxRFU", 
#              "init2", "fluo", "slope_bg", "intercept_bg", 
#              "sd_bg", "head2tail_ratio")
feature <- c("b_slope", "f_intercept", "minRFU", "maxRFU", 
             "init2", "fluo", "slope_bg", "intercept_bg", 
             "sigma_bg", "sd_bg", "head2tail_ratio")

par(mfrow = c(3,4))

x <- data$decision

for(i in 1L:length(feature)) {
    y <- data[, colnames(data) == feature[i]]
    res <- stats::wilcox.test(y ~ x)
    h <- max(na.omit(y))
    l <- min(na.omit(y)) 
    h_text <- rep(h * 0.976, 2)
    
    par(bg=NA)
    stripchart(y ~ x, vertical = TRUE, ylab = feature[i],
               method = "jitter", pch = 19, cex = 1, 
               col = adjustcolor("darkgrey", alpha.f = 0.65), 
               ylim = c(l * 0.95, h * 1.05))
    
    boxplot(y ~ x, outline = FALSE, add = TRUE, boxwex = 0.35)
    
    legend("topleft", paste0("P = ", signif(res[["p.value"]])), 
           cex = 1, bty = "n")
    
    mtext(paste0(LETTERS[i], "   ", feature[i]), cex = 1.2, side = 3, 
          adj = 0, font = 2, col = ifelse(signif(res[["p.value"]]) < 0.05, 
                                          "black", "red"))
}
```

- `polyarea` is a feature that is based on the Gauss polygon area formula. The feature `polyarea` has been selected to show that the area under an amplification curve can be used to distinguish positive and negative amplification curves. The hypothesis is that positive amplification curves have a larger area than negative amplification curves. As shown in \autoref{plot_cp_area}C, there is a statistically significant difference between positive and negative amplification curves. 

- `amptester_polygon` is another method to calculate the area under an amplification curve. `amptester_polygon`\footnote{This feature is determines from the points in an amplification curve (like a polygon, in particular non-convex polygons) are in a `clockwise` order. The sum over the edges result in a positive value if the amplification curve is `clockwise` and is negative if the curve is `counter-clockwise`.} is part of the ``amptester()`` function from the \texttt{chipPCR} package [@roediger2015chippcr]. In contrast to the implementation in the ``amptester()`` is the `amptester_polygon` value normalized to the total number of cycles. This method may allow comparable predictions (\autoref{plot_cp_area}Ds). However, this question is not to be dealt with in the following example.


##### Examples of use cases

The `batsch1`, `HCU32_aggR`, `stepone_std`, `RAS002`, `RAS003`, `lc96_bACTXY` datasets were used for the calculation of the `polyarea` values for all amplification curves. A binomial logistic regression (aka logit regression or logit model (see \autoref{section_technologies_amplification_curves_ML})) was used to analyze the relationship between the `polyarea` value and the decision (negative, positive). This dataset contains almost equal proportions of positive and negative amplification curves (\autoref{plot_Logistic_Regression}A). Prior to this, the amplification curves were analyzed with the ``encu()`` function (\autoref{section_pcrfit_single_pcrfit_parallel}) and stored in the `data_sample.rda` file to save computing time. The file is part of the \texttt{PCRedux} package. The dataset was split into two chunks. This is an important step during such applications. One chunk is for adapting, i. e. training, the model and the other chunk for testing the model. By convention, 70% to 80% of the data is used for training [@walsh_correct_2015, @kuhn_building_2008]. The binomial logistic regression model was adapted using the function ``glm()`` by using the parameter `family = binomial(link = 'logit')`. To objectify the splitting, the ``sample()`` function was used.


```{r, echo=TRUE}
library(PCRedux)

data <- data_sample[data_sample$dataset %in% 
                c("batsch1",
                  "HCU32_aggR",
                  "lc96_bACTXY",
                  "RAS002", 
                  "RAS003", 
                  "stepone_std"), ]

n_positive <- sum(data[["decision"]] == "y")
n_negative <- sum(data[["decision"]] == "n")

dat <- data.frame(polyarea = data[, "polyarea"], 
                  decision = as.numeric(factor(data$decision, 
                                         levels = c("n", "y"), 
                                         label = c(0, 1))) - 1)

# Select randomly observations from 70% of the data for training.
# n_train is the number of observations used for training.

n_train <- round(nrow(data) * 0.7)

# index_test is the index of observations to be selected for the training
index_test <- sample(1L:nrow(dat), size = n_train)

# index_test is the index of observations to be selected for the testing
index_training <- which(!(1L:nrow(dat) %in% index_test))

# train_data contains the data used for training

train_data <- dat[index_test, ]

# test_data contains the data used for training

test_data <- dat[index_training, ]

# Fit the binomial logistic regression model

model_glm <- glm(decision ~ polyarea, family=binomial(link='logit'), 
                 data = train_data)

predictions <- ifelse(predict(model_glm, 
                                 newdata = test_data, type = 'response') > 0.5,
                         1, 0)

res_performeR <- performeR(predictions, test_data[["decision"]])[, c(1:10, 12)]
```

The ``summary()`` function returns the results of the model fitting. This can be analysed and interpreted. 


```{r, echo=TRUE}
summary(model_glm)
```

Based on the results it can be concluded that the parameters ` (Intercept)` and `polyarea` are statistically significant (*P* < 2e-16). This indicates a strong association between `polyarea` and the probability that an amplification curve is positive.

In order to apply the model to a new dataset, further steps are necessary. ``predict()`` is a generic function for prediction from the results of a model fitting function. All previously split test data is passed to the function argument `newdata`. By setting the `type = 'response'` parameter, the ``predict()`` function returns probabilities in the form of $P(y=1|X)$. In the case in hand, it was decided that a decision limit of 0.5 is to be applied. If $P(y=1|X) < 0.5$ then $y = 0$ (amplification curve negative), otherwise $y = 1$ (amplification curves positive).


```{r plot_Logistic_Regression, echo=TRUE, fig.cap=plot_Logistic_Regression, fig.scap=plot_Logistic_Regression_short, fig.height=3.7}
library(PCRedux)

par(mfrow = c(1,2))

# Plot train_data (grey points) and the predicted model (blue)

plot(train_data$polyarea, train_data$decision, pch = 19, 
     xlab = "polyarea", ylab = "Probability", 
     col = adjustcolor("grey", alpha.f = 0.9), cex = 1.5)
mtext("A", cex = 1.2, side = 3, adj = 0, font = 2, las = 0)
abline(h = 0.5, col = "grey")

curve(predict(model_glm, data.frame(polyarea = x), type = "resp"), 
      add = TRUE, col = "blue")

# Plot test_data (red)

points(test_data$polyarea, test_data$decision, pch = 19,
       col = adjustcolor("red", alpha.f = 0.3))
legend("right", paste("Positive: ", n_positive, 
                      "\nNegative: ", n_negative), bty = "n")


# Plot the sensitivity, specificity and other measures to describe the prediction.

position_bp <- barplot(as.matrix(res_performeR), yaxt = "n", 
                       ylab = "Probability", main = "", las = 2, 
                       col = adjustcolor("grey", alpha.f = 0.5), 
                       border = "white")

par(srt = 90)
text(position_bp, rep(0.8, length(res_performeR)), 
     paste(signif(res_performeR, 2)*100, "%"), cex = 0.6)
axis(2, at = c(0, 1), labels = c("0", "1"), las = 2)
abline(h = 0.85, col = "grey")

mtext("B", cex = 1.2, side = 3, adj = 0, font = 2, las = 0)
```

The sensitivity, specificity and further parameters for estimating the prediction were calculated using the ``performeR()`` function (\autoref{section_performeR}). The results indicate that the sensitivity and specificity for the test dataset provides a good result. However, the results in this case depend heavily on the computer-aided random sampling of the training data and the total size of the dataset. Over-fitting and under-fitting and other problems need to be addressed [@walsh_correct_2015].

To proof the results, further methods such as Likelihood Ratio Test, McFadden’s $R^{2}$, k-fold cross-validation, Receiver Operating Characteristic (ROC) analysis and model interpretation should be used [@arlot_survey_2010, @mcfadden_conditional_1974, @sing_rocr:_2005].


```{r, eval=TRUE, echo=FALSE}
data(testdat, package = "qpcR")
x <- testdat[, 1]
y.pos <- testdat[, 2]
y.neg <- testdat[, 4]
nh <- trunc(length(x) * 0.20)
nt <- trunc(length(x) * 0.15)

y.pos.head <- head(y.pos, n = nh)
y.neg.head <- head(y.neg, n = nh)
y.pos.tail <- tail(y.pos, n = nt)
y.neg.tail <- tail(y.neg, n = nt)

lb.pos <- median(y.pos.head) + 2 * mad(y.pos.head)
ub.pos <- median(y.pos.tail) - 2 * mad(y.pos.tail)

lb.neg <- median(y.neg.head) + 2 * mad(y.neg.head)
ub.neg <- median(y.neg.tail) - 2 * mad(y.neg.tail)

res.shapiro.pos <- shapiro.test(y.pos)
res.shapiro.neg <- shapiro.test(y.neg)

res.wt.pos <- wilcox.test(head(y.pos, n = nh), tail(y.pos, n = nt), alternative = "less")
res.wt.neg <- wilcox.test(head(y.neg, n = nh), tail(y.neg, n = nt), alternative = "less")

###
RGt <- function(y) {
  ws <- ceiling((15 * length(y)) / 100)
  if (ws < 5) {
    ws <- 5
  }
  if (ws > 15) {
    ws <- 15
  }
  y.tmp <- na.omit(y[-c(1:5)])
  x <- 1:length(y.tmp)
  suppressWarnings(
    res.reg <- sapply(1L:(length(y.tmp)), function(i) {
      round(summary(lm(y.tmp[i:c(i + ws)] ~ x[i:c(i + ws)]))[["r.squared"]], 4)
    })
  )

  # Binarize R^2 values. Everything larger than 0.8 is positive
  res.LRt <- res.reg
  # Define the limits for the R^2 test
  res.LRt[res.LRt < 0.8] <- 0
  res.LRt[res.LRt >= 0.8] <- 1
  # Seek for a sequence of at least six positive values (R^2 >= 0.8)
  # The first five measuring point of the amplification curve are skipped
  # because most technologies and probe technologies tend to overshot
  # in the start (background) region.
  res.out <- sapply(5L:(length(res.LRt) - 6), function(i) {
    ifelse(sum(res.LRt[i:(i + 4)]) == 5, TRUE, FALSE)
  })
  out <- cbind(1L:(length(y.tmp)), res.reg)
  # res.out
}
```


```{r statistical_methods_amptester, echo = FALSE, fig.cap=statistical_methods_amptester, fig.scap=statistical_methods_amptester_short, fig.height=6.8}

layout(matrix(c(1, 1, 2, 2, 1, 1, 2, 2, 
                3, 3, 4, 4, 5, 5, 6, 6), 4, 4, byrow = TRUE))

y_lim <- range(testdat[, 2]) * c(1, 1.3)

plot(
    x, y.pos, ylim = y_lim, xlab = "Cycles",
     ylab = "RFU", main = "", type = "b", pch = 19
)
mtext("A    Positive amplification", side = 3, adj = 0, font = 2)
abline(v = c(nh, length(x) - nt), lty = 3)

abline(h = lb.pos, lty = 2, col = "red")
text(35, 1.5, "Noise\nmedian + 2 * MAD", col = "red", cex = 1)

abline(h = ub.pos, lty = 2, col = "green")
text(35, 10, "Signal\nmedian - 2 * MAD", col = "green", cex = 1)

arrows(4.5, 12.5, 42.5, 12.5, length = 0.1, angle = 90, code = 3)
text(25, 14.5, paste(
    "W=", signif(res.wt.pos$statistic, 3), "\np-value = ",
                     signif(res.wt.pos$p.value, 3)
))
# text(25, 5, paste("Fold change: \n", round(ub.pos / lb.pos, 2)))


y_lim <- range(testdat[, 4]) * c(1.5, 2.5)

plot(
    x, y.neg, ylim = y_lim, xlab = "Cycles",
     ylab = "RFU", main = "", type = "b", pch = 19
)
mtext("B    Negative amplification", side = 3, adj = 0, font = 2)
abline(v = c(nh, length(x) - nt), lty = 3)

abline(h = lb.neg, lty = 2, col = "red")
text(10, 0.04, "Noise\nmedian + 2 * MAD", col = "red", cex = 1)

abline(h = ub.neg, lty = 2, col = "green")
text(40, -0.02, "Signal\nmedian - 2 * MAD", col = "green", cex = 1)

arrows(4.5, 0.04, 42.5, 0.04, length = 0.1, angle = 90, code = 3)
text(10, 0.06, paste(
    "W = ", res.wt.neg$statistic, "\np-value = ",
    signif(res.wt.neg$p.value, 3)
))

qqnorm(y.pos, pch = 19, main = "")
legend("bottomright", paste(
    "SHt, W = ", signif(res.shapiro.pos$statistic, 3),
                            "\np-value = ", signif(res.shapiro.pos$p.value, 3)
), bty = "n")
mtext("C", side = 3, adj = 0, font = 2)
qqline(y.pos, col = "orange", lwd = 2)


qqnorm(y.neg, pch = 19, main = "")
legend(
    "bottomright", paste(
        "SHt, W = ", signif(res.shapiro.neg$statistic, 3),
                         "\np-value = ", signif(res.shapiro.neg$p.value, 3)
    ),
    bty = "n"
)
mtext("D", side = 3, adj = 0, font = 2)
qqline(y.neg, col = "orange", lwd = 2)


plot(
    RGt(y.pos), xlab = "Cycles", ylab = expression(R ^ 2), main = "", pch = 19,
     type = "b"
)
mtext("E    LRt", side = 3, adj = 0, font = 2)
abline(h = 0.8, col = "black", lty = 2)


plot(
  RGt(y.neg), xlab = "Cycles", ylab = expression(R ^ 2), main = "",
  pch = 19, type = "b"
)
mtext("F    LRt", side = 3, adj = 0, font = 2)
abline(h = 0.8, col = "black", lty = 2)
```

\newpage


#### ``earlyreg()`` - A Function to Calculate the Slope and Intercept in the Ground Phase of an Amplification Curve \label{section_earlyreg}

The signal height and the slope in the first cylces (1 - 10) of amplification curves are potentially useful because some qPCR systems calibrate themselves by fluorescence intensity of the first cycles. This is noticeable as strong signal changes which appear spontaneously between the first and second cycle. For another, the signal level can be used to determine which background signal is present and whether the ground phase already has a slope. Moreover, characteristics of the detection probe system are noticeable (see \autoref{section_DataAnalysis}). From the slope it could be deduced whether amplification has already started (see \autoref{section_DataAnalysis}). 

In addition, the ``earlyreg()`` function was developed. This function uses an ordinary least squares linear regression within a limited number of cycles. As ROI, the first 10 cycles were defined. This restriction is based on empirical data suggesting that during the first ten cycles only a significant increase in signal strength can be measured within few qPCRs. However, ``earlyreg()`` does not ignore the first cycle, as many thermo-cyclers use this cycle for sensor calibration. Extreme values are therefore included. As standard, the next nine amplitude values are used for the linear regression. The number of cycles can also be adjusted via the parameter `range`. Since all amplification curves are normalized to the 99%-percentile, there is also a comparability between the background signals and the slopes. The output of the the ``earlyreg()`` function is:

- `slope_bg`, which is the slope of the ordinary least squares linear regression model,
- `intercept_bg`, which is the intercept of the linear model and
- `sigma_bg`, which is the square root of the estimated variance of the random error.

[//]: # ( $\sigma^{2} = \frac{1}{n-p} \sum (w_{i} R_{i}^2)$, where $R_{i}$ is the i$_{th}$ residual, residuals$_{i}$)

The following example illustrates a possible use of the function ``earlyreg()``. For that purpose amplification curves from the `C127EGHP` dataset were analysed (\autoref{earlyreg_slopes}A). In figure \autoref{earlyreg_slopes}A the amplification curves for all cycles are shown. Next, the ``earlyreg()`` function was used to determine the `slope_bg`, `intercept_bg` and `sigma_bg` in the range of the first ten PCR cycles. The results were used in a cluster analysis using k-means clustering. Therefore, the slope seems to be an indicator of differences between the amplification curves. The \autoref{earlyreg_slopes}B shows the first 8 cycles colored according to their cluster. 
After the cluster analysis this could also be observed (\autoref{earlyreg_slopes}D-F). Hence, it can be postulated that the slope in the background phase is useful for the amplification curve classification.


```{r earlyreg_slopes, echo = TRUE, fig.cap =earlyreg_slopes, fig.scap=earlyreg_slopes_short, fig.height=5}
library(PCRedux)

# Function for the Box-Cox transformation of data

box_cox <- function(x, lambda = 1, offset = 0) {
    if (lambda == 0) {
        log(x + offset)
    } else 
    {
        ((x + offset)^lambda - 1)/lambda
    }
}

# Load the C127EGHP data set

data <- C127EGHP[, -1]

# Normalize each amplification curve to their 0.99 percentile and use the
# earlyreg function to determine the slope and intercept of the first
# cycles 'user_range'

user_range <- 8

res_earlyreg <- do.call(rbind, lapply(2L:ncol(data), function(i) {
  earlyreg(x = data[, 1], y = data[, i], range = user_range, normalize = TRUE)
})) %>% box_cox(.)

# Label the observation with their names
rownames(res_earlyreg) <- substr(colnames(data)[-1], 1, 10)

# Show the first five lines of the res_earlyreg data matrix

head(res_earlyreg)

# Perform k-means clustering on the res_earlyreg data matrix
cl <- kmeans(res_earlyreg, centers = 2)

# Plot the results
# Use x_roi (cycles) and rfu_range (RFU values) to limit the 
# range for the detailed plot of first cycles. 

x_roi <- 1:(user_range + 1)
rfu_range <- range(data[x_roi, -1])

layout(matrix(c(1, 2, 1, 2, 3, 3), 3, 2, byrow = TRUE))

# Plot of raw amplification curves

matplot(
    data[, 1], data[, -1], ylim = range(data[, -1]), pch = 19, lty = 1, 
        type = "l", xlab = "Cycles", ylab = "RFU", main = "", col = "grey"
)
mtext("A", cex = 1.2, side = 3, adj = 0, font = 2)
abline(v = c(1,user_range))
text(3, range(data[, -1])[2], "ROI")

# Detailed plot of the first cycles and the clusters according
# to the k-means clustering
# Define some user colors (blue: EvaGreen, orange: Hydrolysis probes)
colors <- c(
    adjustcolor("blue", alpha.f = 0.5),
            adjustcolor("orange", alpha.f = 0.8)
)

matplot(NA, NA, xlim = range(x_roi), ylim = rfu_range, xlab = "Cycles", 
        ylab = "RFU", main = "")

for(i in 1L:length(unique(cl$cluster))) {
cl_id <- which(cl[["cluster"]] == i) + 1
par(new=TRUE)
matplot(data[x_roi, 1], xlab = "", ylab = "", xaxt = "n", yaxt = "n", 
        data[x_roi, cl_id], ylim = rfu_range, pch = 19, lty = 1, type = "l", 
        col = colors[i])
}


mtext("B", cex = 1.2, side = 3, adj = 0, font = 2)
abline(v = c(1,user_range))
text(3, rfu_range[2], "ROI")
legend("bottomleft", c("Cluster 1", "Cluster 2"), pch = 15, cex = 1.2, 
       col = colors, bty = "n")

# Overview of clusters and corresponding detection chemistry

eghp <- rep(0.7, length(cl$cluster))
names(eghp) <- names(cl$cluster)

barplot(eghp, las = 2, col = colors[cl[["cluster"]]], 
        border = "white", xlab = "", ylab = "", 
        yaxt = "n", ylim = c(0,2.2), cex.axis = 0.7)
mtext("C", cex = 1.2, side = 3, adj = 0, font = 2)
legend("topleft", c("Cluster 1", "Cluster 2"), pch = 15, col = colors, 
       bty = "n", box.col =  "white", cex = 0.9)
arrows(0.5, 2, 38, 2, angle = 90, code = 3)
arrows(39, 2, 76.5, 2, angle = 90, code = 3)
text(c(18.75, 57.5), c(1.6, 1.6), c("EvaGreen", "Hydrolysis probes"))
```


#### ``head2tailratio()`` - A Function to Calculate the Ratio of the Head and the Tail of a Quantitative PCR Amplification Curve\label{section_head2tailratio}

The ratios from the ground and plateau phase can be used to search for patterns in amplification curves. Positive amplification curves have different slopes and intercepts at the start of the amplification curve (head, background region) and the end of the amplification curve (tail, plateau region). Therefore, these regions are potentially useful to extract a feature for an amplification curve classification. Negative amplification curves (no slope) are assumed to have a ratio of about 1. In contrast, positive amplification curves should have a ratio of less than 1.

The ``head2tailratio()`` function calculates the ratio of the head and the tail of a quantitative PCR amplification curve. As ROI, the areas in the ground phase (head) and plateau phases (tail) are used (\autoref{amplification_curve_ROI}A). For the calculation, the median from the first six data points of the amplification curve and the median from the last six data points are used. The determination of six data points in both regions was made on the basis of \emph{empirical experience}. As a rule, no increase in amplification signals can be measured in the first six cycles and in the last six cycles, the amplification curve is usually about to transition into the plateau. This assumption is sometimes violated and might lead to false estimates.


```{r, echo = TRUE}
library(PCRedux)

# Load the RAS002 dataset and assign it to the object data

data <- RAS002
data_decisions <- RAS002_decisions

# Calculate the head2tailratio of all amplification curves

res_head2tailratio <- lapply(2L:ncol(data), function(i) {
  head2tailratio(
    y = data[, i], normalize = TRUE, slope_normalizer = TRUE,
    verbose = TRUE
  )
})

# Fetch all values of the head2tailratio analysis for a later comparison
# by a boxplot.

res <- sapply(1L:length(res_head2tailratio), function(i)
  res_head2tailratio[[i]]$head_tail_ratio)

data_normalized <- cbind(
  data[, 1],
  sapply(2L:ncol(data), function(i) {
    data[, i] / quantile(data[, i], 0.99)
  })
)

# Assign colors to the classes (n: black, y: green).
colors <- as.character(factor(
  data_decisions, levels = c("y", "n"),
  labels = c(
      adjustcolor("green", alpha.f = 0.25), adjustcolor("black", alpha.f = 0.25))
))

res_wilcox.test <- stats::wilcox.test(res ~ data_decisions)
```

The amplification curves in \autoref{figure_head2tailratio} show a signal increase within the first three cycles and the amplification curves in \autoref{amplification_curve_ROI}C have a negative slope in the tail. The median is used to minimize the influence of outliers.


```{r figure_head2tailratio, echo=TRUE, fig.cap=figure_head2tailratio, fig.scap=figure_head2tailratio_short, fig.higth=3}
# Plot the results of the analysis
#
# Position and plot parameters
h <- max(na.omit(res))
h_text <- rep(h * 0.976, 2)

layout(matrix(c(1,1,2), 1, 3, byrow = TRUE))

matplot(
  data_normalized[, 1], data_normalized[, -1],
  xlab = "Cycles", ylab = "normalized RFU", main = "",
  type = "l", lty = 1, lwd = 2, col = colors
)
for (i in 1L:(ncol(data_normalized) - 1)) {
  points(
    res_head2tailratio[[i]]$x_roi, res_head2tailratio[[i]]$y_roi,
    col = colors[i], pch = 19, cex = 1.5
  )
  abline(res_head2tailratio[[i]]$fit, col = colors[i], lwd = 2)
}
mtext("A", cex = 1.2, side = 3, adj = 0, font = 2)

# Boxplot of the head2tail ratios of the positive and negative
# amplification curves.

boxplot(res ~ data_decisions, col = unique(colors), ylab = "Head to Tail Ratio")

lines(c(1, 2), rep(h * 0.945, 2))
text(1.5, h_text, paste0("P = ", signif(res_wilcox.test[["p.value"]])), 
     cex = 1)

mtext("B", cex = 1.2, side = 3, adj = 0, font = 2)
```

In \autoref{section_DataAnalysis} and in \autoref{htPCR_nap} it was shown that negative amplification curves may have a slope with the positive or negative sign. The is no consent in the literate and among peers how to deal with this during the processing. One solution is to include the slope as factor in the ratio calculation. The ``head2tailratio()`` function uses a linear model that calculates the slope between the ground and plateau phases. If the slope of the model is significant, then the ratio from the head and tail is normalized to this slope. This requires setting the `slope_normalizer` parameter in the ``head2tailratio()`` function. By default, this parameter is not set.


#### ``hookreg()`` and ``hookregNL()`` - Functions to Detect Hook Effect-like Curvatures\label{section_hookreg}

``hookreg()`` and ``hookregNL()`` are functions to detect amplification curves bearing a hook effect [@barratt_improving_2002] or negative slope at the end of the amplification curve. Both functions calculate the slope and intercept of an amplification curve data. The idea is that a strong negative slope at the end of an amplification curve is indicative for a hook effect. ``hookreg()`` and ``hookregNL()`` are part of a peer-reviewed publication. For this reason, the functions will not be discussed here.

#### ``mblrr()`` - A Function Perform the Quantile-filter Based Local Robust Regression \label{section_mblrr}


``mblrr()`` is a function to perform the \textbf{m}edian \textbf{b}ased \textbf{l}ocal \textbf{r}obust \textbf{r}egression (mblrr) from a quantitative PCR experiment. In detail, this function attempts to break the amplification curve in two ROIs (head (~background) and tail (~plateau)). As opposed to the ``earlyreg()`` function, the ``mblrr()`` function does not use a fixed interval. Instead, the ``mblrr()`` function dynamically determines cut points for each amplification curve. For the ``mblrr()`` function was defined:

- The 25% quantile is the value for which 25% of all values are smaller than this value. 
- The 75% quantile is the value for which 75% of all values are greater than this value. 

Subsequent, a robust linear regression analysis (``lmrob()``) is preformed individually on both regions of the amplification curve. The rationale behind this analysis is that the slope and intercept of an amplification curve differ in the background and plateau region. This is also shown by the simulations in \autoref{figure_sigmoid_curve_models}A-C. In the example shown below, the observations "P01.W19", "P06.W35", "P33.W66", "P65.W90", "P71.W23" and "P87.W01" were arbitrarily selected for demonstration purposes \autoref{plot_mblrr}. Another example is shown in \autoref{HCU32}A. Those amplification curves have a slight negative trend in the base-line region and a positive trend in the plateau region.

The correlation coefficient\footnote{Product moment correlation coefficient (Pearson)} is a measure to quantify the dependence on variables (e.&nbsp;g., number of cycles, signal height). The correlation coefficient is always between -1 and 1, with a value close to -1 describing a strong-negative dependency and close to 1 describing a strong-positive dependency; if the value is 0, there is no dependency between the variables. The most frequently used correlation coefficient to describe a linear dependency is the Pearson correlation coefficient *r*.

The correlation coefficient can also be used as a feature. Similar data structures have similar correlation coefficients. Correlation coefficients are between -1 and +1, with -1 being a strong negative correlation and 1 a strong positive correlation. The values of -1 and 1 have a perfect correlation. If the value is 0, there is no correlation between the two variables. However, variables that are not strongly correlated can also be important for modeling.

```{r plot_mblrr, echo=TRUE, eval = TRUE, fig.cap=plot_mblrr, fig.scap=plot_mblrr_short, fig.height=6}
library(PCRedux)

# Select four amplification curves from the RAS002 dataset

data <- RAS002[, c(1, 2, 3, 4, 5)]

colors <- c(
            adjustcolor("blue", alpha.f = 0.5),
            adjustcolor("orange", alpha.f = 0.8)
           )

par(mfrow = c(2, 2))

for (i in 2L:ncol(data)) {
  x <- data[, 1]
  y_tmp <- data[, i] / quantile(data[, i], 0.99)
  res_q25 <- y_tmp < quantile(y_tmp, 0.25)
  res_q75 <- y_tmp > quantile(y_tmp, 0.75)
  res_q25_lm <- try(
    suppressWarnings(lmrob(y_tmp[res_q25] ~ x[res_q25])),
    silent = TRUE
  )
  res_q75_lm <- try(
    suppressWarnings(lmrob(y_tmp[res_q75] ~ x[res_q75])),
    silent = TRUE
  )

  plot(x, y_tmp, xlab = "Cycles", ylab = "RFU (normalized)",
    main = "", type = "b", pch = 19)
  
  mtext(paste0(LETTERS[i], "   ", colnames(data)[i]), cex = 1, side = 3, 
        adj = 0, font = 2)
  abline(res_q25_lm, col = colors[1])
  points(x[res_q25], y_tmp[res_q25], cex = 2.5, col = colors[1])
  abline(res_q75_lm, col = colors[2])
  points(x[res_q75], y_tmp[res_q75], cex = 2.5, col = colors[2])
}
```

Finally, the results of the analysis were printed in a tabular format.

```{r, eval=TRUE, echo=TRUE, results='asis'}
# Load the xtable library for an appealing table output
library(xtable)

# Analyze the data via the mblrr() function

res_mblrr <- do.call(cbind, lapply(2L:ncol(data), function(i) {
  suppressMessages(mblrr(
    x = data[, 1], y = data[, i],
    normalize = TRUE
  )) %>% data.frame()
}))
colnames(res_mblrr) <- colnames(data)[-1]

# Transform the data for a tabular output and assign the results to the object
# output_res_mblrr.

output_res_mblrr <- res_mblrr %>% t()

# The output variable names of the mblrr() function are rather long. For better
# readability the variable names were changed to "nBG" (intercept of head region),
# "mBG" (slope of head region), "rBG" (Pearson correlation of head region),
# "nTP" (intercept of tail region), "mTP" (slope of tail region), "rBG" (Pearson
# correlation of tail region)

colnames(output_res_mblrr) <- c(
  "nBG", "mBG", "rBG",
  "nTP", "mTP", "rTP"
)

print(xtable(
  output_res_mblrr, caption = "mblrr() text intro. nBG, intercept of 
             head region; mBG, slope of head region; rBG, Pearson 
             correlation of head region; nTP, intercept of tail region; mTP, 
             slope of tail region; rBG, Pearson correlation of tail region",
  label = "tablemblrrintroduction"
), comment = FALSE, caption.placement = "top")
```

In another example, the results from the ``mblrr()`` function were combined with the classifications (positive, negative) by a human to apply them in an analysis with Fast and Frugal Trees (FFTrees). FFTrees belong to class of simple decision rules. DT's are a classic approach to machine learning [@quinlan_induction_1986]. Here relatively simple algorithms and simple tree structures are used to create a model. A general introduction to decision trees is given in [@quinlan_induction_1986, @luan_signal-detection_2011]. In many situations, FFTrees make fast decisions based on a few features (N = 1 - 5). In this example six features were used for the analysis.

The \texttt{FFTrees} package [@FFTrees_package] provides an implementation for the \texttt{R} statistical computing language. All that is needed for the present example are:

* the data assessed by the ``mblrr()`` function,
* the classification of the amplification curve data by a human,
* and a standard formula, which looks like $outcome \leftarrow var1 + var2 + \ldots$ along with the data arguments. The function ``FFTrees()`` returns a fast and frugal tree object. This rich object contains the underlying trees and many classification statistics (similar to \autoref{section_performeR}). In the following example, the `RAS002` dataset from the \texttt{qpcR} package was used.

```{r, eval=TRUE, echo=TRUE}
# Load the xtable library for an appealing table output
library(FFTrees)
library(PCRedux)

# The RAS002 amplification curves were analyzed with the mblrr() function 
# to save computing time and the.results of this analysis are stored in the 
# `data_sample` dataset.

data <- data_sample[data_sample$dataset == "RAS002", c("mblrr_intercept_bg", 
                                                       "mblrr_slope_bg", 
                                                       "mblrr_cor_bg", 
                                                       "mblrr_intercept_pt", 
                                                       "mblrr_slope_pt", 
                                                       "mblrr_cor_pt")]

# The output variable names of the mblrr() function are rather long. For better
# readability the variable names were changed to "nBG" (intercept of head
# region), "mBG" (slope of head region), "rBG" (Pearson correlation of head
# region), "nTP" (intercept of tail region), "mTP" (slope of tail region),
# "rBG" (Pearson correlation of tail region).


res_mblrr <- data.frame(
    class = as.numeric(as.character(factor(RAS002_decisions, 
                                           levels = c("y", "n"), 
                                           label = c(1, 0)))),
  data
)

colnames(res_mblrr) <- c("class", "nBG", "mBG", "rBG", "nTP", "mTP", "rTP")

res_mblrr.fft <- suppressMessages(
            FFTrees(formula = class ~., data = res_mblrr)
            )
``` 

\autoref{plot_FFTrees} shows the Fast and Frugal Trees by using the features nBG (intercept of head region), mBG (slope of head region), rBG (Pearson correlation of head region), nTP (intercept of tail region), mTP (slope of tail region), and rBG (Pearson correlation of tail region). 


```{r plot_FFTrees, echo=FALSE, fig.cap=plot_FFTrees, fig.scap=plot_FFTrees_short, fig.height=11, fig.width=11}
plot(res_mblrr.fft, decision.lables = c("Positive", "Negative"))
```

\texttt{R} offers several packages like \texttt{party} [@hothorn_unbiased_2006], \texttt{rpart} [@rpart_2017] and \texttt{Rattle} [@williams_rattle:_2009] for creating decision trees. 

#### ``autocorrelation_test()`` - A Function to Detect Positive Amplification Curves \label{section_autocorrelation_test}

Autocorrelation analysis is a technique that is used in the field of time series analysis. It can be used to reveal regularly occurring patterns in one-dimensional data [@spiess_system-specific_2016]. The autocorrelation measures the correlation of a signal $f(t)$ with itself shifted by some time delay $f(t - \tau)$.

The ``autocorrelation_test()`` function coercers the amplification curve data to an object of the class "zoo" (\texttt{zoo} package) as indexed totally ordered observations. Next follows the computation of a lagged version of the amplification curve data. The shifting of the amplification curve data is based on the number of observations (number of cycles 'c') with the following $\tau$.

Number of Cycles (c) | $\tau$
---------------------|----------------
$c \leq 35$          | 8
$35 > c \leq 40$     | 10
$40 < c \leq 45$     | 12
$c > 45$             | 14


This is followed by a significance test for correlation between paired observations (amplification curve data & lagged amplification curve data). The hypothesis is that the paired observation of positive amplification curves has a significant correlation (`stats::cor.test`, significance level is 0.01) in contrast to negative amplification curves (noise). The application of the ``autocorrelation_test()`` function is shown in the following example.

In addition, the decisions file `decision_res_RAS002.csv` from the user was analyzed for the most frequent decision (modus) using the ``decision_modus()`` function (\autoref{section_decision_modus}).


```{r autocorrelation, echo=TRUE, fig.cap=autocorrelation, fig.scap=autocorrelation_short, fig.height=5.5, fig.width=11}
# Test for autocorrelation in amplification curve data
# Load the libraries magrittr for pipes and the amplification curve the data
# The amplification curve data from the `htPCR` dataset was used.
# The data.table package was used for fast import of the csv data
library(magrittr)
library(PCRedux)

# Load the decision_res_htPCR.csv dataset from a csv file.
filename <- system.file("decision_res_htPCR.csv", package = "PCRedux")
decision_res_htPCR <- data.table::fread(filename, data.table = FALSE)

# Select only amplification curves (obs) were all 
# classifications were in concordance ("conformity == TRUE" ).
# This subset of the htPCR dataset contains the classes 
# n: negative
# a: ambiguous
# p: positive

obs_number <- which(decision_res_htPCR[["conformity"]] == TRUE)
decs <- decision_res_htPCR[obs_number, "test.result.1"] %>% unlist(.)

# Give tabular output of classes
table(decs)

# Load only the amplification curves from the htPCR data set that were 
# uniquely assigned to one class (e.g., eight out of eight positive).

data <- qpcR::htPCR[, c(1, obs_number + 1)]

# Since the number of ambiguous is low (n = 2), they were re-assigned to the 
# class negative

decs[decs == "a"] <- "n"

table(decs)

# Assign colors to the classes (n: black, y: green).

colors <- factor(decs, levels = c("y", "n"), label = c("black", "green"))


# Test for autocorrelation in the RAS002 dataset

res_ac <- sapply(2:ncol(data), function(i) {
    autocorrelation_test(data[, i], ns_2_numeric = TRUE)
})

# Curves classified by a human after analysis of the overview. 1 = positive,
# 0 = negative

# # dec <- data.table::fread(system.file(
# #   "decision_res_RAS002.csv",
# #   package = "PCRedux"
# # ))

head(dec)


# decs <- sapply(1L:nrow(dec), function(i) {
#   res <- decision_modus(dec[i, 2L:(ncol(dec) - 1)])
#   if (length(res) > 1) res[[1]] <- "n"
#   res[[1]]
# }) %>% unlist()


# Plot curve data as overview
# Names of the observations

layout(matrix(c(1, 2, 3, 1, 4, 4), 2, 3, byrow = TRUE))
matplot(
    data[, 1], data[, -1], xlab = "Cycles", ylab = "RFU",
    main = "", type = "l", lty = 1,
    col = colors, lwd = 2
)
legend("topleft", c("positive", "negative"), pch = 19, col = c(1, 2), bty = "n")
mtext("A    RAS002 dataset", cex = 1.2, side = 3, adj = 0, font = 2)


# Convert the n.s. (not significant) in 0 and others to 1.
# Combine the results of the aromatic autocorrelation_test as variable "ac",
# the human classified values as variable "hc" in a new data frame (res_ac_hc).

cutoff <- 0.85

res_ac_hc <- data.frame(
    ac = ifelse(res_ac > cutoff, 1, 0),
                        hc = ifelse(decs == "y", 0, 1)
) %>% as.matrix()
res_performeR <- performeR(s = res_ac_hc[, "ac"], r = res_ac_hc[, "hc"])


plot(density(res_ac), xlab = "Autocorrelation", ylab = "Density", main = "")
rug(res_ac)

abline(v = cutoff)
mtext("B", cex = 1.2, side = 3, adj = 0, font = 2, las = 0)


cdplot(
    as.factor(decs) ~ res_ac, xlab = "Autocorrelation",
       ylab = "Class decision"
)
mtext("C", cex = 1.2, side = 3, adj = 0, font = 2, las = 0)

barplot(
    as.matrix(res_performeR[, c(1:10, 12)]), yaxt = "n", ylab = "",
        main = "Performance of autocorrelation_test",
        col = adjustcolor("grey", alpha.f = 0.5), border = "white"
)
text(position_bp, rep(0.8, length(res_performeR[, c(1:10, 12)])), 
     paste(signif(res_performeR[, c(1:10, 12)], 2)*100, "%"))
axis(2, at = c(0, 1), labels = c("0", "1"), las = 2)
mtext("D", cex = 1.2, side = 3, adj = 0, font = 2, las = 0)
```

As shown in this example, the ``autocorrelation_test()`` function is able to distinguish between positive and negative amplification curves. Negative amplification curve were in all cases non-significant. In contrast, the coefficients of correlation for positive amplification curves ranged between `r signif(range(as.numeric(res_ac), na.rm = TRUE), 3)[1]` and `r signif(range(as.numeric(res_ac), na.rm = TRUE), 3)[2]` at a significance level of 0.01 and a lag of 3.

\newpage

```{r plot_peaks_ratio, eval = FALSE, echo=FALSE, fig.cap=plot_peaks_ratio, fig.scap=plot_peaks_ratio_short, fig.height=8}

par(mfrow = c(2,1))

for(i in 1:2){
dat_smoothed <- chipPCR::smoother(RAS002[, 1], RAS002[, i+1])
res_diffQ <- suppressMessages(MBmca::diffQ(cbind(RAS002[-c(1,2), 1], dat_smoothed[-c(1,2)]), verbose = TRUE)$xy)
res_mcaPeaks <- MBmca::mcaPeaks(res_diffQ[, 1], res_diffQ[, 2])

range_p.max <- range(res_mcaPeaks$p.max[, 2])
diff_range_p.max <- diff(range_p.max)


range_p.min <- range(res_mcaPeaks$p.min[, 2])
diff_range_p.min <- diff(range_p.min)


peaks_ratio <- diff_range_p.max / diff_range_p.min


plot(res_diffQ, xlab = "Cycles", ylab = "dRFU/dCycle", type = "b", 
    ylim = c(diff_range_p.min, diff_range_p.max))

points(res_mcaPeaks$p.max, col = "red", pch = 19)
points(res_mcaPeaks$p.min, col = "cyan", pch = 19)

arrows(mean(res_mcaPeaks$p.max[, 1]), 
       range(res_mcaPeaks$p.max[, 2])[[1]], 
       mean(res_mcaPeaks$p.max[, 1]), 
       range(res_mcaPeaks$p.max[, 2])[[2]], col = "red", angle = 90, code = 3)

points(mean(res_mcaPeaks$p.max[, 1]), diff_range_p.max, col = "red", pch = 13, cex = 2)

arrows(mean(res_mcaPeaks$p.min[, 1]), 
       range(res_mcaPeaks$p.min[, 2])[[1]], 
       mean(res_mcaPeaks$p.min[, 1]), 
       range(res_mcaPeaks$p.min[, 2])[[2]], col = "cyan", angle = 90, code = 3)
       
points(mean(res_mcaPeaks$p.min[, 1]), diff_range_p.min, col = "cyan", pch = 13, cex = 2)

mtext(LETTERS[i], cex = 1.2, side = 3, adj = 0, font = 2)


legend("bottomleft", c("Local maxima", "Mean local maxima", 
"Local minima", "Mean local minima", paste("peaks_ratio", signif(diff_range_p.max/diff_range_p.min, 2))), 
                       col = c("red", "cyan", "red", "cyan", "black"), pch = c(19, 13, 19, 13, 1), 
                       bty = "n")
}
```


```{r plot_cp_area, echo=FALSE, fig.cap=plot_cp_area, fig.scap=plot_cp_area_short, fig.height=11, fig.width=11}
library(PCRedux)

data <- data_sample[data_sample$dataset %in% c("stepone_std", "RAS002", "RAS003", "lc96_bACTXY", "C126EG595", "dil4reps94"), ]

feature <- c(
    "polyarea", "peaks_ratio", "changepoint_e.agglo", "changepoint_bcp", 
  "amptester_polygon", "amptester_slope.ratio")

par(mfrow = c(2,3))

x <- data$decision

for(i in 1L:length(feature)) {
    y <- data[, colnames(data) == feature[i]]
    res <- stats::wilcox.test(y ~ x)
    h <- max(na.omit(y))
    l <- min(na.omit(y))
    h_text <- rep(h * 0.976, 2)
    
    par(bg=NA)
    stripchart(y ~ x, vertical = TRUE, ylab = feature[i],
               method = "jitter", pch = 19, cex = 1, 
               col = adjustcolor("darkgrey", alpha.f = 0.65), 
               ylim = c(l * 0.95, h * 1.05))
    
    boxplot(y ~ x, outline = FALSE, add = TRUE, boxwex = 0.35)

    legend("topleft", paste0("P = ", signif(res[["p.value"]])), 
           cex = 1, bty = "n")
    
    mtext(paste0(LETTERS[i], "   ", feature[i]), cex = 1.2, side = 3, 
          adj = 0, font = 2, col = ifelse(signif(res[["p.value"]]) < 0.05, 
                                          "black", "red"))
}
```


#### Change point analysis\label{section_change_point_analysis}

```{r plot_cpa, echo=FALSE, fig.cap=plot_cpa,fig.scap=plot_cpa_short, fig.height=5}

# Analyze a positive and a negative amplification curve from the `RAS002` dataset 
# for change points using the `bcp` and `ecp` packages.
# 
library(bcp)
library(ecp)

# The MBmca package is used to calculate the approximate first derivative 
# of the amplification curve.
library(MBmca)

index <- which(grepl("B.Globin", colnames(RAS002)))

data <- RAS002[, c(1, index)]

amp_data <- data[, c(1,25, 2)]

# Smooth data with moving average for other data
# analysis steps.
dat_smoothed <- cbind(
    chipPCR::smoother(amp_data[, 1], amp_data[, 2]),
                      chipPCR::smoother(amp_data[, 1], amp_data[, 3])
)

# Calculate the first derivative
dat_smoothed_deriv <- cbind(
    suppressMessages(MBmca::diffQ(cbind(amp_data[-c(1,2), 1], dat_smoothed[-c(1,2), 1]), verbose = TRUE)$xy)[, 2],
                            suppressMessages(MBmca::diffQ(cbind(amp_data[-c(1,2), 1], dat_smoothed[-c(1,2), 2]), verbose = TRUE)$xy)[, 2]
)


# Bayesian analysis of change points
# Positive amplification curve
res_bcp_pos <- bcp(dat_smoothed_deriv[, 1])

y2_range <- range(res_bcp_pos$posterior.prob, na.rm = TRUE)

# Negative amplification curve
res_bcp_neg <- bcp(dat_smoothed_deriv[, 2])


# Change point analysis by energy agglomerative clustering

# Positive amplification curve
res_ecp_pos <- ecp::e.agglo(as.matrix(dat_smoothed_deriv[, 1]))$estimates
# Negative amplification curve
res_ecp_neg <- ecp::e.agglo(as.matrix(dat_smoothed_deriv[, 2]))$estimates

par(mfrow = c(2, 3))

plot(amp_data[, 1], amp_data[, 2], xlab = "Cycles", ylab = "RFU", type = "l", lwd = 2)
mtext("A    Negative", cex = 1.2, side = 3, adj = 0, font = 2)

plot(res_bcp_pos$data, xlab = "Cycles", ylab = "d(RFU) / d(cycle)", type = "l", lwd = 2)
mtext("B 1st Derivative", cex = 1.2, side = 3, adj = 0, font = 2)

plot(res_bcp_pos$posterior.prob, xlab = "Cycles", ylab = "Probability", type = "b", lwd = 2, pch = 19, col = "red", ylim = c(0, 1))
res_ecp_pos <- ecp::e.agglo(as.matrix(dat_smoothed_deriv[, 1]))$estimates
abline(v = res_ecp_pos, col = "green", pch = 19)
abline(h = 0.6, col = "grey")
mtext("C   Changepoints", cex = 1.2, side = 3, adj = 0, font = 2)
legend("topright", c("Baysian", "Agglomerative"), pch = c(19, 19), 
       col = c("red", "green"), bty="n")

plot(amp_data[, 1], amp_data[, 3], xlab = "Cycles", ylab = "RFU", type = "l", lwd = 2)
mtext("D    Positive", cex = 1.2, side = 3, adj = 0, font = 2)

plot(res_bcp_neg$data, xlab = "Cycles", ylab = "d(RFU) / d(cycle)", type = "l", lwd = 2)
mtext("E    1st Derivative", cex = 1.2, side = 3, adj = 0, font = 2)

plot(res_bcp_neg$posterior.prob, xlab = "Cycles", ylab = "Probability", type = "b", lwd = 2, pch = 19, col = "red", ylim = c(0, 1))
res_ecp_neg <- ecp::e.agglo(as.matrix(dat_smoothed_deriv[, 2]))$estimates
abline(v = res_ecp_neg, col = "green", pch = 19)
abline(h = 0.6, col = "grey")
mtext("F   Changepoints", cex = 1.2, side = 3, adj = 0, font = 2)
legend("topleft", c("Baysian", "Agglomerative"), pch = c(19, 19), 
       col = c("red", "green"), bty="n")
```

Change point analysis (CPA) encompasses methods to identify or estimate single or multiple locations of distributional changes in a series of data points indexed in time order. A change herein refers to a statistical property. There exist several change point algorithms such as the binary segmentation algorithm [@scott_cluster_1974]. In the change point analysis one assumes independent ordered observations $X_{1}, X_{2}, \ldots, X_{n} \in \mathbb{R}^{\textit{d}}$ [@james_ecp:_2013]. The case of qPCR this is simply the cycle-dependent fluorescence. This is be used to create $k$ homogeneous subsets of unknown size [@erdman_bcp:_2007]. While frequentist methods make an estimation of the parameter at the location (e.&nbsp;g., mean, variance) of the change points at specific points, change point analysis using the Bayesian method produces a probability for the occurrence of a change point at certain points. CPA is used for example in econometrics and bioinformatics [@Killick_2014, @erdman_bcp:_2007]. For the analysis of the amplification curves it was hypothesized that the number of change points differs between positive (sigmoidal) and negative (noise) amplification curves.

The ``pcrfit_single()`` function uses two independent approaches for change point analysis. These are the ``bcp()`` function from the \texttt{bcp} package [@erdman_bcp:_2007] and the ``e.agglo()`` function from the \texttt{ecp} package [@james_ecp:_2013]. The ``e.agglo()`` function performs a non-parametric change point analysis based on agglomerative hierarchical estimation and is useful to "detect changes within the marginal distributions" [@james_ecp:_2013]. Measurement from the qPCR systems typically shows noise that typically has rapidly changing components. Differentiators amplify these rapidly changing noise components [@roediger_RJ_2013]. Therefore, the first derivation of the amplification curve was used for both change point analyses. It was assumed for the change point analysis of amplification curves that this leads to larger differences between positive and negative amplification curves. An example is shown on \autoref{plot_cpa}. In contrast the ``bcp()`` [\texttt{bcp}] function performs a change point analysis based on a Bayesian approach. This method can detect changes in the mean of independent Gaussian observations. As result the analysis returns the posterior probability of a change point at each $X_{i}$. An example is shown on \autoref{plot_cpa}. Both the change point analysis methods provide additional information to distinguish positive and negative amplification curves \autoref{plot_cp_area}E & F).

#### Frequentist Approaches to Test of an Amplification Reaction\label{section_amptester}

A part of the ``pcrfit_single()`` function is the ``amptester()`` function from the \texttt{chipPCR} package. This function contains tests to determine whether an amplification curve is positive or negative. The input values for the function differ due to the different pre-processing steps in the ``pcrfit_single()`` function. Therefore, the concepts of the tests are briefly described below.

- The first test, designated as SHt, is based on this Shapiro-Wilk test of normality. This relatively simple procedure can be used to check whether the underlying population of a sample (amplification curve) is significantly ($\alpha \leq 5e-04$) normal distributed. The name of the output of the ``pcrfit_single()`` function is `amptester_shapiro`.
- The second test is the *Resids growth test* (RGt), which tests if the fluorescence values in linear phase are stable. Whenever no amplification occurs, fluorescence values quickly deviate from linear model. Their standardized residuals will be strongly correlated with their value. For real amplification curves, situation is much more stable. Noise (that means deviations from linear model) in  background do not correlate strongly with the changes in fluorescence. The decision is based on the threshold value (here 0.5). The output is binary coded (negative = 0, positive = 1). The output name of the ``pcrfit_single()`` function is `amptester_rgt`.
- The third test is the *Linear Regression test* (LRt). This test determines the coefficient of determination ($R^{2}$) by an ordinary least squares linear (OLS) regression. The $R^{2}$ are determined from a run of circa 15% range of the data. If a sequence of more than six $R^{2}$s is larger than 0.8 is found that is likely a nonlinear signal. This is a bit counterintuitive because $R^{2}$ of nonlinear data should be low. The output is binary coded (negative = 0, positive = 1). The output name of the ``pcrfit_single()`` function is `amptester_lrt`.
- The fourth test is called *Threshold test* (THt), which is based on the Wilcoxon rank sum test. As a simple rule the first 20% (head) and the last 15% (tail) of an amplification curve are used as input data. From that a one-sided Wilcoxon rank sum tests of the head versus the tail is performed ($\alpha \leq 1e-02$). The output is binary coded (negative = 0, positive = 1). The output name of the ``pcrfit_single()`` function is `amptester_tht`.
- The fifth test is called *Signal level test* (SLt). he test compares the signals of the head and the tail by a robust "sigma" rule (median + 2 * MAD) and the comparison of the head/tail ratio. If the returned value is less than 1.25 (25 percent), then the amplification curve is likely negative.The output is binary coded (negative = 0, positive = 1). The output name of the ``pcrfit_single()`` function is `amptester_slt`.
- The sixth test is called *Polygon test* (pco). The pco test determines if the points in an amplification curve (like a polygon) are in a "clockwise" order. The sum over the edges result in a positive value if the amplification curve is "clockwise" and is negative if the curve is counter-clockwise. From experience is noise positive and "true" amplification curves "highly" negative. In contrast to the implementation in the ``amptester()`` function, the result is normalized by a division to the number of PCR cycles. The output is numeric. The output name of the ``pcrfit_single()`` function is `amptester_polygon`.
- The seventh test is the *Slope Ratio test* (SlR). This test uses the approximated first derivative maximum, the second derivative minimum and the second derivative maximum of the amplification curve. Next the raw fluorescence at the approximated second derivative minimum and the second derivative maximum are taken from the original dataset. The fluorescence intensities are normalized to the maximum fluorescence of this data. This data is used for a linear regression. Where the slope is used. The output is numeric. The output name of the ``pcrfit_single()`` function is `amptester_slope.ratio`.

##### Application of the ``amptester()`` Features

Random Forest is an enhancement of decision tree algorithms. Random Forest uses *n* random data subsets. The subset is to be used to capture trends precisely without taking into account the whole data. To do this, an ensemble consisting of *n* small decision trees is generated. Each decision tree contains a biased classifier. The majority of the previous classes are then selected for classification.
Compared to a single tree classifier, the Random Forest has a high robustness against noise, outliers and over-fitting [@williams_rattle:_2009, @breiman_random_2001].

In the following example, the ``randomForest()`` function from the \texttt{randomForest} package [@liaw_classification_2002] was used for the classification. The aim was to classify positive and negative amplification curves. 
As response vector ($y$) served `decision` with its possible states "positive" and "negative" (factor). The features `amptester_shapiro`, `amptester_lrt`, `amptester_rgt`, `amptester_tht`, `amptester_slt`, `amptester_polygon` and `amptester_slope.ratio` served as a matrix of predictors describing the model to be adapted. The `batsch1`, `HCU32_aggR`, `stepone_std`, `RAS002`, `RAS003`, `lc96_bACTXY` datasets were used for the analysis. This dataset contains almost equal proportions of positive and negative amplification curves (\autoref{plot_Logistic_Regression}A). Prior to this, the amplification curves were analyzed with the ``encu()`` function (\autoref{section_pcrfit_single_pcrfit_parallel}) and stored in the `data_sample.rda` file to save computing time. The file is part of the \texttt{PCRedux} package.


```{r}
library(randomForest)
library(PCRedux)

data <- data_sample[data_sample$dataset %in% 
                c("batsch1",
                  "HCU32_aggR",
                  "lc96_bACTXY",
                  "RAS002", 
                  "RAS003", 
                  "stepone_std"), ]

n_positive <- sum(data[["decision"]] == "y")
n_negative <- sum(data[["decision"]] == "n")

dat <- data.frame(data[, c("amptester_shapiro", 
                           "amptester_lrt", 
                           "amptester_rgt", 
                           "amptester_tht", 
                           "amptester_slt",
                           "amptester_polygon", 
                           "amptester_slope.ratio")],
                  decision = as.numeric(factor(data$decision, 
                                         levels = c("n", "y"), 
                                         label = c(0, 1))) - 1)

# Select randomly observations from 70% of the data for training.
# n_train is the number of observations used for training.

n_train <- round(nrow(data) * 0.7)

# index_test is the index of observations to be selected for the training
index_test <- sample(1L:nrow(dat), size = n_train)

# index_test is the index of observations to be selected for the testing
index_training <- which(!(1L:nrow(dat) %in% index_test))

# train_data contains the data used for training

train_data <- dat[index_test, ]

# test_data contains the data used for training

test_data <- dat[index_training, ]


model_rf = randomForest(decision ~ ., data = train_data, ntree = 4000, 
                        importance = TRUE)


# Determine variable importance
res_importance <- importance(model_rf)
```




```{r plot_random_forest, echo=TRUE, fig.cap=plot_random_forest, fig.scap=plot_random_forest_short, fig.height=3}
par(mfrow = c(1,3))

plot(model_rf, main = "", las = 2)
mtext("A", cex = 1.2, side = 3, adj = 0, font = 2, las = 0)

rownames(res_importance) <- substr(rownames(res_importance), 11, 22)


barplot(t(as.matrix(sort(res_importance[, 1]))), 
        ylab = "%IncMSE", main = "", las = 2,
        col = adjustcolor("grey", alpha.f = 0.5), border = "white")
mtext("B", cex = 1.2, side = 3, adj = 0, font = 2, las = 0)

barplot(t(as.matrix(sort(res_importance[, 2]))), 
        ylab = "IncNodePurity", main = "", las = 2,
        col = adjustcolor("grey", alpha.f = 0.5), border = "white")
mtext("C", cex = 1.2, side = 3, adj = 0, font = 2, las = 0)
```

# Summary and Conclusions  \label{section_Summary_and_conclusions}

Positive amplification curves usually have a sigmoid shape of a ground phase, exponential phase, plateau phase. Negative amplification curves resemble a flat noisy signal.  As a result, the experienced user is normally able to interpret the shape of the curve well. Outliers and measurement errors can also be easily identified by the user in amplification curves. When setting up a qPCR assay, manual handling is a useful and necessary approach to familiarize oneself with the properties of the qPCR amplification curves.

It is challenging to analyze and classify ampilication curves if they deviate significantly from the sigmoid shape or if their number is no longer feasible for manual analysis. The supposed objectivity of the user must also be questioned. In the scientific environment there is often the temptation, or rather the compulsion, to use all data for publications. As a result, amplification curves of rather poor quality are used. In \autoref{htPCR_nap_frequency} it was shown that the reproducible and objective analysis of amplification curves is not always given. For a novice user, the quality of an amplification curve can be acceptable, for an experienced user, however, not. In \autoref{why_PCRedux} further reasons were given why statistical methods are necessary for the objective and reproducible interpretation of the amplification curves.

In the past, developers of qPCR software focused mainly on tools and pipelines for calculating quantitative parameters of amplification curves. Examples of qPCR software are \texttt{quantGenius} [@baebler_quantgenius:_2017] and \texttt{Chainy} [@mallona_chainy:_nodate]. With these software, Cq values and amplification efficiencies can be determined. Both parameters are necessary for expression analysis or genotyping[@pabinger_2014].

There were few approaches where machine learning was used for the analysis of amplification curves. The intention of @gunay_machine_2016 was to improve the determination of Cq values. The authors postulated that they had developed an improved prediction of Cq values using a modified three-parameter model. One assumption of their approach was that their modified three-parameter model could be applied to any amplification curve. However, there are reasons why such an assumption is not valid. In chapters \autoref{section_software} and \autoref{htPCR_nap_frequency} it was described that a considerable proportion of amplification curves deviate clearly from a three-parameter model. Multiparametric models with more than four parameters are more frequently adapted to amplification curves. In addition, the multiparametric models tend to adapt to noise (\autoref{curve_fit_fail}). Unsurprisingly, a Cq-value is calculated for actually negative amplification curves (\autoref{plot_dat_Cq}). This demonstrates that a three-parameter model alone cannot provide reliable predictions. However, a correct model is important for the extraction of Cq values, for the determination of features from the curves and consequently for classification.

Besides the determination of quantification points, the classification of amplification curves is necessary. For example, a diagnostician is interested in whether a sample is positive or negative. In research using high-throughput screening methods, it is important to classify large data sets quickly and cost-effectively. It is important to borne in mind that in manual classification, the classification result is influenced by the subjective perception of the experimenter and that manual analysis is comparatively time-consuming.

An automatic computer-assisted classification of amplification curves is important because it makes the entire analysis process faster, more objective and more reproducible. The objectives were therefore 

1. create a collection of classified amplification curve data,
2. propose algorithms that can be used to calculate features from amplification curves,
3. to develop pipelines (e.g., machine learning, decision trees) that can be used for automatic classification of amplification curves and
4. distribute the results in a public repository as open source software and open data.class

For this purpose the \texttt{PCRedux} package was developed. This package contains proof-of-concept algorithms and functions with which features (mathematically describable properties) of amplification curves can be calculated. The ``pcrfit_single()`` function is an extensible wrapper function for all algorithms and functions developed. Currently `r length(pcrfit_single(chipPCR::C126EG595[, 2]))` features can be calculated from an amplification curve. Other parameters (e.g. hydrolysis probe, DNA binding dye) can be converted into binary classifiers and used for modelling. Machine learning requires features to train a model [@saeys_review_2007]. The model should then be able to put new unknown data into a meaningful context. The features have a significant influence on the accuracy of the prediction model. The features of the  \texttt{PCRedux} package have not yet been described in the literature for the classification of amplification curves. Presumably the collection of features in the \texttt{PCRedux} package were the most extensive at the time of the first release on \url{https://github.com/devSJR/PCRedux} (summer 2017). Nonetheless, it is important to design and test additional features.

To a modest extent, the usefulness of the features was tested using examples. The aim was to achieve a high degree of objectivity and reproducibility. It is probably not possible to fulfill this ideal, since the algorithms were designed from the limited human perspective. That is a generic problem in machine learning. In particular, data records can be distorted if the user excludes seemingly problematic data. It is advisable to perform a more comprehensive analysis of all features. 

Several ROIs (see \autoref{amplification_curve_ROI}) can be obtained from amplification curves. Sigmoidal amplification curves have turning points that can serve as an indicator of a positive amplification curve. These turning points, for example, are mathematical and statistical starting points for calculating features. Amplification curves can have unique forms and sometimes deviate drastically from the ideal sigmoid models (see \autoref{figure_sigmoid_curve}A and \autoref{figure_sigmoid_curve}B). Some amplification curves, for example, have only a slight increase with positive or negative signs without a clear sigmoid curvature.

Almost all real-time thermo-cyclers have built-in software that performs pre-processing steps such as smoothing, baseline correction and normalization on the amplification curves[@roediger2015chippcr, @spiess_impact_2015, @spiess_system-specific_2016]. For this reason it is almost impossible to get access to the raw data. The impact on the feature extraction process cannot be adequately estimated.

At this point it should be mentioned that these approaches were deliberately sought in order to make the approach and implementation more understandable. Self-learning machine-learning methods are expected to be included in the \texttt{PCRedux} package sometime later.

For datasets, it is important that their data volume and classification are representative. The \texttt{PCRedux} package contains a large number of manually classified amplification curves. Data preparation is an important step. It encompasses data cleansing, data transformation and data integration [@herrera_multiple_2016]. The \texttt{xray} package [@Seibelt_xray] can be used to analyze the distribution form and variables in records for anomalies such as missing values, zeros, infinite values and their categories. The ``anomalies()`` function included in the \texttt{xray} package searches for anomalies (including missing values (NA), zero values (Zero), empty strings (Blank) and infinite numbers (Inf)). Users of the\texttt{PCRedux} package should use this kind of tools before proceeding with the datasets. Even though the records in the \texttt{PCRedux} package have a very similar data structure, some records contain missing values or have different dimensions. For example, the data set `C127EGHP` comprises a matrix of `r data_dim <- dim(chipPCR::C127EGHP); paste0(data_dim[1]," x " , data_dim[2])` (35 cycles x observations (65 amplification curves)), while the data set `htPCR` comprises a matrix of `r data_dim <- dim (qpcR::htPCR); paste0 (data_dim[1]," x ", data_dim[2])`.

Although the dataset in the \texttt{PCRedux} package is large compared to what has been available so far, there is no conclusive evidence of how well it represents amplification curves in general. Bellman coined the term "Curse of Dimensionality" in 1961 when he dealt with adaptive control processes. The "Curse of Dimensionality" vaguely describes the practical difficulties of high-dimensional analysis and estimation. There is only a maximum number of features for a sample of a certain size. If there are too many features, the performance of an algorithm decreases rather than improves. As a result, many data mining algorithms fail with high dimensionality because the data points are thin and far apart [@herrera_multiple_2016]. In addition, the following applies:

- The user's knowledge, prejudices and skills are reflected in the classified data. For example, an amplification can be classified as "ambiguous" by one user and "positive" by another user.
- Not all data sets have a comparable number of cases as described above. The majority of the amplification curves of the `C127EGHP` data set have an *ideal* sigmod waveform. In contrast, the amplification curves from the `htPCR` data set are noisy and difficult to classify.
- The user decides
    + how the data is preprocessed,
    + which features are determined by the amplification curves, 
    + which data set is used for machine learning and to what extent,
    + how the models are tested and
    + which results are reported.


For the examples in the \texttt{PCRedux} package, the question was not clarified whether imbalances in the datasets cause a confirmation distortion.

A new concept for the fast groupwise classification of amplification curves was introduced within the \texttt{PCRedux} package. Based on experience with the collected datasets it can be said that only a few iterations are necessary to classify a large dataset. This technology is intended to be a handy assistance for the users of the package.

Actions to minimize the aforementioned sources of error are the implementation of algorithms in the form of an open source package and the publication of the datasets. Unlike black box algorithms and inaccessible data sets, every user can check and correct all steps. Users of \texttt{PCRedux} packages should independently verify whether their models are objective. Failure to do so may have severe implications. For instance, amplification curves classified as false negative may lead to misleading inferences in a forensic analysis. 

Machine-learning algorithms require careful data pre-processing and quality management. In a first step, relatively large data sets of known characteristic vectors have to be collected and features calculated. In a second step, these characteristics are used to classify unknown characteristic vectors using the automatic learning algorithm. As an example, the amplification curves would need to be randomly divided into training data and test data from the whole dataset. Some examples were used in the \texttt{PCRedux} package to show how features of an amplification curve dataset can be calculated and used for classifications.

The scope of the \texttt{PCRedux} package is presumably wide. The quantitation of nucleic acids by curve parameters such as the quantification point (Cq) and the amplification efficiency (AE) is meaningful only if the kinetics has a sigmoid shape [@ruijter_evaluation_2013, @ruijter_2014, @Ritz2008]. In principle, this can be checked out with the \texttt{PCRedux} package. The concepts may also be applied to other bioanalytical methods (e.g. enzyme kinetics) with sigmoidal curves. This will require further studies. Possibly the \texttt{PCRedux} software can also be coupled with other technologies, e.g. Next Generation Sequencing. A typical situation is that the result classes are positive, negative or ambiguous. Ambiguous amplification curves are a big challenge for the user, as both classes (positive and negative) can be true. In most cases, however, the user is interested in an automatic distinction between positive and negative samples. This is important for screening applications. Next generation sequencing depends on pre-tests of the input DNA before sequencing. qPCR is used for pre-testing and as a confirmation test for RNA-Seq quantification. For this purpose, automated quality control and decision support are conceivable.

\newpage

# References
