---
title: "PCRedux package - an overview"
author: "Stefan R&ouml;diger, Micha&#322; Burdukiewicz, Andrej-Nikolai Spiess"
date: "2017-09-15"
output: 
  rmarkdown::html_vignette:
    toc: true
bibliography: "PCRedux.bib"
vignette: >
  %\VignetteIndexEntry{PCRedux package - an overview}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

# Introduction to the analysis of simgmoid shaped curves – or the analysis of 
amplification curve data from quantitative real-time PCR experiments

**PCRedux** is an R package designed to perform analysis of sigmoid curves. A 
sigmoid function results a S-shaped curve. For example, the equation 
\autoref{sigmoid} produces such a curvature.

\begin{equation}
\label{sigmoid}
f(x) = \frac{1}{1 +  e^{-x}}
\end{equation}

These non-linar functions are real-valued and can be differentiated (first derivative 
maximum, with one local minimum and one loacl maximum).

This 
kind of curve is common in many biological assays such as quantitative 
real-time 
PCR (qPCR) experiments. qPCRs are performed in thermo-cyclers. There are many 
manufactures of such systems, which offer such technologies as commercial 
products. Moreover, thermo-cyclers are produced in house as part of scientific 
projects. An example is the VideoScan-technology [@rodiger_highly_2013].

A challenge for the user is, to make sense from the data produced by these 
devices. This is a manageable task if the volume of data is low, or dedicated 
software is available for the analysis thereof. An example for a low number of 
amplification curves is shown in Figure \ref{samples_of_qPCRs}A. All `r 
ncol(chipPCR::C127EGHP)-1` curves exhibit a sigmoid curve shape. In contrast, 
the example in Figure \ref{samples_of_qPCRs}B is no longer manageable by simple 
visual inspection. The data originate from an high-throghput experiment an 
encompas in total `r ncol(qpcR::htPCR)-1` amplification curves. Similarly, a 
manual analysis of the data is time consuming and prone to errors. Dedicated 
software that can distinguish the amplification curves automatically is needed.


```{r, echo=FALSE, fig.cap="Sample data from A) the **C127EGHP** data set (*chipPCR* package, [@rodiger_chippcr:_2015]) and B) from the **htPCR** data set (*qpcR* package, [@ritz_qpcr:_2008]). \\label{samples_of_qPCRs}", fig.height=8, fig.width=8}
    
library(chipPCR)
suppressMessages(library(qpcR))


par(mfrow=c(1,2), las=0, bty="n", oma=c(1,1,1,1))
matplot(C127EGHP[, 2], C127EGHP[, c(-1, -2)], xlab="Cycle", ylab="RFU", 
        main="C127EGHP dataset", type="l", lty=1, lwd=2)
mtext("A", cex = 2, side = 3, adj = 0, font = 2)

matplot(htPCR[, 1], htPCR[, c(-1)], xlab="Cycle", ylab="RFU", 
        main="htPCR dataset", type="l", lty=1, lwd=2)
mtext("B", cex = 2, side = 3, adj = 0, font = 2)
```

There are several open source and closed source software tools, which can be 
used 
for the analysis of qPCR data. The software packages deal for example with
    
    * missing values, 
    * noise,
    * inter run calibration, 
    * normalization, 
    * quantification cycle estimation, 
    * amplification efficiency estimation, 
    * data import, 
    * datat export, 
    * data exchange, 
    * realative gene expression analysis,
    
and other tasks [@pabinger_survey_2014, @rodiger_chippcr:_2015, 
@ritz_qpcr:_2008, @rodiger_r_2015, @lefever_rdml:_2009, 
@ruijter_rdml-ninja_2015, @rodiger_enabling_2017, @perkins_readqpcr_2012, 
@pabinger_qpcr:_2009, @neve_unifiedwmwqpcr:_2014, @feuer_lemming:_2015, 
@mccall_non-detects_2014, @ruijter_evaluation_2013, @ruijter_removal_2015, 
@dvinge_htqpcr:_2009, @ronde_practical_2017, @mallona_pcrefficiency:_2011, 
@mallona_chainy:_nodate]. Many of the are implemented in the R statistical 
computing language. It should be mentioned that this list is far from 
completeness. Dedicated literature is available from peer-reviewed publications 
and textbooks.

Others and we have shown that amplification curve analysis is not a simple 
task. For example selected qPCR systems (device and/or detection chemistry) 
cause a periodicity in the amplification curve data 
[@spiess_system-specific_2016]. Or the commonly employed smoothing of data 
might 
case alterations to the raw data that affects both the estimation of the 
template material (Cq value) and the amplification efficiency 
[@spiess_impact_2015].

At this level all software assumes that the amplification resemble a sigmoid 
curve shape (ideal positive amplification reaction) or a flat low line (ideal 
negative amplification reaction). For example, @ritz_qpcr:_2008 published an R 
package that contains functions to fit several multi-parameter models. This 
includes the 5-parameter Richardson function, which is often used for the 
analysis of qPCR data.

In general, the software packages do not take into account if an amplification 
curve fulfills a certain criterion (e.g., positive or negative, good quality or 
bad quality). This is a needed imformation for some algorithms. For example, 
the 
*linreg* method by @ruijter_amplification_2009 requires a decision, if an 
amplification curve is positive or negative. 

In summary, during our past research we noticed that a many challenges are well 
addressed. In particular, algorithms for the processing of the positive 
amplification curves are widely available. One bottleneck in qPCR data 
analysis is the lack of amplification curve data classifier. Classifier herein 
refer to a vector of features that can be used to distinguish the amplification 
curves by their shape only.

One reason for this is the lack of features that are known for amplification 
curve data. Only few features for amplification curves are described in the 
literature. For example, tools described by us are the ``amptester`` and the 
``amptester.gui`` functions, which are part of the **chipPCR** package 
([@rodiger_chippcr:_2015]). The **qpcR** package ([@ritz_qpcr:_2008]) contains 
an amplification curve test via the ``modlist`` function. The parameter check = 
"uni2" offers an analytical approach, as part of a method for the kinetic 
outlier detection. It checks for a sigmoid structure of the amplification curve. 
Then it tests for the location of the first derivative maximum and the second 
derivative maximum. However, multi-parameter functions, like sigmoid models, fit 
"successful" in most cases including noise and give false positive results. 

# Aim of the `PCRedux` package

The `PCRedux` package contains function and data sets for machine learning on 
quantitative PCR data in R. There are numerous software packages for R which 
can 
be used for the analysis of quantitative PCR (qPCR) data [@rodiger_r_2015, 
@pabinger_survey_2014].

This vignette covers important features of the package and should be used as an 
addendum to the manual.

Reproducible research is an important fundation of research. Both, the technical 
and the experimental aspects need to be performed under principles that follow 
good practics of reproducible research. Serveral authors addressed the matter 
[@huggett_how_2014, @bustin_reproducibility_2014, @bustin_continuing_2017, 
@rodiger_r_2015, @rodiger_enabling_2017, @wilson_good_2016]. During our 
literature studies we found barely material, which can be used for further 
studies. Therefore, we included data set of amplification curves, human rated 
amplification curves (negative, ambiguos, positive) and examples in this 
package.

# Functions of the `PCRedux` package

The function described following are aimed for experimental studies. It is 
important to note that the features proposed herein emerged during a critical 
reasoning process. The aim of the package is to propose a set of features, 
functions and data for an independent research.

## autocorrelation_test

Autocorrelation analysis is a technique that is used in the field of time series 
analysis. It can be used to reveal regularly occurring patterns in 
one-dimensional data [@spiess_system-specific_2016]. The autocorrelation 
function measures the correlation of a signal $f(t)$ with itself shifted by some 
time delay $f(t - \tau)$.
The `autocorrelation_test` function coercres the amplification curve data to an 
object of the class "zoo" (*zoo* package) as indexed totally ordered 
observations. Next follows the computation of a lagged version of the 
amplification curve data. The shifting the amplification curve data is based 
back by a given number of observations (default $\tau = 3$). Then follows a 
significance test for correlation between paired samples (amplification curve 
data & lagged amplification curve data). The hypothesis is that the paired 
sample of positive amplification curves has is significant correlation in 
contrast to negative amplification curves (noise).

The application of the `autocorrelation_test` function is shown in the following 
example. 

```{r, echo=TRUE, fig.cap="Autocorrelation analysis for amplification curves of the **testdat** data set (*qpcR* package, [@ritz_qpcr:_2008]). \\label{autocorrelation}", fig.height=8, fig.width=8}
# Test for autocorrelation in amplification curve data
# Load the libraries magrittr for pipes and qpcR for the data
library(magrittr)
library(PCRedux)
library(qpcR)

# Test for autocorrelation in the testdat dataset
res_ac <- sapply(2:ncol(testdat), function(i) {
    autocorrelation_test(testdat[, i])
}
)

# Plot curve data as overview
# Define the colors for the amplification curves
colors <- rainbow(ncol(testdat)-1, alpha=0.3)
# Names of samples
samples <- colnames(testdat)[-1]
layout(matrix(c(1,2,1,3), 2, 2, byrow = TRUE))
matplot(testdat[, 1], testdat[, -1], xlab="Cycle", ylab="RFU", 
        main="testdat dataset", type="l", lty=1, col=colors, lwd=2)
legend("topleft", samples, pch=19, col=colors, ncol=2, bty="n")

# Curves rated by a human after analysis of the overview. 1 = positive, 
# 0 = negative
human_rating <- c(1,1,0,0,1,1,0,0,
                  1,1,0,0,1,1,0,0,
                  1,1,0,0,1,1,0,0)

# Convert the n.s. (not significant) in 0 and others to 1. 
# Combine the results of the atomatic autocorrelation_test as variable "ac", 
# the human rated values as variable "hr" in a new data frame (res_ac_hr).
res_ac_hr <- data.frame(ac=ifelse(res_ac=="n.s.", 0, 1), 
                        hr=human_rating) %>% as.matrix
                        
# Add ratings by humna and autocorrelation_test to the plot
par(las=2)
barplot(res_ac_hr[, "ac"], main="Rated by autocorrelation_test", col=colors, 
        xlab="Sample", ylab="rating", names.arg=samples)
barplot(res_ac_hr[, "hr"], main="Rated by a human", col=colors, xlab="Sample", 
        ylab="rating", names.arg=samples)
```

## decission_modus

## earlyreg

## head2tailratio

## hookreg

## mblrr

## pcrfit_helper

## pcrfit_parallel

## qPCR2fdata



# Summary and conlusion

The `PCRedux` enables the user to extract features from amplification curve 
data. We implemented numerous features that can be extracted from the 
amplification curve data. Some of them have not been described in the 
literature. We consider `PCRedux` as enabling technology for further research. 
For example, the proposed features can be used for machine learing applications 
or quality assassemt of data.

The ``pcrfit_helper`` function is a wrapper that can be extended if new features 
emerge.

Of note, we would like to emphasise that the functianlity of this package is not 
limited to amplificatiton curve data from qPCR experiments. As stated before, 
amplification curves have a sigmoid curve shape. So do many other processes. 

# References
