---
title: "Supplement to: Algorithms for Automatized Detection of Hook Effect-bearing Amplification Curves"
author: "Stefan R&ouml;diger, Micha&#322; Burdukiewicz, Konstantin A. Blagodatskikh, Werner Lehmann, Peter Schierack, Andrej-Nikolai Spiess"
date: "September 23, 2017"
output: 
  rmarkdown::pdf_document:
    toc: true
bibliography: "literature.bib"
vignette: >
  %\VignetteIndexEntry{Hook Effect}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


# Introduction

This is a supplement document for the study *Algorithms for Automatized 
Detection of Hook Effect-bearing Amplification Curves*. The functions and data 
presented in the paper are available from <https://github.com/devSJR/PCRedux>. 
The data, including the RDML file are part of the **PCRedux** package and are 
made available in the CSV or RDML format [@rodiger_enabling_2017] for a vendor 
independent analysis. We have decided use the HTML format for the presentation 
of the supplementary information to tables hand the reader an documents with 
interactive tables (sorting and filter option).

All analysis were implemented and done with the **R** statistical computing 
language [@R_language, @rodiger_r_2015] and dedicated integrated development 
environments such as **RKWard** [@rodiger_rkward:_2012]. Further documentation 
can be found in the help files of the **R** packages.

We added the source code for the generation of the figures in the main text at 
the end of the document.

> Note: Some rows in the tables below appear to be empty. This may occur in 
cases where the corresponding functions were not able to calculate the 
coefficients due to a failed model fit or violation of truncation criterion.

## Results for the analysis of the hookreg.rdml data set by human rater

All calculations in the following sections were done with `hookreg.rdml` data 
set, which is part of the **PCRedux** package. The data were transfered to the 
**R** environment by the **RDML** package [@rodiger_enabling_2017].

```{r, echo=TRUE, message=FALSE, warning=FALSE}
library(RDML)
library(PCRedux)
library(magrittr)
library(qpcR)
library(DT)
library(htmlwidgets)

raw_data <- RDML$new(filename=system.file("hookreg.rdml", package="PCRedux"))

raw_data_tab <- raw_data$AsTable(name.pattern=paste(
    react$position,
    react$sample$id,
    # run id added to names
    sep="~"))
data <- as.data.frame(raw_data$GetFData(raw_data_tab, long.table=FALSE))

matplot(data[, 1], data[, -1], type="l", lty=1, lwd=2, ylab="RFU", xlab="Cycle")

head(data, 3)
```

Printout of all measured samples, their rating by two humans (rater 1 and rater 
2) with their dichotomous ratings (0, no hook; 1, hook) and their sources.

- The boggy data (qpcR::boggy) set was taken from the **qpcR** package 
[@Ritz2008, @spiess_highly_2008].
- The C127EGHP data (chipPCR::C127EGHP) set was taken from the **chipPCR** package 
[@rodiger_chippcr:_2015].
- The testdat data (qpcR::testdat) set was taken from the **qpcR** package 
[@Ritz2008, @spiess_highly_2008].
- Other data were prepared by Evrogen laboratory experiments.


```{r, echo=TRUE, message=FALSE, warning=FALSE}
library(readxl)
library(DT)
library(htmlwidgets)

Table_human_rated <- read_xlsx(path=system.file("Table_human_rated.xlsx", package="PCRedux"))

datatable(Table_human_rated, options = list(pageLength = nrow(Table_human_rated)))
```

## Results for the analysis of the hookreg.rdml data set with hookreg


```{r, echo=TRUE, message=FALSE, warning=FALSE}
library(PCRedux)
library(magrittr)
library(qpcR)

res_hookreg <- sapply(2L:ncol(data), function(i) {
    hookreg(x=data[, 1], y=data[, i])
}) %>% t %>% data.frame(sample=colnames(data)[-1],.)


res_hookreg_table <- data.frame(sample=as.character(res_hookreg[["sample"]]),
                                intercept=signif(res_hookreg[["intercept"]], 2),
                                slope=signif(res_hookreg[["slope"]], 1),
                                hook.start=signif(res_hookreg[["hook.start"]], 0),
                                hook.delta=signif(res_hookreg[["hook.delta"]], 0),
                                p.value=signif(res_hookreg[["p.value"]], 4),
                                CI.low=signif(res_hookreg[["CI.low"]], 2),
                                CI.up=signif(res_hookreg[["CI.up"]], 2),
                                hook.fit=res_hookreg[["hook.fit"]],
                                hook.CI=res_hookreg[["hook.CI"]],
                                hook=res_hookreg[["hook"]]
)
```

Printout of all measured samples and their sources.


```{r, echo=TRUE, message=FALSE, warning=FALSE}
library(DT)
library(htmlwidgets)
datatable(res_hookreg_table, options = list(pageLength = nrow(res_hookreg_table)))
```

## Results for the analysis of the hookreg.rdml data set with hookregNL


```{r, echo=TRUE, message=FALSE, warning=FALSE}
res_hookregNL <- suppressMessages(sapply(2L:ncol(data), function(i) {
    hookregNL(x=data[, 1], y=data[, i])
}) %>% t %>% data.frame(sample=colnames(data)[-1],.))

res_hookregNL_table <- data.frame(sample=as.character(res_hookregNL[["sample"]]),
                                  slope=signif(as.numeric(res_hookregNL[["slope"]]), 1),
                                  CI.low=signif(as.numeric(res_hookregNL[["CI.low"]]), 2),
                                  CI.up=signif(as.numeric(res_hookregNL[["CI.up"]]), 2),
                                  hook.CI=unlist(res_hookregNL[["hook"]])
)

datatable(res_hookregNL_table, options = list(pageLength = nrow(res_hookregNL_table)))
```


## Comparison of the methods

```{r, echo=FALSE, message=FALSE, warning=FALSE}
library(readxl)
library(dplyr)

Table_human_rated <- read_xlsx(path=system.file("Table_human_rated.xlsx", package="PCRedux"))

res <- data.frame(sample=Table_human_rated[, "Sample"],
                  hr=Table_human_rated[, "Hook effect-like\nRater 2"],
                  hookeg=res_hookreg_table[, "hook"],
                  hookregNL=data.frame(unlist(res_hookregNL[, "hook"]))
          )
colnames(res) <- c("Sample",
                   "Human rater",
                   "hookreg",
                   "hookregNL"
                )
datatable(res, options = list(pageLength = nrow(res)))
```


```{r, echo=FALSE, message=FALSE, warning=FALSE}
meta_hookreg <- sapply(1:nrow(res), function(i){
    ifelse(res[i, "hookreg"] ==1 || res[i, "hookregNL"] == 1, 1, 0)
})

res_out <- data.frame(Sample=res[["Sample"]], res[["Human rater"]], res_hookreg[["hook"]], res_hookregNL_table[["hook.CI"]], meta_hookreg)

colnames(res_out) <- c("Sample",
                   "Human rater",
                   "hookreg",
                   "hookregNL",
                   "hookreg and hoohkreNL combined"
)

datatable(res_out, options = list(pageLength = nrow(res_hookregNL_table)))
```

# Performance analysis

The performance of the **hookreg** and **hookregNL** function was analyzed with 
the **performance** function of the **PCRedux** package.

Abbreviations: TP, true positive; FP, false positive; TN, true negative; FN, false negative

Measure                                    | Formula
-------------------------------------------|-----------------------------------------------
Sensitivity - TPR, true positive rate      | $TPR = \frac{TP}{TP + FN}$
Specificity - SPC, true negative rate      | $SPC = \frac{TN}{TN + FP}$
Precision - PPV, positive predictive value | $PPV = \frac{TP}{TP +  FP}$
Negative predictive value - NPV            | $NPV = \frac{TN}{TN + FN}$
Fall-out, FPR, false positive rate         | $FPR = \frac{FP}{FP + TN} = 1 - SPC$
False negative rate - FNR                  | $FNR = \frac{FN}{TN + FN} = 1 - TPR$
False discovery rate - FDR                 | $FDR = \frac{FP}{TP + FP} = 1 - PPV$
Accuracy - ACC                             | $ACC =  \frac{(TP + TN)}{(TP + FP + FN + TN)}$
F1 score - F1                              | $F1 = \frac{2TP}{(2TP + FP + FN)}$
Matthews correlation coefficient - MCC     | $MCC = \frac{(TP*TN - FP*FN)}{\sqrt{(TP+FP)*(TP+FN)*(TN+FP)*(TN+FN)}}$
Likelihood ratio positive - LRp            | $LRp = \frac{TPR}{1-SPC}$
Cohen''s kappa (binary classification)     | $\kappa=\frac{p_{0}-p_{c}}{1-p_{0}}$


r (reference) is the trusted label and s (sample) is the predicted value

     | r=1 | r=0 
-----|-----|-----
s=1  | a   | b
s=0  | c   | d 

n=a+b+c+d

pc=((a+b)/n)((a+c)/n)+((c+d)/n)((b+d)/n)

po=(a+d)/n

```{r, echo=TRUE, message=FALSE, warning=FALSE}
library(xtable)
res_performance <- rbind(
hookreg=performance(res_out[["hookreg"]], res_out[["Human rater"]]),
hookregNL=performance(res_out[["hookregNL"]], res_out[["Human rater"]]),
combined_hookreg=performance(res_out[["hookreg and hoohkreNL combined"]], res_out[["Human rater"]])
) %>% t %>% signif(., 4)

colnames(res_performance) <- c("hookreg", "hookregNL", "hookreg and hookregNL")
# rownames(res_performance) <- c("Sensitivity (TPR)",
#                                "Specificity (SPC)",
#                                "Precision (PPV)",
#                                "Negative predictive\nvalue (NPV)",
#                                "Fall-out (FPR)",
#                                "False negative rate (FNR)",
#                                "False discovery rate (FDR)",
#                                "Accuracy (ACC)",
#                                "F1 score (F1)",
#                                "Matthews correlation\ncoefficient (MCC)"
# )

xtable(res_performance)

datatable(res_performance, options = list(pageLength = nrow(res_performance)))
```

# References
